### 2023-05-23 writing/abstract.tex
kicked off abstract for AMCTM

\documentclass{article}

\newcommand{\pct}{\texttt{\symbol{37}}}
\newcommand{\dir}{\texttt{\symbol{62}}}
\newcommand{\res}{\texttt{\symbol{60}}}
\newcommand{\lcb}{\texttt{\symbol{123}}}
\newcommand{\rcb}{\texttt{\symbol{125}}}
\newcommand{\lsb}{\texttt{\symbol{91}}}
\newcommand{\rsb}{\texttt{\symbol{93}}}

\newcommand{\lr}{\textsc{LabRat}}
\newcommand{\ma}{\textsc{MatLab}}
\newcommand{\remph}{\emph}
\begin{document}
\title{\lr --- an interactive assistant for \ma}
\author{Fredrik Nordvall Forsberg, Conor McBride, Georgi Nakov}
\maketitle

We set out the prospectus for and report our progress on \lr, an interactive computer system designed to help human users of \ma\ write programs which actively make more of the sense those humans intend. This work is supported by the UK's National Physical Laboratory, and is significantly informed by a corpus of \ma\ code arising from their work, along with conversations and observations of their praxis. The latter is often healthy from an engineering perspective, with a discipline of documenting the structure and significance of each variable in \remph{comments} which, although not formal, exhibit a certain regularity of format. That is to say, the humans communicate to \remph{one another} information vital for comprehension of and interaction with their software, whilst withholding the very same from the \remph{machine}. We can and should do better.

\lr\ introduces a language of \emph{directives} within the existing \ma\ comment syntax:
\[\pct\dir\;\mathit{directive}
\]
\ma\ ignores these directives entirely, but \lr\ does not. \lr's input is \ma\ source code with directives on which to act; its output is a new version of that source code, transformed and queried as directed. We envisage four varieties of directive:
\begin{itemize}
\item documenting the \remph{type} of a function or variable, to determine not only its \ma\ data representation, but also more refined metadata, e.g., units of measure;
\item querying the type of a function, a variable, or a hole in the program;
\item requesting the systematic transformation of the user's code, e.g., renaming all occurrences of a given variable;
\item generating low-level code serving some stated high-level purpose on the basis of the metadata supplied, e.g., to read a data set from a file.
\end{itemize}
Our intent is that programmers run \lr\ iteratively on their work in progress, gaining support, feedback, and the alleviation of such system\remph{atic} labour that we can render a \remph{system}. \lr\ acts on source code files, rather than by direct manual controls, so may readily be integrated with existing editors with minimal assumptions as to which. \lr\ communicates back to its users by including \emph{responses} (which are \ma\ comments meriting user attention)
\[
\pct\res\;\mathit{response}
\]
and \emph{syntheses} (which are \ma\ code segments meriting user indifference)
\[\begin{array}{l}
\pct\res\lcb\qquad \mbox{(which is a \remph{single line} \ma\ comment)}\\
\mathit{synthesized}\; \mathit{code}\\
\pct\res\rcb\qquad \mbox{(also a \remph{single line} \ma\ comment)} \\
\end{array}\]

The type of a \ma\ matrix variable is given by, at worst, its sizes in each dimension and its entry type. However, a size is but a number and such numbers are rightly seen as a special case of a \remph{list of headers}, such as you might find in the top row or left column of a spreadsheet: the type of each cell may depend on the headers which govern its position. An important use of this refinement takes headers to be \remph{units of measure}: our intent is for \lr\ to check that the units of each cell respect these headers and thus ensure that each inner product computed in matrix multiplication is dimensionally meaningful and in commensurate units.

Current status.

Ulterior motive.

\end{document}

### 2023-05-23 writing/AMCTM2023/abstract.tex
[ AMCTM ] first draft

\documentclass{ws-procs9x6}

%%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorlinks=true]{hyperref}
\usepackage{microtype}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%% Macros %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\pct}{\texttt{\symbol{37}}}
\newcommand{\dir}{\texttt{\symbol{62}}}
\newcommand{\res}{\texttt{\symbol{60}}}
\newcommand{\lcb}{\texttt{\symbol{123}}}
\newcommand{\rcb}{\texttt{\symbol{125}}}
\newcommand{\lsb}{\texttt{\symbol{91}}}
\newcommand{\rsb}{\texttt{\symbol{93}}}

\newcommand{\lr}{\textsc{LabRat}}
\newcommand{\ma}{\textsc{MatLab}}
\newcommand{\remph}{\emph}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\title{\lr --- an interactive assistant for \ma}
\author{Conor McBride, Georgi Nakov, and Fredrik Nordvall Forsberg}
\address{University of Strathclyde}

\bodymatter

\paragraph{Introduction}

We set out the prospectus for and report our progress on \lr, an interactive computer system designed to help human users of \ma{} write programs which actively make more of the sense those humans intend. This work is supported by the UK's National Physical Laboratory, and is significantly informed by a corpus of \ma{} code arising from their work, along with conversations and observations of their praxis. The latter is often healthy from an engineering perspective, with a discipline of documenting the structure and significance of each variable in \remph{comments} which, although not formal, exhibit a certain regularity of format. That is to say, the humans communicate to \remph{one another} information vital for comprehension of and interaction with their software, whilst withholding the very same from the \remph{machine}. We can and should do better.

%\paragraph{Ulterior motive}

Our goal is not to get practitioners to give up their existing toolchains in favour of some research level prototype language --- this would neither be productive nor feasible. Instead, \lr{} works with existing (work in progress) \ma{} programs, and offers encouragement both in carrot and stick form: by making more of the structure and meaning of programs formal, \lr{} can both generate boring parts of programs automatically (carrot), as well as complain when things do not seem to make sense (stick). Our operation is inspired by Andrew Kennedy's work on adding type to PHP at Facebook~\cite{hack}, which was carried out as a similar stealth operation.

\paragraph{Directives and Responses}

\lr{} introduces a language of \emph{directives} within the existing \ma{} comment syntax:
\[\pct\dir\;\mathit{directive}
\]
\ma{} ignores these directives entirely, but \lr{} does not. \lr's input is \ma{} source code with directives on which to act; its output is a new version of that source code, transformed and queried as directed. We envisage four varieties of directive:
\begin{itemize}
\item documenting the \remph{type} of a function or variable, to determine not only its \ma{} data representation, but also more refined metadata, e.g., units of measure;
\item querying the type of a function, a variable, or a hole in the program;
\item requesting the systematic transformation of the user's code, e.g., renaming all occurrences of a given variable;
\item generating low-level code serving some stated high-level purpose on the basis of supplied metadata, e.g., to read a data set from a file.
\end{itemize}
Our intent is that programmers run \lr{} iteratively on their work in progress, gaining support, feedback, and the alleviation of such system\remph{atic} labour that we can render a \remph{system}. \lr{} acts on source code files, rather than by direct manual controls, so may readily be integrated with existing editors with minimal assumptions as to which. \lr{} communicates back to its users by including \emph{responses} (which are \ma{} comments meriting user attention)
\[
\pct\res\;\mathit{response}
\]
and \emph{syntheses} (which are \ma{} code segments meriting user indifference)
\[\begin{array}{l}
\pct\res\lcb\qquad \mbox{(which is a \remph{single line} \ma{} comment)}\\
\mathit{synthesized}\; \mathit{code}\\
\pct\res\rcb\qquad \mbox{(also a \remph{single line} \ma{} comment)} \\
\end{array}\]

\paragraph{\lr{} types}

The type of a \ma{} matrix is given by, at worst, its sizes in each dimension and its entry type. However, a size is but a number and such numbers are rightly seen as a special case of a \remph{list of headers}, such as you might find in the top row or left column of a spreadsheet: the type of each cell may depend on the headers which govern its position. An important use of this refinement takes headers to be \remph{units of measure}: our intent is for \lr{} to check that the units of each cell respect these headers and thus ensure that inner products computed in matrix multiplication are dimensionally meaningful and in commensurate units~\cite{dimTypes}.

\paragraph{Current status}

\lr{} is under active development, and can be downloaded from \url{https://github.com/msp-strath/LabRat}. We can currently parse most \ma{} code we have encountered in the wild, and have implemented directives for renaming variables and generating code for reading data sets from a file based on the \textsc{MGen} input description format~\cite{mgen}.

\bibliographystyle{ws-procs9x6}
\bibliography{labrat}

\end{document}

### 2023-05-26 writing/AMCTM2023/abstract.tex
[ AMCTM ] missing bib file
tweaked the hack bit
[ AMCTM ] prose boiling

@@
 
-\newcommand{\lr}{\textsc{LabRat}}
-\newcommand{\ma}{\textsc{MatLab}}
+\newcommand{\lr}{\textsc{LabMate}}
+\newcommand{\repo}{\url{https://github.com/msp-strath/LabMate}}
+\newcommand{\ma}{\textsc{MATLAB}}
 \newcommand{\remph}{\emph}
@@
 \begin{document}
+
 \title{\lr --- an interactive assistant for \ma}
@@
 We set out the prospectus for and report our progress on \lr, an interactive computer system designed to help human users of \ma{} write programs which actively make more of the sense those humans intend.
-This work is supported by the UK's National Physical Laboratory, and is significantly informed by a corpus of \ma{} code arising from their work, along with conversations and observations of their praxis.
-The latter is often healthy from an engineering perspective, with a discipline of documenting the structure and significance of each variable in \remph{comments} which, although not formal, exhibit a certain regularity of format.
-That is to say, the humans communicate to \remph{one another} information vital for comprehension of and interaction with their software, whilst withholding the very same from the \remph{machine}.
+This work is supported by the UK's National Physical Laboratory, and is significantly informed by a code corpus arising from their work, along with observations of their praxis.
+The latter is solid from an engineering perspective, with a discipline of documenting the structure and significance of each variable in \remph{comments} which, although not formal, exhibit a certain regularity of format.
+That is to say, the humans communicate to \remph{one another} information vital for comprehension of their software, whilst withholding the very same from the \remph{machine}.
 We can and should do better.
@@
 
-Our goal is not to get practitioners to give up their existing toolchains in favour of some research level prototype language --- this would neither be productive nor feasible.
-Instead, \lr{} works with existing (work in progress) \ma{} programs, and offers encouragement both in carrot and stick form: by making more of the structure and meaning of programs formal, \lr{} can both generate boring parts of programs automatically (carrot), as well as complain when things do not seem to make sense (stick).
-Our operation is inspired by Andrew Kennedy's work on adding type to PHP at Facebook~\cite{hack}, which was carried out as a similar stealth operation.
+Our goal is to enhance practitioners' experience of their existing toolchains, rather than to insist on the adoption of a radically different language prototype.
+\lr{} gives a formal language of comments for \ma{} programs, and can thus help both preventatively (with safety checks ahead of execution) and constructively (generating correct code from richer information).
+Our approach is inspired by Andrew Kennedy's work at Facebook on adding types incrementally to PHP~\cite{hack}.
 
@@
 
-\lr{} introduces a language of \emph{directives} within the existing \ma{} comment syntax:
+\lr{} introduces a language of \emph{directives}:
 \[\pct\dir\;\mathit{directive}
 \]
-\ma{} ignores these directives entirely, but \lr{} does not.
-\lr's input is \ma{} source code with directives on which to act; its output is a new version of that source code, transformed and queried as directed.
-We envisage four varieties of directive:
+which \ma{} treats as comments, but \lr{} acts on, transforming code with directives to a new version of that code in four ways:
 \begin{itemize}
-\item documenting the \remph{type} of a function or variable, to determine not only its \ma{} data representation, but also more refined metadata, e.g., units of measure;
-\item querying the type of a function, a variable, or a hole in the program;
-\item requesting the systematic transformation of the user's code, e.g., renaming all occurrences of a given variable;
+\item documenting \remph{types} of functions and variables, giving not only data representation, but also richer metadata, e.g., units of measure;
+\item querying the type of a part of the program, or a gap in it;
+\item requesting the systematic transformation of the user's code, e.g., consistently renaming all occurrences of a given variable;
 \item generating low-level code serving some stated high-level purpose on the basis of supplied metadata, e.g., to read a data set from a file.
 \end{itemize}
-Our intent is that programmers run \lr{} iteratively on their work in progress, gaining support, feedback, and the alleviation of such system\remph{atic} labour that we can render a \remph{system}.
-\lr{} acts on source code files, rather than by direct manual controls, so may readily be integrated with existing editors with minimal assumptions as to which.
-\lr{} communicates back to its users by including \emph{responses} (which are \ma{} comments meriting user attention)
+Programmers can run \lr{} iteratively on their work in progress, gaining support, feedback, and the alleviation systematic drudgery.
+\lr{} is controlled only by the directives in its input, so may readily be integrated with existing editors with minimal assumptions as to which.
+\lr{} communicates back by including \emph{responses} (comments meriting attention)
 \[
@@
 \]
-and \emph{syntheses} (which are \ma{} code segments meriting user indifference)
+and \emph{syntheses} (\ma{} code segments meriting user indifference)
 \[\begin{array}{l}
@@
 
-The type of a \ma{} matrix is given by, at worst, its sizes in each dimension and its entry type.
-However, a size is but a number and such numbers are rightly seen as a special case of a \remph{list of headers}, such as you might find in the top row or left column of a spreadsheet: the type of each cell may depend on the headers which govern its position.
-An important use of this refinement takes headers to be \remph{units of measure}: our intent is for \lr{} to check that the units of each cell respect these headers and thus ensure that inner products computed in matrix multiplication are dimensionally meaningful and in commensurate units~\cite{dimTypes}.
+The type of a \ma{} matrix is given by, at worst, its sizes in each dimension and cell type.
+However, a size is but a number and thus a special case of a list of \remph{headers}, such as you might find in a spreadsheet: the type of each cell may depend on the headers which govern its position.
+A key use of this refinement takes headers to be \remph{units}: \lr{} will check that the units of each cell respect these headers so that inner products computed during matrix multiplication are dimensionally meaningful~\cite{dimTypes}.
 
@@
 
-\lr{} is under active development, and can be downloaded from \url{https://github.com/msp-strath/LabRat}.
-We can currently parse most \ma{} code we have encountered in the wild, and have implemented directives for renaming variables and generating code for reading data sets from a file based on the \textsc{MGen} input description format~\cite{mgen}.
+\lr{} is in development and available at \repo.
+We can currently parse most \ma{} code we have encountered in the wild, and have implemented directives for renaming variables and generating code for reading data sets from files based on \textsc{MGen} input descriptions~\cite{mgen}.
 

### 2023-05-26 writing/AMCTM2023/abstract.tex
[ AMCTM ] more prose repair

@@
 \end{itemize}
-Programmers can run \lr{} iteratively on their work in progress, gaining support, feedback, and the alleviation systematic drudgery.
-\lr{} is controlled only by the directives in its input, so may readily be integrated with existing editors with minimal assumptions as to which.
-\lr{} communicates back by including \emph{responses} (comments meriting attention)
+Programmers can run \lr{} iteratively on their work in progress, gaining support, feedback, and the alleviation of systematic drudgery.
+Controlled only by the directives in its input, \lr\ may readily be integrated with existing editors with minimal assumptions as to which.
+It communicates back by including \emph{responses} (comments with useful information)
 \[
@@
 \]
-and \emph{syntheses} (\ma{} code segments meriting user indifference)
+and \emph{syntheses} (code segments unlikely to be of interest)
 \[\begin{array}{l}
 \pct\res\lcb\qquad \mbox{(which is a \remph{single line} \ma{} comment)}\\
-\mathit{synthesized}\; \mathit{code}\\
+\mathit{synthesized}\; \ma\;\mathit{code}\\
 \pct\res\rcb\qquad \mbox{(also a \remph{single line} \ma{} comment)} \\

### 2023-05-26 writing/AMCTM2023/abstract.tex
[ AMCTM ] MATLAB -> Matlab
[ AMCTM ] unlikely to merit inspection

@@
 \newcommand{\repo}{\url{https://github.com/msp-strath/LabMate}}
-\newcommand{\ma}{\textsc{MATLAB}}
+\newcommand{\ma}{\textsc{Matlab}}
 \newcommand{\remph}{\emph}
@@
 Programmers can run \lr{} iteratively on their work in progress, gaining support, feedback, and the alleviation of systematic drudgery.
-Controlled only by the directives in its input, \lr\ may readily be integrated with existing editors with minimal assumptions as to which.
-It communicates back by including \emph{responses} (comments with useful information)
+Controlled only by the directives in its input, \lr\ may readily be integrated with existing editors.
+It communicates back only via its output, including \emph{responses} (comments with useful information)
 \[
@@
 \]
-and \emph{syntheses} (code segments unlikely to be of interest)
+and \emph{syntheses} (code segments unlikely to merit inspection)
 \[\begin{array}{l}

### 2023-05-26 src/Machine.hs
rename to LabMate

@@ -70,7 +70,7 @@ data DeclarationType
       [(String, ResponseLocation)] -- requested name and how to reply
                                    -- (we hope of length at most 1)
       Bool     -- is it capturable?
-  | LabratDecl
+  | LabmateDecl
   deriving Show
 
 data Problem
@@ -114,12 +114,12 @@ initMachine :: File -> Map Nonce String -> MachineState
 initMachine f t = MS
   { position = B0 :<+>: []
   , problem = File f
-  , nameSupply = (B0 :< ("labrat", 0), 0)
+  , nameSupply = (B0 :< ("labmate", 0), 0)
   , nonceTable = t
   }
 
 findDeclaration :: DeclarationType -> Bwd Frame -> Maybe (Name, Bwd Frame)
-findDeclaration LabratDecl fz = Nothing
+findDeclaration LabmateDecl fz = Nothing
 findDeclaration (UserDecl old seen news _) fz = go fz False where
   go B0 _ = Nothing
   go (fz :< Locale ScriptLocale) True = Nothing
@@ -129,7 +129,7 @@ findDeclaration (UserDecl old seen news _) fz = go fz False where
   go (fz :< f) b = fmap (:< f) <$> go fz b
 
 makeDeclaration :: DeclarationType -> MachineState -> (Name, MachineState)
-makeDeclaration LabratDecl ms = error "making labratdecl declaration"
+makeDeclaration LabmateDecl ms = error "making labmatedecl declaration"
 makeDeclaration d@(UserDecl s seen news capturable) ms = case fresh s ms of
   (n, ms) -> case position ms of
     fz :<+>: fs -> (n, ms { position = findLocale fz (Declaration n d) :<+>: fs })

### 2023-05-26 src/Machine/Reassemble.hs

@@ -87,7 +87,7 @@ renamePass ms = inbound ms
             cap <- captured fz b s
             when cap $ tellGripes Captured d
           pure s
-        _ -> error "Impossible: renaming LabRat var"
+        _ -> error "Impossible: renaming LabMate var"
       else do
         s <- renamer fz n b
         s' <- newName d

### 2023-05-26 src/Main.hs

@@ -28,7 +28,7 @@ import Machine
 import Machine.Reassemble
 
 import Data.Version
-import Paths_LabRat
+import Paths_LabMate
 
 type ParseError = (Maybe FilePath, Reach, Int)
 
@@ -48,7 +48,7 @@ main = do
    stdin = T.getContents >>= actInput
    go (Right (tab, cs@(_ :<=: (n,src)))) = do
       let out = run (initMachine cs tab)
-      putStrLn ("%< LabRat " ++ showVersion version)
+      putStrLn ("%< LabMate " ++ showVersion version)
       putStrLn $ reassemble n out
    go (Left e) = do printError e; exitWith fatalExitCode
 

### 2023-09-25 examples/assign.m
[ wip ] typesynth dir responses

@@ -1,3 +1,4 @@
 %> n :: int
 n = 5;
 x = n
+%> typeof x

### 2023-09-25 src/CoreTT.hs

@@ -31,7 +31,7 @@ instance Show Meta where
   show = \case
     Meta{..} -> nattily (vlen mctxt) $
       concat [ "Meta{", show mctxt
-             , " , ", show mtype , " , ", show mdefn
+             , " , ", show mtype, " , ", show mdefn
              , " , ", show mstat
              , "}"]
 

### 2023-09-25 src/Machine.hs

@@ -30,7 +30,7 @@ import qualified Data.Set as Set
 
 import Debug.Trace
 
-debug = trace
+debug = const id --trace
 
 type Elab = State MachineState
 
@@ -61,6 +61,13 @@ initMachine f t = MS
 type TERM = Term Chk ^ Z
 type TYPE = Type ^ Z
 
+data DiagnosticData
+  = SynthD
+      TYPE {- T -}
+      TERM {- t :: T -}
+      Expr {- e, t = elab e -}
+  deriving Show
+
 data Frame where
   Source :: Source -> Frame
   Declaration :: Name -> DeclarationType TYPE -> Frame
@@ -71,6 +78,7 @@ data Frame where
   Fork :: (Either Fork Fork) -> [Frame] -> Problem -> Frame
   Problems :: [Problem] -> Frame
   LocalNamespace :: NameSupply -> Frame
+  Diagnostic :: DiagnosticData -> ResponseLocation -> Frame
   deriving Show
 
 data Fork = Solved | FunctionName Name
@@ -290,6 +298,12 @@ normalise :: Context n -> Type ^ n -> Term Chk ^ n -> Elab (Norm Chk ^ n)
 normalise ctx ty tm  = elabTC ctx (checkEval ty tm) >>= \case
   Left err -> error $ "normalise : TC error" ++ err
   Right tm -> pure tm
+
+normaliseDiagnostic :: DiagnosticData -> Elab DiagnosticData
+normaliseDiagnostic (SynthD ty tm e) = do
+  ty <- normalise emptyContext (atom SType) ty
+  tm <- normalise emptyContext ty tm
+  pure $ SynthD ty tm e
 
 constrain :: String -> Constraint -> Elab ConstraintStatus
 constrain s c@Constraint{..} = debug ("Constrain call " ++ s ++ " constraint = " ++ show c) $ case (traverse (traverse isHom) constraintCtx, constraintType) of
@@ -625,12 +639,13 @@ runDirective rl (dir :<=: src, body) = do
     Typecheck ty e -> do
       (tySol, tyProb) <- elab "typeToCheck" emptyContext (atom SType) (TensorTask <$> ty)
       (eSol, eProb) <- elab "exprToCheck" emptyContext tySol (ExprTask <$> e)
-      push $ Problems [eProb]
+      --traverse push [Diagnostic _ rl, Problems [eProb]]
       newProb tyProb
       run
     SynthType e -> do
[…]

### 2023-09-25 src/Machine/Reassemble.hs

@@ -44,7 +44,6 @@ reassemble n ms =
       tab' = foldr (\((n, c), grp) rs -> Map.insertWith (++) n (concat ["\n", replicate c ' ', "%< ", renameGripeResponse grp]) rs) tab probs
   in fromMaybe [] $ Map.lookup n (updateNonceTable tab' (resetCursor (position $ if Set.null probs then ms' else ms)))
 
-
 updateNonceTable :: Map Nonce String -> Cursor Frame -> Map Nonce String
 updateNonceTable table (fz :<+>: []) = table
 updateNonceTable table (fz :<+>: Fork _ fs' e : fs) = updateNonceTable (updateNonceTable table (fz :<+>: fs)) (fz :<+>: fs')
@@ -146,7 +145,6 @@ renamePass ms = inbound ms
        inner [] = Nothing
     respond (t : ts) = t : respond ts
     respond [] = []
-
 
     tellGripes :: RenameGripe -> DeclarationType a -> Writer RenameProblems ()
     tellGripes grp = \case

### 2023-09-25 src/Parse/Matlab.hs

@@ -119,6 +119,7 @@ pdirhead = Declare <$> psep1 (pspc <|> punc ",") (pws pnom) <* pospc <* psym "::
     <|> InputFormat <$ pkin Nom "input" <* pospc <*> pnom
     <|> Typecheck <$ pkin Nom "typecheck" <* pospc <*> ptensortype <* pspc <*> pexpr topCI
     <|> SynthType <$ pkin Nom "typeof" <* pospc <*> pexpr topCI
+--    <|> EverythingOkay <$ pkin Nom ""
 
 ptensortype :: Parser TensorType
 ptensortype = pws ptensortype'

### 2023-09-25 src/Syntax.hs

@@ -34,6 +34,7 @@ data DirHeader
   | InputFormat String {- name of function -}
   | Typecheck ConcreteType Expr
   | SynthType Expr
+--  | EverythingOkay
   deriving Show
 
 data DirBody

### 2023-09-26 writing/AMCTM2023/presentation.tex
[ wip ] amctm presentation

\documentclass[handout]{beamer}
\usepackage[utf8x]{inputenc}
\usepackage{pgfpages}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xspace}
\usepackage{xparse}
%\setbeameroption{show notes on second screen}

\usepackage{microtype}

\usepackage{tikz}
\usetikzlibrary{shapes.callouts, shapes.geometric, arrows, arrows.meta, chains}

% speech bubbles
\tikzset{
  invisible/.style={opacity=0,text opacity=0},
  visible on/.style={alt=#1{}{invisible}},
  alt/.code args={<#1>#2#3}{%
    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}} % \pgfkeysalso doesn't change the path
  },
}
\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture,baseline=-0.5ex] \node (#1) {};}

\NewDocumentCommand{\mycallout}{r<> O{opacity=0.8,text opacity=1} m m}{%
  \tikz[remember picture, overlay]\node[align=center, fill=blue!20, text width=15em,
  #2,visible on=<#1>, rounded corners,
  draw,rectangle callout,anchor=pointer,callout relative pointer={(230:3em)}]
  at (#3) {#4};
}

% flowcharts
\tikzset{
  reg/.style={
    rectangle,
    rounded corners,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=black
  },
  arrow/.style={thick,->},
  fade/.style={
    rectangle,
    rounded corners,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=gray!30,
    text=gray!30,
  },
  fadearrow/.style={thick,->, gray!30},
  highlight/.style={
    rectangle,
    rounded corners,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=black,
    fill=blue!10
  },
  end/.style={
    tape,
    tape bend top=none,tape bend height=2mm,
    tape bend bottom=out and in,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=black
  },
  fadeend/.style={
  tape,
  tape bend top=none,tape bend height=2mm,
  tape bend bottom=out and in,
  minimum width=3cm,
  minimum height=1cm,
  align=center,
  draw=gray!30,
  text=gray!30
  }
}

\newcommand{\pct}{\texttt{\symbol{37}}}
\newcommand{\dir}{\texttt{\symbol{62}}}
\newcommand{\res}{\texttt{\symbol{60}}}
\newcommand{\lcb}{\texttt{\symbol{123}}}
\newcommand{\rcb}{\texttt{\symbol{125}}}
\newcommand{\lsb}{\texttt{\symbol{91}}}
\newcommand{\rsb}{\texttt{\symbol{93}}}
\newcommand{\keyword}[1]{\textcolor{blue}{#1}}
\newcommand{\lr}{\textsc{LabMate}}
\newcommand{\repo}{\url{https://github.com/msp-strath/LabMate}}
\newcommand{\ma}{\textsc{Matlab}}

\setbeamertemplate{frametitle}[default][center]
\title{\huge \lr\ :\\ an interactive assitant for \ma }
\author[McBride, Nakov, Nordvall Forsberg]{\small Conor McBride, \underline{Georgi Nakov}, Fredrik Nordvall Forsberg}
\institute[]{University of Strathclyde}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{Motivation}
\begin{itemize}
 \item %already a good engineering practice
\end{itemize}

\end{frame}

\begin{frame}{Living with(in) \ma}
\begin{itemize}
  \item enhance the experience of working with the existing toolchains
  \begin{itemize}
    \item rather than insisting on adopting a totally new language
  \end{itemize}
  \item develop a language of formal comments
  \begin{itemize}
    \item \lr\ interacts with \ma\ via comments
    \item can be integrated into editors as an optional processing step
    %\item inspired by gradual typing,
  \end{itemize}
  \item requests to \lr\,, i.e., \emph{directives}
  \[\pct\dir\;\mathit{directive}\]
  \item responses from \lr
\end{itemize}
\end{frame}

\begin{frame}{Classifying directives}
  What can \lr\ do for us?
\begin{itemize}
  \item perform systematic transformation of the user's code, e.g., consistently renaming all occurrences of a given variable;
  \item generate low-level code serving some stated high-level purpose via supplied metadata, e.g., to read a data set from a file.
  \item offer means to document \emph{types} of functions and variables
  \begin{itemize}
    \item provide richer metadata, e.g., units of measure;
  \end{itemize}
  \item query the type of a part of the program
\end{itemize}
\end{frame}

\begin{frame}{Code transformations}
 \begin{itemize}
   \item
 \end{itemize}
\end{frame}

\begin{frame}{Insights into scope management}

\end{frame}

\begin{frame}{Code generation}

\end{frame}

\begin{frame}{Types as machine-checkable documentation}
What is a type system?
\begin{itemize}
  \item set of \keyword{logical rules} that assign a type to various expressions in a programming language.
  \item can prevent some classes of errors by \keyword{typechecking during compilation}
  \item varies in expressivity and complexity of implementation
\end{itemize}
Expressive types systems:
\begin{itemize}
  \item understand semantic facts about our programs
  \item enable the developers to state their intent
  \item can offer guidance during writing the code
  \item gives machine-certified correctness guarantees at compile-time
\end{itemize}
\end{frame}

%{
%  \setbeamertemplate{background}
%  {
%    \begin{tikzpicture}[remember picture, overlay]
%    \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
%      \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
%    };
%    \end{tikzpicture}
%  }
\begin{frame}{Anatomy of a MatLab assistant}
  Software dealing with \keyword{user code as input} (e.g., compilers) typically follow a common architecture:\vspace{1em}
  \begin{columns}
  \begin{column}{0.3\textwidth}
  \begin{tikzpicture}[
    start chain=going below,
    every join/.style={arrow},
    node distance=0.6cm
    ]
    \node (s0) [reg,on chain,join] {Parsing};
    \node (s1) [reg, on chain, join] {Elaboration};
    \node (s2) [reg, on chain, join] {Typechecking};
    \node (s3) [end, on chain, join] {Other stuff};
  \end{tikzpicture}
  \begin{tikzpicture}[remember picture, overlay]
    \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
      \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
    };
  \end{tikzpicture}
  \end{column}
  \begin{column}{0.55\textwidth}
    \begin{itemize}
      \item converts the user code to some very \keyword{faithful internal representation}
      \vspace{0.5em}\item translates to a \keyword{simpler}, potentially more verbose representation
      \vspace{0.2em}\item \keyword{validates} and \keyword{checks} the simpler representation
      \vspace{1em}\item optimisations, environments for code execution, emitting native code, etc.
    \end{itemize}
  \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Parsing Matlab code}
  \begin{columns}
  \begin{column}{0.3\textwidth}
    \begin{tikzpicture}[
      start chain=going below,
      every join/.style={fadearrow},
      node distance=0.6cm
    ]
    \node (s0) [highlight, on chain,join] {Parsing};
    \node (s1) [fade, on chain, join] {Elaboration};
    \node (s2) [fade, on chain, join] {Typechecking};
    \node (s3) [fadeend, on chain, join] {Other stuff};
    \end{tikzpicture}
    \begin{tikzpicture}[remember picture, overlay]
      \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
        \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
      };
    \end{tikzpicture}
  \end{column}
  \begin{column}{0.55\textwidth}
  Parsing actually consists of two phases --- \keyword{lexing} and \keyword{parsing}:
  \begin{itemize}
    \item a \keyword{lexer} converts stream of \keyword{characters} to a stream of \keyword{tokens}, e.g., by grouping digits into numbers, finding comments, etc;
    \item a parser then converts the stream of \keyword{tokens} to an \keyword{abstract syntax tree} - a data structure amenable to further processing
  \end{itemize}
  There is much more to that phase...\\
  Remark: Detecting \lr\ directive happens during lexing, i.e., very early in the pipeline.
  \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Core Type Theory}
  \begin{columns}
  \begin{column}{0.3\textwidth}
    \begin{tikzpicture}[
      start chain=going below,
      every join/.style={fadearrow},
      node distance=0.6cm
      ]
      \node (s0) [fade, on chain,join] {Parsing};
      \node (s1) [fade, on chain, join] {Elaboration};
      \node (s2) [highlight, on chain, join] {Typechecking};
      \node (s3) [fadeend, on chain, join] {Other stuff};
    \end{tikzpicture}
    \begin{tikzpicture}[remember picture, overlay]
      \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
        \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
      };
    \end{tikzpicture}
  \end{column}
  \begin{column}{0.55\textwidth}
    The algorithmic implementation of the rules of our type system.
    \begin{itemize}
      \item can report any typing errors during compilation
      \item but can also be delayed until program execution, as in \ma
      \begin{itemize}
        \item to the peril of the user
      \end{itemize}
    \end{itemize}
  \end{column}
\end{columns}
\end{frame}

\begin{frame}{A Tour of Types: Unit type}
\end{frame}

\begin{frame}{A Tour of Types: Free Abelian Groups}

\end{frame}

\begin{frame}{A Tour of Types: Matrices}

\end{frame}

\begin{frame}{Elaborating Matlab expressions to CoreTT}

\end{frame}

\begin{frame}[fragile]{Elaborating Assignments}
  \begin{lstlisting}[escapechar=@, language=Matlab, xleftmargin=10em, gobble=4]
    %> n@\tikzmark{ndecl}@ :: int
    n = 5;
    x = n
    %> typeof x
  \end{lstlisting}
  \mycallout<0>{ndecl}{see a variable declaration $n$; invent the type }
\end{frame}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2023-09-26 src/Machine.hs

@@ -30,7 +30,7 @@ import qualified Data.Set as Set
 
 import Debug.Trace
 
-debug = const id --trace
+debug = trace
 
 type Elab = State MachineState
 
@@ -822,7 +822,7 @@ move = pull >>= \case
   Nothing -> do
     cleanup
     st@MS{..} <- get
-    if (clock > epoch)
+    if clock > epoch
       then do
         put st{ epoch = clock }
         run
@@ -875,10 +875,10 @@ cleanup = do
     modify $ \st@MS{..} -> st{ metaStore = Map.insert name meta metaStore }
   modify $ \st@MS{..} -> st{ metaStore = Map.filter (\Meta{..} -> mstat /= Hoping || isNothing mdefn) metaStore }
 
-
 diagnosticPass :: Elab ()
 diagnosticPass = do
   st <- get
+  pure ()
 
 generateInputReader :: String -- ^
   -> [String] -- ^

### 2023-09-27 writing/AMCTM2023/presentation.tex
[ AMCTM23 ] fix typo
[ AMCTM23 ] presentation

@@
-\documentclass[handout]{beamer}
+\documentclass[]{beamer}
 \usepackage[utf8x]{inputenc}
@@
 \usepackage{listings}
+\usepackage{lstautogobble}
 \usepackage{xspace}
 \usepackage{xparse}
+\usepackage{matlab-prettifier}
 %\setbeameroption{show notes on second screen}
@@
 
+\lstset{ %
+  language=Matlab,
+  style=Matlab-editor,
+  autogobble=true,
+  escapechar=!
+}
+
 \newcommand{\pct}{\texttt{\symbol{37}}}
@@
 \newcommand{\ma}{\textsc{Matlab}}
+\newcommand{\List}{\mathsf{List}}
+\newcommand{\Unit}{\mathsf{Unit}}
+\newcommand{\Abel}{\mathsf{Abel}}
+\newcommand{\Int}{\mathsf{Int}}
+\newcommand{\Matrix}{\mathsf{Matrix}}
+\newcommand{\Enum}{\mathsf{Enum}}
+\newcommand{\LL}{\mathsf{L}}
+\newcommand{\TT}{\mathsf{T}}
+\newcommand{\MM}{\mathsf{M}}
+\newcommand{\Kg}{\mathsf{kg}}
+\newcommand{\Sec}{\mathsf{s}}
+\newcommand{\Km}{\mathsf{m}}
+\newcommand{\NameTimes}{\!\!\times\!\!}
+
+%\usepackage[backend=biber, style=apa, citestyle=apa, isbn=false, doi=false]{biblatex}
+%\bibliography{labrat}
 
 \setbeamertemplate{frametitle}[default][center]
-\title{\huge \lr\ :\\ an interactive assitant for \ma }
+\title{\huge \lr: \\ \centerline{an interactive assistant for \ma} }
 \author[McBride, Nakov, Nordvall Forsberg]{\small Conor McBride, \underline{Georgi Nakov}, Fredrik Nordvall Forsberg}
@@
 
-\begin{frame}{Motivation}
-\begin{itemize}
- \item %already a good engineering practice
+\begin{frame}[fragile]{Motivation}
+\begin{itemize}[<+->]
+\item A lot of metrology and testing software involves complex processing of data in various formats, converting between multiple units of measure
+\item And lot of that software is implemented in \ma
+\item A good and common practice is to document the \keyword{structure} and \keyword{significance} of each variable
+\begin{lstlisting}[basicstyle=\small]
+  function x_d = trajectory(x, ...)
+  % x is the vector of angular positions
+  % of the actuators, a 4x1 double
+  R = 6371e3; % Earth radius in metres
+  laa = vehicle(1)*pi/180; % Lat of vehicle in radians
+  ...
+\end{lstlisting}
 \end{itemize}
-
 \end{frame}
 
-\begin{frame}{Living with(in) \ma}
+\begin{frame}{Motivation}
+  These comments are a treasure trove:
+  \begin{itemize}[<+->]
+    \item convey \keyword{information vital for comprehension} of that software
+    \item written by the user for other human users, and expressing \keyword{intent}
     \begin{itemize}
-  \item enhance the experience of working with the existing toolchains
+      \item and completely invisible to the machine
+    \end{itemize}
+    \item although in free form, the prose is \keyword{semi-structured} with emerging patterns
+  \end{itemize}
+  \visible<5->{\lr{} is our attempt to unearth the hidden riches.}
+\end{frame}
+
+\begin{frame}[fragile]{Living with(in) \ma}
+\lr{} --- a processor and a language of \keyword{formal \ma{} comments}
+\pause
+\begin{itemize}[<+->]
+  \item enhances the experience of working with the existing toolchains
   \begin{itemize}
@@
   \end{itemize}
-  \item develop a language of formal comments
-  \begin{itemize}
-    \item \lr\ interacts with \ma\ via comments
+  \item requests (input) to \lr\,, i.e., \keyword{directives}:\\
+  \begin{center}
+  \begin{tabular}{p{8em}p{4em}l}
+    &  & $\pct\lcb$ \\
+    $\pct\dir\;\mathit{directive}$& or & $\quad\dir\;\mathit{multiline\;directive}$  \\
+    &  & $\pct\rcb$ \\
+  \end{tabular}
+  \end{center}
+  \item responses (output) from \lr:
+  \begin{center}
+  \begin{tabular}{p{8em}p{4em}p{9.3em}}
+    &  & $\pct\res\lcb$ \\
+    $\pct\res\;\mathit{response}$& or & $\mathit{synthesized}\; \ma\;\mathit{code}$  \\
+    &  & $\pct\res\rcb$ \\
+  \end{tabular}
+  \end{center}
     \item can be integrated into editors as an optional processing step
-    %\item inspired by gradual typing,
-  \end{itemize}
-  \item requests to \lr\,, i.e., \emph{directives}
-  \[\pct\dir\;\mathit{directive}\]
-  \item responses from \lr
+  \item run iteratively to maintain a dialogue with the user
 \end{itemize}
@@
 \begin{frame}{Classifying directives}
-  What can \lr\ do for us?
-\begin{itemize}
-  \item perform systematic transformation of the user's code, e.g., consistently renaming all occurrences of a given variable;
-  \item generate low-level code serving some stated high-level purpose via supplied metadata, e.g., to read a data set from a file.
-  \item offer means to document \emph{types} of functions and variables
+  What can \lr\ do for us?\pause
+\begin{itemize}[<+->]
+  \item perform \keyword{systematic transformation} of the user's code, e.g., consistently renaming all occurrences of a given variable;
+  \item \keyword{generate} low-level \keyword{code} serving some stated high-level purpose via supplied metadata, e.g., to read a data set from a file.
+  \item offer means to document \keyword{types} of functions and variables
   \begin{itemize}
@@
   \end{itemize}
-  \item query the type of a part of the program
+  \item \keyword{query} the type of a part of the program
 \end{itemize}
@@
 
-\begin{frame}{Code transformations}
+\begin{frame}[fragile]{Code transformations}
+Renaming a variable is perhaps the simplest and most common refactoring action\pause
+ \begin{itemize}[<+->]
+   \item \lstinline|%> rename x y|
+   \item happens \keyword{simultaneously} and \keyword{transactionally} with all the other rename directives
    \begin{itemize}
-   \item
+     \item if all renamings \keyword{succeed}, notify the user: \lstinline|%< renamed x y|
+     \item if any renaming \keyword{fails}, do not change the file...
+   \end{itemize}
+   \item but how can renaming fail?
  \end{itemize}
@@
 \begin{frame}{Insights into scope management}
-
+When exactly should renaming fail? \pause
+\begin{itemize}
+  \item renaming should be \keyword{non-capturing}, i.e., keep distinct variables distinct
+\end{itemize}
+\pause
+\begin{columns}
+  \begin{column}{0.4\textwidth}
+    \includegraphics[width=0.3\paperwidth, keepaspectratio]{renaming.png}
+  \end{column}
+  \begin{column}{0.5\textwidth}
+  \keyword{Scope of a variable}:
+  \begin{itemize}[<+->]
+    \item a variable defined at the \keyword{top level of the script} is distinct from a variable defined in a function (even if they share the same name)
+    \item a variable defined \keyword{in a function} $f$ is visible to any nested functions of $f$
+  \end{itemize}
+  \visible<5->{\keyword{Observation}: two subsequent renamings $x \mapsto y \mapsto x$ must not change the meaning of the program}
+  \vspace{3em}
+  \end{column}
+\end{columns}
 \end{frame}
@@
 \begin{frame}{Code generation}
-
+There are lot of menial programming tasks in the wild\pause
+\begin{itemize}[<+->]
+  \item e.g., code that handles data serialisation
+  \begin{itemize}
+    \item have to keep in mind what input is required, what are the dimensions, what format, etc...
+    \item but this is meta information, not necessarily evident in the \ma{} implementation
+  \end{itemize}
+\end{itemize}
+\visible<5->{We can do better}
+\pause
+\begin{itemize}[<+->]
+  \item develop a high-level \keyword{metalanguage of descriptions}
+  \item read and validate descriptions, make sanity checks and generate \ma{} code that closely resembles hand crafted one
+\end{itemize}
 \end{frame}
 
-\begin{frame}{Types as machine-checkable documentation}
-What is a type system?
+\begin{frame}[fragile]{Code generation}
+ \begin{columns}
+   \begin{column}{0.45\textwidth}
+    \begin{lstlisting}[basicstyle=\small]
+%{
+  > input readFromDisk
+  myStruct :
+    count : @Int
+    totalLen : @Km
+    unitLens : @count elements of @m
+%}
+    \end{lstlisting}
+   \end{column}
+   \begin{column}{0.45\textwidth}
+   \begin{lstlisting}[basicstyle=\tiny]
+%<{
+  function myStruct = readFromDisk(fname);
+  f1 = fopen(fname);
+  c1 = textscan(f1, `%f`);
+  src = c1{1};
+  fclose(f1);
+  readPtr = 1;
+  myStruct.count = src[readPtr];
+  readPtr = readPtr + 1;
+  myStruct.totalLen = src[readPtr];
+  readPtr = readPtr + 1;
+  for i = 1:myStruct.count
+    myStruct.unitLens[i] = src[readPtr + i];
+  end
+%>}
+   \end{lstlisting}
+   \end{column}
+ \end{columns}
+\end{frame}
+
+\begin{frame}[fragile]{Types as machine-checkable documentation}
+Types are a well established method of \keyword{communication} between the user and the machine.\pause\\
+\vspace{1em}
+\lr{} should enable the user to: \pause
+\begin{itemize}[<+->]
+  \item \keyword{annotate} variables with type information\\
+  \begin{center}
+    \lstinline|%> n :: int|
+  \end{center}
   \begin{itemize}
+    \item support \keyword{inferring} the types of most variables, thus eliminating the need for thorough annotations
+  \end{itemize}
+  \item make sure a specific type \keyword{admits} some \ma{} expression
+  \begin{center}
+    \lstinline|%> typecheck int n|
+  \end{center}
+  \item \keyword{ask} for the type of an \ma{} expression
+  \begin{center}
+    \lstinline|%> typeof n|
+  \end{center}
+\end{itemize}
+\visible<7->{Crucially, this relies on developing a \keyword{type system} for \ma{}.}
+\end{frame}
+
+\begin{frame}{Interlude: Type Systems}
+What is a \keyword{type system}? \pause
+\begin{itemize}[<+->]
   \item set of \keyword{logical rules} that assign a type to various expressions in a programming language.
@@
 \end{itemize}
-Expressive types systems:
-\begin{itemize}
-  \item understand semantic facts about our programs
-  \item enable the developers to state their intent
-  \item can offer guidance during writing the code
-  \item gives machine-certified correctness guarantees at compile-time
+\visible<5->{\keyword{Expressive} type systems:}\pause
+\begin{itemize}[<+->]
+  \item understand \keyword{semantic facts} about the programs
+  \item enable the developers to state their \keyword{intent}
+  \item provide \keyword{guidance} during writing the code
+  \item offer machine-certified \keyword{correctness guarantees} at compile-time
 \end{itemize}
@@
 
-%{
-%  \setbeamertemplate{background}
-%  {
-%    \begin{tikzpicture}[remember picture, overlay]
-%    \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
-%      \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
-%    };
-%    \end{tikzpicture}
-%  }
-\begin{frame}{Anatomy of a MatLab assistant}
-  Software dealing with \keyword{user code as input} (e.g., compilers) typically follow a common architecture:\vspace{1em}
+\begin{frame}{Implementing a type system}
+  A software implementation of an \keyword{expressive type system} within a compiler typically follows a common architecture:\vspace{1em}\pause
   \begin{columns}
@@
     ]
-    \node (s0) [reg,on chain,join] {Parsing};
+    \node (s0) [reg, on chain, join] {Parsing};
     \node (s1) [reg, on chain, join] {Elaboration};
@@
   \begin{column}{0.55\textwidth}
-    \begin{itemize}
-      \item converts the user code to some very \keyword{faithful internal representation}
-      \vspace{0.5em}\item translates to a \keyword{simpler}, potentially more verbose representation
-      \vspace{0.2em}\item \keyword{validates} and \keyword{checks} the simpler representation
+    \pause
+    \begin{itemize}[<+->]
+      \item convert the user code to some very \keyword{faithful internal representation}
+      \vspace{0.5em}\item translate to a \keyword{simpler}, potentially more verbose representation
+      \vspace{0.2em}\item \keyword{validate} and \keyword{check} the simpler representation
       \vspace{1em}\item optimisations, environments for code execution, emitting native code, etc.
@@
 
-\begin{frame}{Parsing Matlab code}
-  \begin{columns}
-  \begin{column}{0.3\textwidth}
-    \begin{tikzpicture}[
-      start chain=going below,
-      every join/.style={fadearrow},
-      node distance=0.6cm
-    ]
-    \node (s0) [highlight, on chain,join] {Parsing};
-    \node (s1) [fade, on chain, join] {Elaboration};
-    \node (s2) [fade, on chain, join] {Typechecking};
-    \node (s3) [fadeend, on chain, join] {Other stuff};
-    \end{tikzpicture}
-    \begin{tikzpicture}[remember picture, overlay]
-      \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
-        \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
-      };
-    \end{tikzpicture}
-  \end{column}
-  \begin{column}{0.55\textwidth}
-  Parsing actually consists of two phases --- \keyword{lexing} and \keyword{parsing}:
-  \begin{itemize}
-    \item a \keyword{lexer} converts stream of \keyword{characters} to a stream of \keyword{tokens}, e.g., by grouping digits into numbers, finding comments, etc;
-    \item a parser then converts the stream of \keyword{tokens} to an \keyword{abstract syntax tree} - a data structure amenable to further processing
-  \end{itemize}
-  There is much more to that phase...\\
-  Remark: Detecting \lr\ directive happens during lexing, i.e., very early in the pipeline.
-  \end{column}
-  \end{columns}
-\end{frame}
+%\begin{frame}{Parsing Matlab code}
+%  \begin{columns}
+%  \begin{column}{0.3\textwidth}
+%    \begin{tikzpicture}[
+%      start chain=going below,
+%      every join/.style={fadearrow},
+%      node distance=0.6cm
+%    ]
+%    \node (s0) [highlight, on chain,join] {Parsing};
+%    \node (s1) [fade, on chain, join] {Elaboration};
+%    \node (s2) [fade, on chain, join] {Typechecking};
+%    \node (s3) [fadeend, on chain, join] {Other stuff};
+%    \end{tikzpicture}
+%    \begin{tikzpicture}[remember picture, overlay]
+%      \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
+%        \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
+%      };
+%    \end{tikzpicture}
+%  \end{column}
+%  \begin{column}{0.55\textwidth}
+%  Parsing actually consists of two phases --- \keyword{lexing} and \keyword{parsing}:
+%  \begin{itemize}
+%    \item a \keyword{lexer} converts stream of \keyword{characters} to a stream of \keyword{tokens}, e.g., by grouping digits into numbers, finding comments, etc;
+%    \item a parser then converts the stream of \keyword{tokens} to an \keyword{abstract syntax tree} - a data structure amenable to further processing
+%  \end{itemize}
+%  There is much more to that phase...\\
+%  Remark: Detecting \lr\ directive happens during lexing, i.e., very early in the pipeline.
+%  \end{column}
+%  \end{columns}
+%\end{frame}
 
@@
   \begin{column}{0.55\textwidth}
-    The algorithmic implementation of the rules of our type system.
+    The implementation of the \keyword{algorithmic} rules of the type system. \pause
+    \begin{itemize}[<+->]
+      \item reports any typing \keyword{errors} during compilation
       \begin{itemize}
-      \item can report any typing errors during compilation
-      \item but can also be delayed until program execution, as in \ma
+        \item \lr{}'s approach
+      \end{itemize}
+      \item but in principle can also be delayed until program execution
       \begin{itemize}
         \item to the peril of the user
+        \item \ma{}'s approach
       \end{itemize}
@@
 \begin{frame}{A Tour of Types: Unit type}
+Let's take a look at some of the types implemented in \lr{}.\pause \\
+\begin{itemize}[<+->]
+  \item Starting with the most important type.
+  \item The type ($\Unit$) with a single inhabitant ($\star$)
+  \[\star :: \Unit\]
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{A Tour of Types: Free Abelian Groups}
-
+\begin{itemize}[<+->]
+  \item $\Abel(G)$ - the type of free Abelian groups over an arbitrary type of generators $G$
+  \begin{itemize}
+    \item can then define the type of integers, $\Int$, as the type $\Abel(\Unit)$.
+  \end{itemize}
+  \item $\Enum(xs)$ - the enumeration type for some list of constants $xs$
+  \begin{itemize}
+    \item letting $xs$ be the constants for length ($\LL$), time ($\TT$), and mass ($\MM$), we can model the type of \keyword{fundamental dimensions}, $\mathsf{D} := \Enum([\LL, \TT, \MM])$
+  \end{itemize}
+  \item An \keyword{arbitrary dimension} $d$ can be represented as an element of the free Abelian group over the type of fundamental dimensions:
+  \[d :: \Abel(\mathsf{D})\]
+\end{itemize}
 \end{frame}
 
-\begin{frame}{A Tour of Types: Matrices}
+\begin{frame}[fragile]{A Tour of Types: Matrices}
+\begin{itemize}[<+->]
+  \item $\Matrix(T, n, m)$ - a matrix over the type of elements $T$ with dimensions $n :: \Abel(\Unit)$ and $m :: \Abel(\Unit)$
+  \item good attempt, but models only \keyword{uniform matrices}
+  \item many applications in metrology collect together distance, speed, acceleration, etc, in one matrix — of related, but different dimensions, e.g.,
+  \[\begin{pmatrix}
+    1\, \Km/\Kg & 2\, \Km/\Sec & 3\, \Km \cdot \Km \\
+    4\, \Sec/\Kg & 5\, \Sec/\Sec & 6\, \Sec \cdot \Km
+  \end{pmatrix}\]
+\end{itemize}
+\end{frame}
 
+\begin{frame}[fragile]{A Tour of Types: Matrices}
+\begin{itemize}[<+->]
+  \item $\Matrix(R, C, T, rs, cs)$
+  \begin{itemize}
+  \item $rs$ and $cs$ - row and column headers
+  \item $R$, $C$ - types of the elements in the row and column headers
+  \item $T$ - type of the cell entries in the matrix
+  \end{itemize}
+  \item very similar representation to a spreadsheet
+  \begin{center}
+    \includegraphics[width=0.7\linewidth]{spreadsheet}
+  \end{center}
+
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{Elaborating Matlab expressions to CoreTT}
-
+  \begin{columns}
+  \begin{column}{0.3\textwidth}
+    \begin{tikzpicture}[
+      start chain=going below,
+      every join/.style={fadearrow},
+      node distance=0.6cm
+      ]
+      \node (s0) [fade, on chain,join] {Parsing};
+      \node (s1) [highlight, on chain, join] {Elaboration};
+      \node (s2) [fade, on chain, join] {Typechecking};
+      \node (s3) [fadeend, on chain, join] {Other stuff};
+    \end{tikzpicture}
+    \begin{tikzpicture}[remember picture, overlay]
+      \node [xshift=0.12\paperheight,yshift=0.15\paperheight] at (current page.south west){
+        \includegraphics[width=0.2\paperwidth, keepaspectratio]{mate.png}
+      };
+    \end{tikzpicture}
+  \end{column}
+  \begin{column}{0.55\textwidth}
+    The \keyword{interface} between the type system and the user written code. \pause
+    \begin{itemize}[<+->]
+      \item tries to fill the gaps between the two
+      \begin{itemize}
+        \item e.g., comes up with types for \ma{} expressions
+      \end{itemize}
+      \item makes sure the type errors are human readable
+    \end{itemize}
+  \end{column}
+\end{columns}
 \end{frame}
@@
 \begin{frame}[fragile]{Elaborating Assignments}
-  \begin{lstlisting}[escapechar=@, language=Matlab, xleftmargin=10em, gobble=4]
-    %> n@\tikzmark{ndecl}@ :: int
-    n = 5;
-    x = n
-    %> typeof x
+  \begin{lstlisting}[xleftmargin=10em]
+    %> n!\tikzmark{ndecl}! :: int!\tikzmark{intdecl}!
+    n !\tikzmark{nlhs}!=!\tikzmark{ass1}! 5!\tikzmark{five}!;
+    x !\tikzmark{xlhs}!=!\tikzmark{ass2}! n !\tikzmark{nrhs}!
+    %> typeof x!\tikzmark{xtypeof}!
   \end{lstlisting}
-  \mycallout<0>{ndecl}{see a variable declaration $n$; invent the type }
+  \pause
+  \mycallout<+>{ndecl}{see a new variable $n$}
+  \mycallout<+>{intdecl}{simplify $n$'s type}
+  \mycallout<+>{ndecl}{declare $n$ in scope with type $int$}
+  \mycallout<+>{ass1}{see an assignment at yet unknown type}
+  \mycallout<+>{five}{process the RHS, $5$, using the the type of LHS $n$}
+  \mycallout<+>{nlhs}{find the type of the declared $n$}
+  \mycallout<+>{ass1}{constrain the type of RHS,$5$, to be equal to that of LHS, $n$}
+  \mycallout<+>{ass2}{see an assignment at yet unknown type}
+  \mycallout<+>{nrhs}{simplify the RHS, $n$, finding its type}
+  \mycallout<+>{xlhs}{declare $x$ in scope with unknown type}
+  \mycallout<+>{ass2}{constrain the type of RHS, $n$, to be equal to that of LHS, $x$}
+  \mycallout<+>{xtypeof}{find the type of the declared $x$}
 \end{frame}
+
+\begin{frame}[fragile]{Current Progress}
+  \begin{itemize}[<+->]
+    \item The code is available at \repo
+    \item We are currently parsing a lot of common \ma{} constructions
+    \item Directives for renaming and code generation are ready
+    \item Implemented most of the underlying type system for \ma
+    \item Actively working on elaboration
+  \end{itemize}
+  \pause
+  \vspace{3em}
+  \LARGE \centerline{Thank you for your attention!}
+\end{frame}
+
+%\begin{frame}[allowframebreaks]
+%  \frametitle{References}
+%  \printbibliography
+%\end{frame}
 

### 2023-09-27 writing/AMCTM2023/demo/minimal.m

@@ -1,0 +1,7 @@
+%{
+  > input readFromDisk
+myStruct :
+  count : @Int (`count` has integral type)
+  totalLen : @Km (`len` carries unit information)
+  unitLens : @count elements of @m (`unitLens` is an array containing `count` many entries, each stored in m)
+%}

### 2023-09-27 writing/AMCTM2023/demo/nested.m

@@ -1,0 +1,18 @@
+
+f1()
+
+function f1()
+  % This is an example of a renaming the Matlab editor gets wrong.
+  % It should be disallowed, because it captures w in f2.
+
+  %> rename x z
+
+  x = 5;
+  f2();
+  disp(x);
+
+  function y = f2()
+    z = 7;
+    y = z;
+  end
+end

### 2023-09-27 writing/AMCTM2023/demo/simple.m

@@ -1,0 +1,5 @@
+y = 5;
+
+%> rename y x
+
+z = y + y;

### 2023-09-27 writing/AMCTM2023/demo/toomany.m

@@ -1,0 +1,3 @@
+x = 5
+%> rename x z
+%> rename x w

### 2023-09-27 writing/AMCTM2023/presentation.tex
[ AMCTM23 ] minor fixes

@@
     \item have to keep in mind what input is required, what are the dimensions, what format, etc...
-    \item but this is meta information, not necessarily evident in the \ma{} implementation
+    \item but this is metainformation, not necessarily evident in the \ma{} implementation
   \end{itemize}
@@
       \end{itemize}
-      \item makes sure the type errors are human readable
+      \item makes sure that type errors are human readable
     \end{itemize}
@@
   \mycallout<+>{nlhs}{find the type of the declared $n$}
-  \mycallout<+>{ass1}{constrain the type of RHS,$5$, to be equal to that of LHS, $n$}
+  \mycallout<+>{ass1}{constrain the type of RHS, $5$, to be equal to that of LHS, $n$}
   \mycallout<+>{ass2}{see an assignment at yet unknown type}
@@
     \item We are currently parsing a lot of common \ma{} constructions
-    \item Directives for renaming and code generation are ready
+    \item Directives for renaming and code generation are working
     \item Implemented most of the underlying type system for \ma

### 2023-09-27 writing/AMCTM2023/demo/nested.m

@@ -1,9 +1,8 @@
-
 f1()
 
 function f1()
   % This is an example of a renaming the Matlab editor gets wrong.
-  % It should be disallowed, because it captures w in f2.
+  % It should be disallowed, because it captures z in f2.
 
   %> rename x z
 

### 2023-10-05 src/Lex.hs
[ wip ] responding to type synth ... in progress

@@ -71,6 +71,9 @@ dump = Hide (0,0)
 sym :: String -> Tok
 sym s = Tok s Sym dump
 
+spc :: Int -> Tok
+spc n = Tok (replicate n ' ') Spc dump
+
 type Doc = (T.Text, Pos)
 
 lexer :: T.Text -> [Tok]

### 2023-10-05 src/Machine.hs

@@ -827,7 +827,7 @@ move = pull >>= \case
         put st{ epoch = clock }
         run
       else do
-        diagnosticPass
+        diagnosticRun
   Just fr -> case fr of
     Fork (Right frk) fs' p' -> do
       (fs, p) <- switchFwrd (fs', p')
@@ -875,10 +875,43 @@ cleanup = do
     modify $ \st@MS{..} -> st{ metaStore = Map.insert name meta metaStore }
   modify $ \st@MS{..} -> st{ metaStore = Map.filter (\Meta{..} -> mstat /= Hoping || isNothing mdefn) metaStore }
 
-diagnosticPass :: Elab ()
-diagnosticPass = do
-  st <- get
-  pure ()
+diagnosticRun :: Elab ()
+diagnosticRun = llup >>= \case
+  Just fr -> case fr of
+    Diagnostic dd (n, dent) -> do
+      nt <- gets nonceTable
+      msg <- msg dent dd
+      push $ Source (n, Hide msg)
+      --modify $ \st@MS{..} -> st { nonceTable = Map.insert n msg nt }
+      push fr
+      diagnosticRun
+    _ -> push fr >> diagnosticRun
+  Nothing -> diagnosticMove
+  where
+    msg :: Int -> DiagnosticData -> Elab [Tok]
+    msg dent = \case
+      SynthD _ _ (e :<=: (en, _)) -> pure
+        [Tok "\n" Ret dump, spc dent, sym "%<", spc 1, Tok "Hello" Nom dump, spc 1, non en]
+      _ -> pure [Tok "\n" Ret dump, spc dent, sym "%<", spc 1, Tok "Goodbye" Nom dump]
+        {- case dd of
+        SynthD ty tm (e :<=: eSrc) -> case () of
+         _ | Set.null (dependencies ty) -> case () of
+             _ | Set.null (dependencies tm) -> _
+        --concat  ["\n", replicate dent ' ', "%<", " ", "Hello"] -}
+
+diagnosticMove :: Elab ()
+diagnosticMove = pull >>= \case
+  Nothing -> pure ()
+  Just fr -> case fr of
+    Fork (Right frk) fs' p' -> do
+      (fs, p) <- switchFwrd (fs', p')
+      shup $ Fork (Left frk) fs p
+      diagnosticMove
+    Fork (Left frk) fs' p' -> do
+      (fs, p) <- switchFwrd (fs', p')
+      push $ Fork (Right frk) fs p
[…]

### 2024-03-19 src/CoreTT.hs
[ new ] numerical matrix multiplication typechecking v1

@@ -465,6 +465,14 @@ evalSynth tm = withScope $ case tm of
                                       , t' //^ sub0 (D $^ tgt <&> atom Sfst))
             | otherwise -> error "evalSynth: funny pair"
       _ -> fail "evalSynth: eliminator for an unknown type"
+  tm | Just tm <- MX $? tm, (lmx, rmx) <- split tm -> do
+    (lmx, lmxTy) <- evalSynth lmx
+    (rmx, rmxTy) <- evalSynth rmx
+    -- TODO: FIXME
+    retTy <- case (tagEh lmxTy, tagEh rmxTy) of
+      (Just (SMatrix, [rowTy, midTy, lcellTy, rs, _]), Just (SMatrix, [_, colTy, rcellTy, _, cs])) ->
+        pure (mk SMatrix rowTy colTy lcellTy {- YIKES -} rs cs)
+    pure (E $^ MX $^ (R $^ lmx <&> lmxTy) <&> (R $^ rmx <&> rmxTy), retTy)
   M (x, n) :$ sig :^ th -> metaLookup x >>= \case
     Meta {..}
       | Just Refl <- nattyEqEh (vlen mctxt) n ->
@@ -581,8 +589,8 @@ checkEvalMatrixNF nf ty@(rowTy, colTy, cellTy) mx@(rs, cs) tm
            cs1 <- prefixEh colTy cs0 cs
            h   <- nf rowTy rs0
            pure ([(h, [NFNeutral tm])], ((rs1, cs0), (rs0, cs1)))
-         _ -> fail "checkEvalMatrixAbel:"
-  | otherwise = fail "checkEvalMatrixAbel: not a valid matrix ctor"
+         _ -> fail "checkEvalMatrixAbel"
+  | otherwise = withScope $ fail $ "checkEvalMatrixAbel: not a valid matrix ctor " ++ show tm
 
 
 propEh :: Type ^ n -> TC n Bool

### 2024-03-19 src/Machine.hs

@@ -134,7 +134,7 @@ instance PrettyPrint MachineState where
   pprint st = do
     let (fz :<+>: fs) = position st
     lines <- pprint (fz <>> fs, problem st)
-    pure (show (metaStore st) : lines)
+    pure (show (metaStore st) : show (constraintStore st) : lines)
 
 type TERM = Term Chk ^ Z
 type TYPE = Type ^ Z
@@ -227,6 +227,7 @@ data ElabTask where
   LHSTask :: LHS' -> ElabTask
   ExprTask :: Expr' -> ElabTask
   Await :: NATTY n => ConstraintStatus -> Term Chk ^ n -> ElabTask
+  ConstrainTask :: NATTY n => (String, Constraint) -> Term Chk ^ n -> ElabTask
   Abandon :: ElabTask -> ElabTask
 deriving instance Show ElabTask
 
@@ -520,13 +521,29 @@ constrain s c@Constraint{..} =  case (traverse (traverse isHom) constraintCtx, c
     ty  <- normalise ctx (atom SType) ty
     lhs <- normalise ctx ty lhs
     rhs <- normalise ctx ty rhs
+    lhsFlexEh <- case lhs of
+       E :$ (M (x, n) :$ sig) :^ th
+         | Just meta@Meta{..} <- x `Map.lookup` ms
+         , Just Refl <- nattyEqEh n (vlen mctxt)
+         , Just Refl <- nattyEqEh n (vlen ctx) -> do
+           lhs' <- normalise mctxt mtype (nattily n (wrapMeta x))
+           pure (lhs == lhs')
+       _ -> pure False
+    rhsFlexEh <- case rhs of
+       E :$ (M (x, n) :$ sig) :^ th
+         | Just meta@Meta{..} <- x `Map.lookup` ms
+         , Just Refl <- nattyEqEh n (vlen mctxt)
+         , Just Refl <- nattyEqEh n (vlen ctx) -> do
+           rhs' <- normalise mctxt mtype (nattily n (wrapMeta x))
+           pure (rhs == rhs')
+       _ -> pure False
     case debug ("CONSTRAIN " ++ show lhs ++ " = " ++ show rhs) (lhs, rhs) of
       _ | debug "Checking syntactic equality?" (lhs == rhs) -> debug ("Passed syn eq for " ++ show lhs ++ " =?= " ++ show rhs) (pure $ SolvedIf [])
[…]

### 2024-03-19 src/MagicStrings.hs

@@ -21,8 +21,6 @@ pattern Splus = "plus"
 pattern Sone = "one"
 pattern Shjux = "hjux"
 pattern Svjux = "vjux"
--- TODO: cannonical multiplication, remove?
-pattern Smult = "mult"
 
 
 -- Labmate user types

### 2024-03-19 src/Term.hs

@@ -42,8 +42,10 @@ data Ctor (s :: Sort) (t :: Sort) where
   R :: Ctor (Prd Chk Chk) Syn
   -- tuples
   T :: Ctor (Prd Chk Chk) Chk
-  -- desctructors
+  -- unary destructors
   D :: Ctor (Prd Syn Chk) Syn
+  -- matrix mul destructor
+  MX :: Ctor (Prd Syn Syn) Syn
   -- metavar usage (these are unknowns, not schematic vars)
   M :: (Name, Natty k) -> Ctor (Sub k) Syn
   -- subst
@@ -114,6 +116,7 @@ cmpCtor E E = EQ' Refl
 cmpCtor R R = EQ' Refl
 cmpCtor T T = EQ' Refl
 cmpCtor D D = EQ' Refl
+cmpCtor MX MX = EQ' Refl
 cmpCtor (M (n, k)) (M (n', k')) = case truck ("MetaCmp " ++ show n ++" =?= " ++ show n') $ compare n n' of
   LT -> LT'
   GT -> GT'
@@ -129,7 +132,7 @@ cmpCtor t t' = case compare (helper t) (helper t') of
     helper = \case
       { A{} -> 1; I{} -> 2; E -> 3; R -> 4
       ; T -> 5; D -> 6; M{} -> 7; S0 -> 8
-      ; ST{} -> 9; SL{} -> 10; DL{} -> 11 }
+      ; ST{} -> 9; SL{} -> 10; DL{} -> 11; MX -> 12 }
 
 
 -- these instances should only be used for Norm
@@ -341,6 +344,8 @@ tmShow b (T :$ P tl u tr :^ th) ctx =
     s = tmShow False (tl :^ covl u -< th) ctx ++ tmShow True (tr :^ covr u -< th) ctx
 tmShow b (D :$ tm :^ th) ctx = let (a, d) = split (tm :^ th)
   in concat [barIf b, tmShow False a ctx, "(", tmShow False d ctx, ")"]
+tmShow b (MX :$ tm :^ th) ctx = let (a, d) = split (tm :^ th)
+  in concat [barIf b, "(", tmShow False a ctx, " * ", tmShow False d ctx, ")"]
 -- TODO : add the remaining cases
 
 barIf :: Bool -> String

### 2024-03-20 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] paper outline

--- 2023-05-26 writing/AMCTM2023/abstract.tex
@@
 
-\newcommand{\lr}{\textsc{LabMate}}
-\newcommand{\repo}{\url{https://github.com/msp-strath/LabMate}}
-\newcommand{\ma}{\textsc{Matlab}}
 \newcommand{\remph}{\emph}
@@
 
-\title{\lr --- an interactive assistant for \ma}
-\author{Conor McBride, Georgi Nakov, and Fredrik Nordvall Forsberg}
-\address{University of Strathclyde}
+\title{LabMate: supporting types for MATLAB}
+\author{Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Neil Ghani, Alasdair Forbes, Keith Lines, Ian Smith}
+\address{University of Strathclyde, National Physical Laboratory}
+%TODO: Affiliations
 
@@
 
-\paragraph{Introduction}
+\section{Introduction}
 
-We set out the prospectus for and report our progress on \lr, an interactive computer system designed to help human users of \ma{} write programs which actively make more of the sense those humans intend.
-This work is supported by the UK's National Physical Laboratory, and is significantly informed by a code corpus arising from their work, along with observations of their praxis.
-The latter is solid from an engineering perspective, with a discipline of documenting the structure and significance of each variable in \remph{comments} which, although not formal, exhibit a certain regularity of format.
-That is to say, the humans communicate to \remph{one another} information vital for comprehension of their software, whilst withholding the very same from the \remph{machine}.
-We can and should do better.
+% types help
 
-%\paragraph{Ulterior motive}
+% at low cost  (ie: don't need to start from scratch)
 
-Our goal is to enhance practitioners' experience of their existing toolchains, rather than to insist on the adoption of a radically different language prototype.
-\lr{} gives a formal language of comments for \ma{} programs, and can thus help both preventatively (with safety checks ahead of execution) and constructively (generating correct code from richer information).
-Our approach is inspired by Andrew Kennedy's work at Facebook on adding types incrementally to PHP~\cite{hack}.
+% Matlab popular etc
 
-\paragraph{Directives and Responses}
+\section{LabMate in action}
 
-\lr{} introduces a language of \emph{directives}:
-\[\pct\dir\;\mathit{directive}
-\]
-which \ma{} treats as comments, but \lr{} acts on, transforming code with directives to a new version of that code in four ways:
-\begin{itemize}
-\item documenting \remph{types} of functions and variables, giving not only data representation, but also richer metadata, e.g., units of measure;
-\item querying the type of a part of the program, or a gap in it;
-\item requesting the systematic transformation of the user's code, e.g., consistently renaming all occurrences of a given variable;
-\item generating low-level code serving some stated high-level purpose on the basis of supplied metadata, e.g., to read a data set from a file.
-\end{itemize}
-Programmers can run \lr{} iteratively on their work in progress, gaining support, feedback, and the alleviation of systematic drudgery.
-Controlled only by the directives in its input, \lr\ may readily be integrated with existing editors.
-It communicates back only via its output, including \emph{responses} (comments with useful information)
-\[
-\pct\res\;\mathit{response}
-\]
-and \emph{syntheses} (code segments unlikely to merit inspection)
-\[\begin{array}{l}
-\pct\res\lcb\qquad \mbox{(which is a \remph{single line} \ma{} comment)}\\
-\mathit{synthesized}\; \ma\;\mathit{code}\\
-\pct\res\rcb\qquad \mbox{(also a \remph{single line} \ma{} comment)} \\
-\end{array}\]
+% simple matrix multiplication example
 
-\paragraph{\lr{} types}
+\section{Types for matrices}
 
-The type of a \ma{} matrix is given by, at worst, its sizes in each dimension and cell type.
-However, a size is but a number and thus a special case of a list of \remph{headers}, such as you might find in a spreadsheet: the type of each cell may depend on the headers which govern its position.
-A key use of this refinement takes headers to be \remph{units}: \lr{} will check that the units of each cell respect these headers so that inner products computed during matrix multiplication are dimensionally meaningful~\cite{dimTypes}.
+% matrix types, and their motivation
 
-\paragraph{Current status}
+% typechecking
 
-\lr{} is in development and available at \repo.
-We can currently parse most \ma{} code we have encountered in the wild, and have implemented directives for renaming variables and generating code for reading data sets from files based on \textsc{MGen} input descriptions~\cite{mgen}.
+\section{Implementation}
+
+% transducer model
+
+% elaboration into core type theory (with richer equational theory than state-of-the-art)
+
+% Stack machine, with disjunctive problem solving
+
+\section{Supporting dimensional consistency}
+
+% matrix multiplication with units of measure
+
+\section{Conclusions}
+
+% future work etc
 
 \bibliographystyle{ws-procs9x6}
-\bibliography{labrat}
+%\bibliography{labmate}
 
 \end{document}
+
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: t
+%%% End:

### 2024-03-20 src/Machine.hs

@@ -31,7 +31,7 @@ import qualified Data.Set as Set
 import Debug.Trace
 import Control.Monad.Reader
 
-debug = trace --trace
+debug = const id --trace
 debugCatch = const id --trace
 
 type Elab = State MachineState

### 2024-03-21 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] rough draft Section 3

@@
 \usepackage{microtype}
+\usepackage{proof}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\rsb}{\texttt{\symbol{93}}}
+
+\newcommand{\istype}[1]{#1\ \textsf{type}}
+\newcommand{\isadd}[1]{#1\ \textsf{numeric}}
+\newcommand{\ismult}[3]{(#1, #2, #3)\ \textsf{multipliable}}
+
+\newcommand{\append}{+\!\!\!+}
+\newcommand{\hjux}[2]{{#1}|{#2}}
+\newcommand{\vjux}[2]{\frac{#1}{#2}}
+
+\newcommand{\One}{\mathbf{1}}
+\newcommand{\Matrix}[5]{\mathsf{Matrix}\,#1\,#2\,#3\,#4\,#5}
+\newcommand{\List}[1]{\mathsf{List}\,#1}
 
@@
 
-% matrix types, and their motivation
+Under the hood, LabMate translates MATLAB expressions to typed
+expressions in its own core type theory.
+%
+Since matrices feature heavily in MATLAB code, the type of matrices
+play a central role in this type theory.
+%
+We make this type rather precise, by allowing the type of each entry
+in the matrix to vary according to the row and column of the entry:
+%
+we parameterise the type of matrices by a type $R$ associated with
+each row, a type $C$ associated with each column, and a type $E(x,y)$
+which might vary depending on the values of parameters $x : R$ and
+$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
+idea being that the $i$th entry in the list $rs$ and the $j$th entry
+in the list $cs$ determines the type of the $(i, j)$ coordinate in the
+matrix.
+%
+Summarised as an inference rule, matrix types are thus formed as follows:
+%
+\[
+  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
+  {
+    \istype{R}
+    &\quad
+    \istype{C}
+    &\quad
+    x : R, y : C \vdash \istype{E(x,y)}
+    &\quad
+    rs : \List{R}
+    &\quad
+    cs : \List{C}
+  }
+\]
+
+A notable special case is when both $R$ and $C$ are the unit type
+$R = C = \One$, that is, the type with exactly one element.
+%
+In this case, there is only one type of entries $E$ (because both $x$
+and $y$ must be the unique element of the unit type), and the only
+information available in the lists $rs : \List{\One}$ and
+$cs : \List{\One}$ is their length.
+%
+Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
+$m \times n$ matrices with entries of type $E$, where $n$ and $m$ are
+the unique lists over the unit type of length $n$ and $m$
+respectively.
+
+Another noteworthy special case which we want to support takes $R$ and
+$R$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
+type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
+%
+This way, we can reduce dimensional consistency checking \`a la
+Kennedy~\cite{kennedyUOM} also for programs involving matrices,
+following the work of Hart~\cite{hart}.
+%
+We will see in \sref{sec:example-revisited} how this can be helpful.
+
+Terms of matrix type in the core type theory are built up from
+$1 \times 1$ matrices and pasting matrices horizontally and vertically.
+%
+Singleton matrices have the following typing rule
+\[
+  \infer{[e] : \Matrix{R}{C}{E}{[r]}{[c]}}
+    {r : R
+      &\quad
+      c : C
+      &\quad
+      e : E(r, c)
+    }
+\]
+%
+while horizontal and vertical pasting horizontally have the following
+typing rules:
+\[
+  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{rs}{(cs_A \append cs_B)}}
+  {
+    A : \Matrix{R}{C}{E}{rs}{cs_A}
+    &\quad
+    B : \Matrix{R}{C}{E}{rs}{cs_B}
+  }
+\]
+\[
+  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(rs_A \append rs_B)}{cs}}
+  {
+    A : \Matrix{R}{C}{E}{rs_A}{cs}
+    &\quad
+    B : \Matrix{R}{C}{E}{rs_B}{cs}
+  }
+\]
+When comparing matrices for equality, we reassociate horizontal and
+vertical pastings, so that $\hjux{\vjux{A}{C}}{\vjux{B}{D}}$ and
+$\vjux{\hjux{A}{B}}{\hjux{C}{D}}$ indeed are considered equal.
+
+Two matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
+numeric (for example integers, or floating point numbers), and if $A$
+and $B$ have the same matrix type:
+%
+\[
+  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
+  {
+    x : R, y : C \vdash \isadd{E(x, y)}
+    &\quad
+    A : \Matrix{R}{C}{E}{rs}{cs}
+    &\quad
+    B : \Matrix{R}{C}{E}{rs}{cs}
+  }
+\]
+%
+Matrix multiplication is trickier, because it does not demand that $A$
+and $B$ have the same type --- rather, it demands that $A$ and $B$
+have \remph{compatible} types:
+%
+\[
+  \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
+  {
+    \ismult{E}{E'}{E''}
+    &\quad
+    A : \Matrix{R}{C}{E}{rs}{ms}
+    &\quad
+    B : \Matrix{R}{C}{E'}{ms}{cs}
+  }
+\]
+The entry types $E$, $E'$, $E''$ are multipliable if for all $x$, $y$
+and $z$, we have that $E(x, y)$ and $E'(y, z)$ supports a
+multiplication operation with codomain $E''(x, z)$.
+%
+This is certainly the case if all entry types involved are simply
+integers or floating point numbers, but this constraint can also
+ensure that the physical units involved line up.
 
@@
 \section{Supporting dimensional consistency}
+\label{sec:example-revisited}
 
@@
 \bibliographystyle{ws-procs9x6}
-%\bibliography{labmate}
+\bibliography{labmate}
 

### 2024-03-26 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fantasy part of Section 2

@@
 \usepackage{proof}
+\usepackage{mathtools}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \newcommand{\remph}{\emph}
+
+\newcommand{\todo}[1]{\textcolor{red}{\textbf{TODO: #1}}}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 % simple matrix multiplication example
+
+Consider the problem of resistance calibration: \todo{explain here}.
+
+% $t_i$ time ($\mathsf{T}$; in seconds); $v_i$ voltage ($\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$; in volt)
+
+\[
+\begin{pmatrix*}
+  t_1 & 1 & -1 & 0 & 0 \\
+  t_2 & 1 & -1 & 0 & 0 \\
+  t_3 & 1 & -1 & 0 & 0 \\
+  t_4 & 1 & -1 & 0 & 0 \\
+  t_5 & 1 & -1 & 0 & 0 \\
+  t_6 & 0 & 0 & 1 & -1 \\
+  t_7 & 0 & 0 & 1 & -1 \\
+  t_8 & 0 & 0 & 1 & -1 \\
+  t_9 & 0 & 0 & 1 & -1 \\
+\end{pmatrix*}
+\times
+\begin{pmatrix}
+  m \\
+  c \\
+  a \\
+  c_{\text{offset}} \\
+  a_{\text{offset}}
+\end{pmatrix}
+=
+\begin{pmatrix}
+  v_1 \\
+  v_2 \\
+  v_3 \\
+  v_4 \\
+  v_5 \\
+  v_6 \\
+  v_7 \\
+  v_8 \\
+  v_9
+\end{pmatrix}
+\]
+
+This can be converted into a MATLAB problem with LabMate annotations as follows:
+\begin{verbatim}
+%> A :: [ 9 x 5 ] double
+A = [ t1 1 -1 0 0
+      t2 1 -1 0 0
+      t3 1 -1 0 0
+      t4 1 -1 0 0
+      t5 1 -1 0 0
+      t6 0 0 1 -1
+      t7 0 0 1 -1
+      t8 0 0 1 -1
+      t9 0 0 1 -1 ]
+
+%> B :: [ 5 x 1 ] double
+B = [ m c a cOffset aOffset ]
+
+C = A * B
+
+%> typeof C
+\end{verbatim}
+%
+\todo{Is computing C from A and B what we want?}
+The lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{input directives} by LabMate.
+LabMate acts by reading the source code and outputting it again, but possibly with some additional \remph{responses} to the directives added.
+
+Did you notice the error in the above code?
+The matrix \texttt{B} should be of size $5 \times 1$, but is erroneously of size $1 \times 5$ instead.
+LabMate detects this error and produces the following as output, where we have suppressed the same definition of \texttt{A} as above:
+\begin{verbatim}
+%> A :: [ 9 x 5 ] double
+A = ...
+
+%> B :: [ 5 x 1 ] double
+B = [ m c a cOffset aOffset ]
+
+C = A * B
+
+%> typeof C
+%< the type of C is quite a puzzle; unsolved constraints 5 =? 1, 1 =? 5
+\end{verbatim}
+%
+That is, LabMate says that for \texttt{C} to have a sensible type, we need the (impossible) constraints $5 = 1$ and $1 = 5$ to be true, arising from the fact that we have declared \texttt{B} to be a $5 \times 1$ matrix, but in fact it is a $1 \times 5$ matrix.
+Note that the original source code has been retained, but a response has been inserted to our \texttt{typeof} query.
+If we change \texttt{B} to
+\begin{verbatim}
+B = [ m; c; a; cOffset; aOffset ]
+\end{verbatim}
+instead, then LabMate responds
+\begin{verbatim}
+%< C :: [ 9 x 1 ] double
+\end{verbatim}
+This way, LabMate can detect matrix size errors at development time, rather than at runtime.
 
@@
 
+GravCalc example again: $t_i$ time ($\mathsf{T}$; in seconds); $v_i$ voltage ($\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$; in volt)
+
 \section{Conclusions}

### 2024-03-26 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fantasy draft of Section 5

@@
 \section{LabMate in action}
+\label{sec:example}
 
@@
 
-GravCalc example again: $t_i$ time ($\mathsf{T}$; in seconds); $v_i$ voltage ($\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$; in volt)
+Coming back to the example from \sref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
+\begin{verbatim}
+%> A :: [ i <- 9,  j <- 5 ] (if j == 0 then Q T else Q 1)
+%> B :: [ Q TODO; Q TODO; Q TODO; Q TODO; Q TODO; Q TODO ]
+\end{verbatim}
+This says that the entries in the first column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities.
+Meanwhile the entries of \texttt{B} are quantities of dimension \todo{fill in} respectively.
+LabMate reports that the type of \texttt{C} should be
+\begin{verbatim}
+%< C :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
+\end{verbatim}
+that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$.
+Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{C}, and to leave the type of one of the entries in \texttt{B} as an unknown:
+\begin{verbatim}
+%> B :: [ Q TODO; Q TODO; ?aTy; Q TODO; Q TODO; Q TODO ]
+\end{verbatim}
+We can then ask LabMate to solve the \remph{meta variable} \texttt{?aTy}, to which LabMate responds that we must have \todo{\texttt{?aTy} = fill in}.
+This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.
+
+%GravCalc example again: $t_i$ time ($\mathsf{T}$; in seconds); $v_i$ voltage ($\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$; in volt)
 

### 2024-03-27 examples/rescal.m
[ broken ] add simple versions of res cal examples

@@ -1,0 +1,17 @@
+%> A :: [ i <- 9,  j <- 5 ] (if j == 0 then Q T else Q 1)
+A = [ t1 1 -1 0 0
+      t2 1 -1 0 0
+      t3 1 -1 0 0
+      t4 1 -1 0 0
+      t5 1 -1 0 0
+      t6 0 0 1 -1
+      t7 0 0 1 -1
+      t8 0 0 1 -1
+      t9 0 0 1 -1 ]
+
+%> B :: [ 5 x 1 ] int
+B = [ m; c; a; cOffset; aOffset ]
+
+C = A * B
+
+%> typeof C

### 2024-03-27 src/CoreTT.hs
[ broken ] broken matrix elab; error in prefixEh

@@ -67,9 +67,9 @@ instance Monad (TC n) where
 instance MonadFail (TC n) where
   fail  = TC . const . const . Left
 
-must :: MonadFail m => Maybe a -> m a
-must (Just a) = pure a
-must Nothing = fail "must fails"
+must :: MonadFail m => String -> Maybe a -> m a
+must _ (Just a) = pure a
+must s Nothing = fail $ s ++ " must fail"
 
 typeOf :: (S Z <= n) -> TC n (Type ^ n)
 typeOf i = TC $ \_ -> Right . snd . vonly . (i ?^)
@@ -198,7 +198,7 @@ checkCanMatrixEh
 {-
      <------------cs------------>
 
- ^   X------------------cs1----->
+ ^   X-----------X------cs1----->
  |   |           |
  |   |           |
  |   |          rs0
@@ -285,11 +285,11 @@ prefixEh elty pr l = withScope $ propEh elty >>= \case
     True <- track ("NFAbel pr=" ++ show pr) $ pure True
     True <- track ("NFAbel  l=" ++ show l) $ pure True
     guard $ n <= m
-    nfAbelToTerm . NFAbel (m - n) <$> must (sub tns tms)
+    nfAbelToTerm . NFAbel (m - n) <$> must "prefixEh:nfAbelToTerm" (sub tns tms)
   False -> do
     pr <- termToNFList elty pr
     l  <- termToNFList elty l
-    nfListToTerm <$> must (stripPrefix pr l)
+    nfListToTerm <$> must "prefixEh:nfListToTerm" (stripPrefix pr l)
   where
     sub :: [(Norm Chk ^ n, Integer)]
         -> [(Norm Chk ^ n, Integer)]
@@ -468,7 +468,7 @@ evalSynth tm = withScope $ case tm of
   tm | Just tm <- MX $? tm, (lmx, rmx) <- split tm -> do
     (lmx, lmxTy) <- evalSynth lmx
     (rmx, rmxTy) <- evalSynth rmx
-    -- TODO: FIXME
+    -- TODO: FIXME, compute the cell type from lcellTy, rcellTy
     retTy <- case (tagEh lmxTy, tagEh rmxTy) of
       (Just (SMatrix, [rowTy, midTy, lcellTy, rs, _]), Just (SMatrix, [_, colTy, rcellTy, _, cs])) ->
[…]

### 2024-03-27 src/Lex.hs

@@ -448,6 +448,10 @@ nonce = fst . source
 
 noSource :: a -> WithSource a
 noSource a = a :<=: (-1, Hide [])
+
+realSourceEh :: Source -> Bool
+realSourceEh (n, _) = n >= 0
+
 {-
 instance Show a => Show (WithSource a) where
   show (a :<=: (n, src)) = "$" ++ show n ++ " = `" ++ (src >>= showEaten) ++ "`" ++ show a

### 2024-03-27 src/Machine.hs

@@ -33,6 +33,7 @@ import Control.Monad.Reader
 
 debug = const id --trace
 debugCatch = const id --trace
+debugMatrix = trace
 
 type Elab = State MachineState
 
@@ -209,7 +210,6 @@ winning = Winning True mempty
 worried :: Monoid a => Mood' a
 worried = Winning False mempty
 
-
 data DeclarationType a
   = UserDecl
   { varTy :: a             -- (eventual) type
@@ -339,13 +339,17 @@ postRight fs = pull >>= \case
     getStatus (Fork frk) = either fstatus fstatus frk
     getStatus _ = mempty
 
-
 switchFwrd :: ([Frame], Problem) -> Elab ([Frame], Problem)
 switchFwrd (fs', p') = do
   st@MS{..} <- get
   let fz :<+>: fs = position
   put $ st { position = fz :<+>: fs', problem = p' }
   pure (fs, problem)
+
+pushSource :: Source -> Elab ()
+pushSource src
+  | realSourceEh src = push $ Source src
+  | otherwise = pure ()
 
 prob :: Elab Problem
 prob = llup >>= \case
@@ -510,7 +514,7 @@ ctx \\\\ x = fmap (fmap wk) <$> ctx :# x
 
 normalise :: Context n -> Type ^ n -> Term Chk ^ n -> Elab (Norm Chk ^ n)
 normalise ctx ty tm  = elabTC ctx (checkEval ty tm) >>= \case
-  Left err -> error $ "normalise : TC error" ++ err
+  Left err -> nattily (vlen ctx) $ error . concat $ ["normalise: TC error ", err, " for ", show tm, "::", show ty]
   Right tm -> pure tm
 
 constrain :: String -> Constraint -> Elab ConstraintStatus
@@ -731,7 +735,7 @@ onNearestSource :: ([Tok] -> [Tok]) -> Elab ()
 onNearestSource f = excursion go where
   go = pull >>= \case
     Nothing -> error "Impossible: no enclosing Source frame"
-    Just (Source (n, Hide ts)) -> push $ Source (n, Hide $ f ts)
+    Just (Source (n, Hide ts)) -> pushSource (n, Hide $ f ts)
     Just f -> shup f >> go
 
 metaDecl :: Status -> String -> Context n -> Type ^ n -> Elab Name
@@ -809,16 +813,16 @@ run = prob >>= \case
       move winning
   Expression (App (f :<=: src) args) -> do
[…]

### 2024-03-27 src/Parse/Matlab.hs

@@ -150,7 +150,7 @@ ptypeexpr ci = go >>= more ci where
                _ -> mempty)
    <|> pws (TyStringLiteral <$> pstringlit)
   lhsstuff ci = (pws (TyVar <$> pnom) >>= lmore)
-            <|> pws (TyMat <$> pgrp (== Bracket Square) (many (prow ptypeexpr)))
+            <|> pws (tyMat <$> pgrp (== Bracket Square) (many (prow ptypeexpr)))
   lmore l = pcond
               (pws' (nonce l) (TyApp l <$ pcxspc ci <*> pargs Round (ptypeexpr topCI)))
               lmore

### 2024-03-27 src/Syntax.hs

@@ -48,16 +48,32 @@ data TensorType'
   = Tensor ((String, TypeExpr), (String, TypeExpr)) TypeExpr
   deriving Show
 
+data VOrH = Vertical | Horizontal
+  deriving Show
+
 -- possibly incomplete list of type level expressions
 data TypeExpr'
   = TyVar String -- might also be constants, e.g. Double
   | TyNum Int
   | TyApp TypeExpr [TypeExpr]
-  | TyMat [[TypeExpr]]
+  -- | TyMat [[TypeExpr]]
+  | TyJux VOrH TypeExpr TypeExpr
+  | TyNil VOrH
   | TyBinaryOp BinOperator TypeExpr TypeExpr
   | TyUnaryOp UnOperator TypeExpr
   | TyStringLiteral String
   deriving Show
+
+tyMat :: [[TypeExpr]] -> TypeExpr'
+tyMat exps = what $ go Vertical (go Horizontal id) exps
+  where
+    jux _ e (TyNil _ :<=: _) = e
+    jux dir e e' = noSource $ TyJux dir e e'
+
+    nil dir = noSource $ TyNil dir
+
+    go :: VOrH -> (a -> TypeExpr) -> [a] -> TypeExpr
+    go dir pre = foldr (jux dir . pre) (nil dir)
 
 type Res = [Tok]
 
@@ -87,12 +103,14 @@ toTypeExpr' (Var x) = pure $ TyVar x
 toTypeExpr' (IntLiteral n) = pure $ TyNum n
 toTypeExpr' (StringLiteral s) = pure $ TyStringLiteral s
 toTypeExpr' (BinaryOp op x y) = TyBinaryOp op <$> toTypeExpr x <*> toTypeExpr y
+toTypeExpr' (Mat exps) = do
+   exps <- traverse (traverse toTypeExpr) exps
+   pure $ tyMat exps
 -- FIXME : add more
 toTypeExpr' _ = Nothing
 
 toTypeExpr :: Expr -> Maybe TypeExpr
 toTypeExpr = traverse toTypeExpr'
-
 
 data LHS'
   = LVar String

### 2024-03-27 examples/matrixmul1.m
[ fix ] some progress on matrix elab

@@ -1,0 +1,5 @@
+%> A :: [ 3 x 1 ] int
+A = [ 2
+      4
+      6 ]
+%> typeof A

### 2024-03-27 src/CoreTT.hs

@@ -576,7 +576,7 @@ checkEvalMatrixNF nf ty@(rowTy, colTy, cellTy) mx@(rs, cs) tm
     (rm, ((_, cs1), (_, cs''))) <- checkEvalMatrixNF nf ty rt r
     pure (lm `hjux` rm,  ((rs', mk Splus cs0 cs1), (rs0, cs'')))
   | Just (Svjux, [t, b]) <- tagEh tm = withScope $ do
-    (tm, (lb@(_, cs0), (rs0, cs'))) <- checkEvalMatrixNF nf ty mx t
+    (tm, (lb@(_, cs0), (rs0, cs'))) <- track ("checkEval VJUX " ++ show tm) $ checkEvalMatrixNF nf ty mx t
     (bm, ((rs'', _), (rs1, _))) <- checkEvalMatrixNF nf ty lb b
     pure (tm `vjux` bm, ((rs'', cs0), (mk Splus rs0 rs1, cs')))
   | Just tm <- E $? tm = withScope $ do
@@ -585,7 +585,7 @@ checkEvalMatrixNF nf ty@(rowTy, colTy, cellTy) mx@(rs, cs) tm
        Nothing -> checkEvalMatrixNF nf ty mx tm
        Just tm -> case tagEh ty' of
          Just (SMatrix, [_, _, _,  rs0, cs0]) -> do
-           rs1 <- prefixEh rowTy rs0 rs
+           rs1 <- track ("Are we here " ++ show tm) $ prefixEh rowTy rs0 rs
            cs1 <- prefixEh colTy cs0 cs
            h   <- nf rowTy rs0
            pure ([(h, [NFNeutral tm])], ((rs1, cs0), (rs0, cs1)))

### 2024-03-27 src/Machine.hs

@@ -31,7 +31,7 @@ import qualified Data.Set as Set
 import Debug.Trace
 import Control.Monad.Reader
 
-debug = const id --trace
+debug = trace
 debugCatch = const id --trace
 debugMatrix = trace
 
@@ -688,8 +688,8 @@ constrain s c = do
       Unstarted -> SolvedIf [s]
       _ -> constraintStatus c
 
-constrainEqualType :: TYPE -> TYPE -> Elab ()
-constrainEqualType lhs rhs = (() <$) . constrain "Q" $ Constraint
+constrainEqualType :: TYPE -> TYPE -> Elab ConstraintStatus
+constrainEqualType lhs rhs = constrain "Q" $ Constraint
   { constraintCtx   = VN
   , constraintType  = Hom (atom SType)
   , constraintStatus = Unstarted
@@ -709,7 +709,7 @@ findDeclaration UserDecl{varTy = ty, currentName = old, seen, newNames = news} =
       Locale (FunctionLocale _) -> shup f >> go True -- we know we are in a function
       Declaration n u'@UserDecl{varTy = ty', currentName = old', seen = seen', newNames = news'} | old == old' -> do
         case ty of
-          Just ty -> constrainEqualType ty ty'
+          Just ty -> () <$ constrainEqualType ty ty'
           Nothing -> pure ()
         Just (n, ty') <$ push (Declaration n u'{seen = seen' || seen, newNames = news' ++ news})
       _ -> shup f >> go b
@@ -991,19 +991,26 @@ elab x ctx ty (etask :<=: src) =
   fmap (Sourced . (:<=: src)) <$> elab' x ctx ty etask
 
 
-ensureMatrixType :: Context n -> NmTy ^ n -> Elab (NmTy ^ n, NmTy ^ n, NmTy ^ n, Norm Chk ^ n, Norm Chk ^ n)
+ensureMatrixType :: Context n -> NmTy ^ n -> Elab (NmTy ^ n, NmTy ^ n, NmTy ^ n, Norm Chk ^ n, Norm Chk ^ n, ConstraintStatus)
 ensureMatrixType ctxt ty
   | Just (SMatrix, [rowTy, colTy, cellTy, rs, cs]) <- tagEh ty
   {- , Just cellTy <- lamEh cellTy
-  , Just cellTy <- lamEh cellTy -} = pure (rowTy, colTy, cellTy, rs, cs)
[…]

### 2024-03-28 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] draft intro and conclusions

@@
 \title{LabMate: supporting types for MATLAB}
-\author{Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Neil Ghani, Alasdair Forbes, Keith Lines, Ian Smith}
+\author{Neil Ghani, Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Alasdair Forbes, Keith Lines, Ian Smith}
 \address{University of Strathclyde, National Physical Laboratory}
@@
 
-% types help
+MATLAB is a key workhorse in many scientific and engineering
+disciplines that are heavily reliant on numerical methods. It helps us
+do powerful things. However, as with all software, MATLAB code may
+contain implementation errors and bugs. In good programming practice
+for MATLAB developers, programmers use comments to make clear what
+physical systems their data concern and how the data should be
+interpreted, specifying, e.g., units of measure for
+quantities. Regrettably, none of this rich and often disciplined
+metadata is perceptible to MATLAB, which instead enforces the
+compatibility of producers and consumers of data by run-time checking
+of tags which indicate only machine representation, not any form of
+meaning. In reality, the programmers document meaning for each other's
+benefit but keep the machine in the dark.
 
-% at low cost  (ie: don't need to start from scratch)
-
-% Matlab popular etc
+To rectify this situation, we have developed LabMate, which is a tool
+to reify current virtuous engineering practices as a formal language
+of MATLAB comments. MATLAB remains in the dark, but LabMate reads,
+assesses, and transforms MATLAB programs in accordance with these
+comments.
+%
+Behind the scenes, LabMate is retrofitting an expressive type system
+to MATLAB, with the more of the meaning of programs recorded in their
+types.
+%
+This gives a lightweight and low-cost way for MATLAB programmers to
+express their intent, and in a language they are already working in,
+rather than starting over from scratch.
+%
+While type systems and their theory is a well established discipline
+of computer science, the development of LabMate has a several novel
+advances novel focus on the algebraic structure required to classify
+matrices and the meanings of the quantities therein, e.g., their
+physical dimensions. LabMate brings advanced type systems to
+effectiveness within, rather than instead of, existing scientific and
+engineering toolchains and practices.
 
@@
 
-\section{Conclusions}
+\section{Conclusions and future work}
 
-% future work etc
+We have demonstrated how LabMate can help write MATLAB programs that are correct both in the sense that they avoid runtime errors, as well as dimensional inconsistency for physical dimensions.
+%
+Getting an error for types not matching can sometimes feel like a stick, but there are also plenty of opportunities for carrots, for example where the specification encoded in the type of the program admits a single unique solution, which could be generated automatically by LabMate.
+%
+Similarly, LabMate already has support for generating MATLAB code for serialising and deserialising complex data formats to and from files, based on a more high-level description of the data format (for example also taking units of measure into account)~\cite{mgen}.
+
+So far, our focus has been on matrices, as this is what makes MATLAB
+unique as a programming language.
+%
+In the future, we would like to also extend our coverage to
+conditionals and loops, which might necessitate some approximation of
+what is known about program fragments at runtime.
 

### 2024-03-28 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] draft Implementation section

@@
 
-% transducer model
+In contrast to most other typecheckers, LabMate works using a
+\remph{transducer} model of interaction: it inputs MATLAB code with
+formal comments, and outputs a new version of its input, responding to
+the comments, as if it were a development collaborator.  It thus needs
+to make sure that it retains as much of the input formatting as
+possible.  Making sense of MATLAB code presented a significant reverse
+engineering challenge, as the MATLAB syntax is largely specified
+informally and by example.
 
-% elaboration into core type theory (with richer equational theory than state-of-the-art)
+LabMate elaborates MATLAB expressions and commands into terms of its own internal core type theory.
+%
+Because we want to ergonomically handle physical dimensions, matrices
+and their types, this type theory has a richer equational theory than
+other state-of-the-art type theories, which we achieve by a refined
+normalisation algorithm for free Abelian groups, and free monoids.
+%
+The actual typechecking is implemented as a stack based virtual machine which uses unification and a limited form of backtracking to solve typing constraints.
+%
+After the virtual machine has solved as many problems as it can, we analyse its final state to reconstruct the output source code, perhaps with inserted warnings and error messages, or additional responses to queries.
 
-% Stack machine, with disjunctive problem solving
+LabMate is under constant development; the latest version can be found at \url{https://github.com/msp-strath/LabMate/tree/devel}.
 

### 2024-03-28 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fix minor typos

@@
 of computer science, the development of LabMate has a several novel
-advances novel focus on the algebraic structure required to classify
+advances on the algebraic structure required to classify
 matrices and the meanings of the quantities therein, e.g., their
@@
 Since matrices feature heavily in MATLAB code, the type of matrices
-play a central role in this type theory.
+plays a central role in this type theory.
 %

### 2024-04-03 src/CoreTT.hs
[ new ] Boolean type (without normalisation)

@@ -96,6 +96,7 @@ metaLookup s = TC $ \ st _ -> case s `Map.lookup` st of
 typeEh :: Term Chk ^ n -> TC n ()
 typeEh ty | Just cts <- tagEh ty = case cts of
   (SOne, [])   -> pure ()
+  (STwo, [])   -> pure ()
   (SAtom, [])  -> pure ()
   (SType, [])  -> pure ()
   (SChar, [])  -> pure ()
@@ -166,6 +167,7 @@ checkCanEh ty tm | Just x <- tagEh ty = withScope $ case x of
     nfs <- termToNFList (atom SAtom) as
     True <$ checkEnumEh nfs tm
   (SOne, []) -> pure True
+  (STwo, []) | Intg i <- tm, i == 0 || i == 1 -> pure True
   (SChar, []) | Intg i <- tm, i >= 0 && i <= 128 -> pure True
   (SAtom, []) | Atom _ <- tm -> pure True
   (SPi, [s, t]) | Just t' <- lamEh t, Just (x, tm') <- lamNameEh tm ->
@@ -367,6 +369,8 @@ checkNormEval wantTy tm | Just tm <- E $? tm = do
 checkNormEval ty tm | Just ty <- tagEh ty = withScope $ case ty of
   (SType, []) -> typeEval tm
   (SOne, []) -> pure nil
+  -- TODO: FIXME use reduced order BDD
+  (STwo, []) -> pure tm
   (SAtom, []) -> pure tm
   (SChar, []) -> pure tm
   (SAbel, [genTy]) -> nfAbelToTerm <$> termToNFAbel genTy tm
@@ -411,7 +415,7 @@ checkEval ty tm = do
 
 typeEval :: Type ^ n -> TC n (NmTy ^ n)
 typeEval ty | Just ty <- tagEh ty = withScope $ case ty of
-  (x, []) | x `elem` [SAtom, SOne, SType, SChar] -> pure $ atom x
+  (x, []) | x `elem` [SAtom, SOne, STwo, SType, SChar] -> pure $ atom x
   (SAbel, [genTy]) -> mk SAbel <$> typeEval genTy
   (SList, [genTy]) -> mk SList <$> typeEval genTy
   (SEnum, [as]) -> mk SEnum <$> checkNormEval (mk SList (atom SAtom)) as

### 2024-04-03 src/MagicStrings.hs

@@ -3,6 +3,7 @@ module MagicStrings where
 -- core types
 pattern SType = "Type"
 pattern SOne  = "One"
+pattern STwo  = "Two"
 pattern SAbel = "Abel"
 pattern SList = "List"
 pattern SAtom = "Atom"

### 2024-04-03 src/Machine.hs
[ wip ] moving constraints to the frame store

@@ -129,6 +129,7 @@ instance PrettyPrint Frame where
    where
      blat t | Non n <- kin t = '$': show n
      blat t = raw t
+  pprint (ConstraintFrame n c) = pure ["!!!!!CONSTRAINT FRAME!!!!!", show n ++ "  " ++ show c]
   pprint fr = pure [show fr]
 
 instance PrettyPrint MachineState where
@@ -166,6 +167,7 @@ data Frame where
     -> TERM -- rhs
     -> Frame
   LocalStore :: Store -> Frame
+  ConstraintFrame :: Name -> Constraint -> Frame
   deriving Show
 
 data Fork = MkFork
@@ -358,6 +360,9 @@ prob = llup >>= \case
     oldNs <- swapNameSupply ns
     push $ LocalNamespace oldNs
     prob
+  Just (ConstraintFrame name c) -> do
+    debug ("`prob` attempting to solve constrain " ++ show c) $ solveConstraint name c
+    prob
   Just f  -> do
     f <- normaliseFrame f
     push f
@@ -416,6 +421,7 @@ data ConstraintType' t
   = Hom t
   -- heterogenous constraint will become homogenous as soon as the
   -- named constraints have been solved
+  -- TODO: replace the list of names with a closed Boolean term
   | Het t [Name] t
   deriving (Show, Functor)
 
@@ -434,6 +440,9 @@ isHom :: ConstraintType n -> Maybe (Type ^ n)
 isHom (Hom ty) = Just ty
 isHom _ = Nothing
 
+
+-- TODO: use boolean metavariables instead, updating the
+-- status then boils down to a call to `normalise`
 data ConstraintStatus
   = Impossible
   | Blocked
@@ -494,6 +503,7 @@ mkConstraintType :: Type ^ n -> [Name] -> Type ^ n -> ConstraintType n
 mkConstraintType lty [] rty = Hom lty
 mkConstraintType lty names rty = Het lty names rty
 
+-- TODO: purge away
 getConstraintStatus :: Name -> Elab ConstraintStatus
 getConstraintStatus x = do
   cstore <- gets constraintStore
@@ -503,6 +513,7 @@ getConstraintStatus x = do
     Just (SolvedIf xs) -> mconcat <$> traverse getConstraintStatus xs
     Just cstatus -> pure cstatus
 
+-- TODO: purge away
[…]

### 2024-04-06 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] additions by Keith

@@
 \usepackage{mathtools}
+% Added by Keith L 05/04/2024
+% Removed by FNF 06/04/2024 (incompatible with style file)
+%\usepackage[]{biblatex}
+%\addbibresource{ws-procs9x6.bib}
+\usepackage{graphics}
+%\usepackage{caption}
+\usepackage{floatrow}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \title{LabMate: supporting types for MATLAB}
-\author{Neil Ghani, Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Alasdair Forbes, Keith Lines, Ian Smith}
+\author{Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Alistair Forbes, Keith Lines}
 \address{University of Strathclyde, National Physical Laboratory}
@@
 While type systems and their theory is a well established discipline
-of computer science, the development of LabMate has a several novel
+of computer science, the development of LabMate has several novel
 advances on the algebraic structure required to classify
@@
 
-Consider the problem of resistance calibration: \todo{explain here}.
+The following greatly simplified, but still realistic, example from metrology  will help illustrate the principles of how LabMate will work.
+Consider one of the calculations required for measuring resistors using a cryogenic current comparator bridge~\cite{Williams_2010}.
+
+An electrical current with a known value (the applied signal) is passed through the resistor being measured, and a series of voltage readings are taken at fixed time intervals (the recorded output).
+After a specified number of readings are taken, the direction of the input signal must be reversed in order to separate the offset and drift from the voltage readings.
+Therefore, the applied signal (bridge energisation) is a square wave of {\em reversals}.
+A calibration factor is also applied, dividing the signal into two parts with different amplitudes.
+
+The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~\ref{fig:rec_out} provides an example of recorded output, using simulated data.
+\begin{figure}[H]
+    \centering
+    \captionsetup{justification=centering}
+    \includegraphics[width=0.6\linewidth]{example_output.jpg}
+    \caption{Simulated example of recorded output}       
+    \label{fig:rec_out}
+\end{figure}
+
+The calculation of offset and drift is achieved using a least squares line fit calculated by solving a series of simultaneous equations, listed below.
+Note that although offset and drift differ for calibration and non-calibration parts of the input signal, the drift is the same in both cases.
 
@@
 \begin{pmatrix*}
-  t_1 & 1 & -1 & 0 & 0 \\
-  t_2 & 1 & -1 & 0 & 0 \\
-  t_3 & 1 & -1 & 0 & 0 \\
-  t_4 & 1 & -1 & 0 & 0 \\
-  t_5 & 1 & -1 & 0 & 0 \\
-  t_6 & 0 & 0 & 1 & -1 \\
-  t_7 & 0 & 0 & 1 & -1 \\
-  t_8 & 0 & 0 & 1 & -1 \\
-  t_9 & 0 & 0 & 1 & -1 \\
+  d_{\textrm{nc}1} & d_{\textrm{c}1} & n_1 & c_1 & t_1  \\
+  d_{\textrm{nc}2} & d_{\textrm{c}2} & n_2 & c_2 & t_2  \\
+   & \vdots &  & & \\
+  d_{\textrm{nc}N} & d_{\textrm{c}N} & n_N & c_N & t_N  \\
 \end{pmatrix*}
@@
 \begin{pmatrix}
-  m \\
+  a \\
+  a_{cal} \\ 
   c \\
-  a \\
-  c_{\text{offset}} \\
-  a_{\text{offset}}
+  c_{cal} \\ 
+  m
 \end{pmatrix}
@@
   v_2 \\
-  v_3 \\
-  v_4 \\
-  v_5 \\
-  v_6 \\
-  v_7 \\
-  v_8 \\
-  v_9
+  \vdots \\
+  v_N
 \end{pmatrix}
@@
 
-This can be converted into a MATLAB problem with LabMate annotations as follows:
+Where in the matrix:
+\begin{itemize}
+    \item The values in the first four columns are purely numeric and have no associated dimension.
+
+    \item The first column will contain either 1 or -1 for non-calibration voltage readings (i.e., those to which the calibration factor has not been applied) depending on the direction of the input signal.
+For calibration readings the value will be 0.
+
+    \item The second column contains equivalent values for voltage readings in which the calibration factor has been applied.
+I.e.
+0 for non-calibration readings and 1 or -1 depending on the input signal direction for calibration readings.
+
+    \item The third column contains 1 for non-calibration readings and 0 for calibration.
+
+    \item The forth column is the opposite of the third.
+
+    \item The fifth columns contains time values.
+\end{itemize}
+
+The first row vector contains values to be calculated using the line fit.
+The second column vector contains voltage readings.
+The measured value of the resistor is calculated using the contents of the first row vector.
+Details of that calculation are beyond the scope of this paper.
+
+This calculation can be converted into a MATLAB problem with LabMate annotations as follows.
+Note that for reasons of space only twelve values are listed.
+A real measurement would contain considerably more values:
 \begin{verbatim}
-%> A :: [ 9 x 5 ] double
-A = [ t1 1 -1 0 0
-      t2 1 -1 0 0
-      t3 1 -1 0 0
-      t4 1 -1 0 0
-      t5 1 -1 0 0
-      t6 0 0 1 -1
-      t7 0 0 1 -1
-      t8 0 0 1 -1
-      t9 0 0 1 -1 ]
+%> A :: [ 12 x 5 ] double
+A = [ 1  0  1  0  t1
+      1  0  1  0  t2
+      1  0  1  0  t3
+     -1  0  1  0  t4
+     -1  0  1  0  t5
+     -1  0  1  0  t6
+      0  1  0  1  t7
+      0  1  0  1  t8
+      0  1  0  1  t9
+      0 -1  0  1  t10
+      0 -1  0  1  t11
+      0 -1  0  1  t12 ]
 
 %> B :: [ 5 x 1 ] double
-B = [ m c a cOffset aOffset ]
+B = [ a aCal c cCal m ]
 
@@
 %
-\todo{Is computing C from A and B what we want?}
+\todo{Is computing C from A and B what we want? \\ Keith 05/04/2024 ***No. We are computing B from A and C using a least squares line fit.*** }
 The lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{input directives} by LabMate.
@@
 \begin{verbatim}
-%> A :: [ 9 x 5 ] double
+%> A :: [ 12 x 5 ] double
 A = ...
@@
 %> B :: [ 5 x 1 ] double
-B = [ m c a cOffset aOffset ]
+B = [ a aCal c cCal m ]
 
@@
 \begin{verbatim}
-B = [ m; c; a; cOffset; aOffset ]
+B = [ a; aCal; c; cCal; m ]
 \end{verbatim}
@@
 \begin{verbatim}
-%< C :: [ 9 x 1 ] double
+%< C :: [ 12 x 1 ] double
 \end{verbatim}
@@
 \begin{verbatim}
-%> A :: [ i <- 9,  j <- 5 ] (if j == 0 then Q T else Q 1)
-%> B :: [ Q TODO; Q TODO; Q TODO; Q TODO; Q TODO; Q TODO ]
+%> A :: [ i <- 12,  j <- 5 ] (if j == 4 then Q T else Q 1)
+%> B :: [ i <- 5 ] (if i == 4 then Q (ML^2T^(-4)I^(-1)) else Q (ML^2T^(-3)I^(-1)))
 \end{verbatim}
-This says that the entries in the first column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities.
-Meanwhile the entries of \texttt{B} are quantities of dimension \todo{fill in} respectively.
+This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities.
+Meanwhile the entries of \texttt{B} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first three rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) respectively.
 LabMate reports that the type of \texttt{C} should be
@@
 \end{verbatim}
-that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$.
+that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$.
 Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{C}, and to leave the type of one of the entries in \texttt{B} as an unknown:
@@
 
+\section{Acknowledgements}
+
+This work was undertaken jointly by the Mathematically Structured Programming Group of the University of Strathclyde and the National Physical Laboratory’s Data Science department as part of Data Science’s Tools for Trustworthiness National Measurement System (NMS) project 2023 – 2024.
+
+Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from \sref{sec:example}. Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper.
+
 \bibliographystyle{ws-procs9x6}
 \bibliography{labmate}
-
+%\printbibliography
 \end{document}

### 2024-04-08 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] change style file

@@
-\documentclass{ws-procs9x6}
+\documentclass{IMEKO2024}
 
 %%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage{amsmath}
 \usepackage[colorlinks=true]{hyperref}
@@
 \usepackage{mathtools}
-% Added by Keith L 05/04/2024
-% Removed by FNF 06/04/2024 (incompatible with style file)
-%\usepackage[]{biblatex}
-%\addbibresource{ws-procs9x6.bib}
-\usepackage{graphics}
-%\usepackage{caption}
+\usepackage{graphicx}
+\usepackage{rotating}
+\usepackage{subcaption}
+%\usepackage{multirow}
+\usepackage{xkeyval}
+\usepackage[font=small,justification=centering]{caption}
+\captionsetup{labelsep = period}
 \usepackage{floatrow}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \title{LabMate: supporting types for MATLAB}
-\author{Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Alistair Forbes, Keith Lines}
-\address{University of Strathclyde, National Physical Laboratory}
-%TODO: Affiliations
+\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
+\affiliation{%
+\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , Glasgow, UK}  \\
+\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, Teddington, UK}}
 
-\bodymatter
+\maketitle
+
+\begin{abstract}
+  TODO
+\end{abstract}
+\begin{keywords}
+  TODO
+\end{keywords}
 
@@
 
-The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~\ref{fig:rec_out} provides an example of recorded output, using simulated data.
+The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~1%\ref{fig:rec_out}
+provides an example of recorded output, using simulated data.
 \begin{figure}[H]
-    \centering
-    \captionsetup{justification=centering}
+  \begin{center}
     \includegraphics[width=0.6\linewidth]{example_output.jpg}
-    \caption{Simulated example of recorded output}       
-    \label{fig:rec_out}
+
+    \captionwide{Simulated example of recorded output.}
+  \end{center}
 \end{figure}
@@
 %
-We will see in \sref{sec:example-revisited} how this can be helpful.
+We will see in Section~\ref{sec:example-revisited} how this can be helpful.
 
@@
 
-Coming back to the example from \sref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
+Coming back to the example from Section~\ref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
 \begin{verbatim}
@@
 
-Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from \sref{sec:example}. Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper.
+Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from Section \ref{sec:example}.
+Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper.
 

### 2024-04-08 writing/IMEKOWC2024/LabMate/paper.tex
some minor edits to Section 2

@@
     \includegraphics[width=0.6\linewidth]{example_output.jpg}
-
     \captionwide{Simulated example of recorded output.}
+    \label{fig:rec_out}
   \end{center}
@@
 \]
-
-Where in the matrix:
+%
+The matrix on the left is made up of the following values:
+%
 \begin{itemize}
@@
 
-    \item The first column will contain either 1 or -1 for non-calibration voltage readings (i.e., those to which the calibration factor has not been applied) depending on the direction of the input signal.
-For calibration readings the value will be 0.
+    \item The first column will contain either $1$ or $-1$ for non-calibration voltage readings (i.e., those to which the calibration factor has not been applied) depending on the direction of the input signal.
+For calibration readings the value will be $0$.
 
-    \item The second column contains equivalent values for voltage readings in which the calibration factor has been applied.
-I.e.
-0 for non-calibration readings and 1 or -1 depending on the input signal direction for calibration readings.
+    \item The second column contains equivalent values for voltage readings in which the calibration factor has been applied, i.e., $0$ for non-calibration readings and $1$ or $-1$ depending on the input signal direction for calibration readings.
 
-    \item The third column contains 1 for non-calibration readings and 0 for calibration.
+    \item The third column contains $1$ for non-calibration readings and $0$ for calibration.
 
@@
 
-    \item The fifth columns contains time values.
+    \item The fifth column contains time values.
 \end{itemize}
@@
 %
-\todo{Is computing C from A and B what we want? \\ Keith 05/04/2024 ***No. We are computing B from A and C using a least squares line fit.*** }
+\todo{Is computing C from A and B what we want?
+\\ Keith 05/04/2024 ***No.
+We are computing B from A and C using a least squares line fit.*** \\ FNF 08/04/2024 Should the above be rewritten to do such a line fit, then?}
 The lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{input directives} by LabMate.
@@
 %> typeof C
-%< the type of C is quite a puzzle; unsolved constraints 5 =? 1, 1 =? 5
+%< the type of C is quite a puzzle;
+%<   unsolved constraints 5 =? 1, 1 =? 5
 \end{verbatim}

### 2024-04-08 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] added typing rules figure

@@
 \captionsetup{labelsep = period}
+\usepackage{subcaption}
 \usepackage{floatrow}
@@
 %
-Summarised as an inference rule, matrix types are thus formed as follows:
+Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
+%\[
+%  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
+%  {
+%    \istype{R}
+%    &\quad
+%    \istype{C}
+%    &\quad
+%    x : R, y : C \vdash \istype{E(x,y)}
+%    &\quad
+%    rs : \List{R}
+%    &\quad
+%    cs : \List{C}
+%  }
+%\]
 %
+\begin{figure*}[th]
+\begin{subfigure}{\textwidth}
+  \begin{center}
     \[
@@
     \]
+  \end{center}
+  \caption{Type formation rule for matrix types}
+  \label{fig:matrix_intro}
+\end{subfigure}
+\hfill
+\begin{subfigure}{\textwidth}
+  \begin{center}
+    \[
+      \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
+      {
+        x : R, y : C \vdash \isadd{E(x, y)}
+        &\quad
+        A : \Matrix{R}{C}{E}{rs}{cs}
+        &\quad
+        B : \Matrix{R}{C}{E}{rs}{cs}
+      }
+    \]
+  \end{center}
+  \caption{Addition rule for matrices}
+  \label{fig:matrix_add}
+\end{subfigure}
+\hfill
+\begin{subfigure}{\textwidth}
+  \begin{center}
+    \[
+      \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
+      {
+        \ismult{E}{E'}{E''}
+        &\quad
+        A : \Matrix{R}{C}{E}{rs}{ms}
+        &\quad
+        B : \Matrix{R}{C}{E'}{ms}{cs}
+      }
+    \]
+  \end{center}
+  \caption{Multiplication rule for matrices}
+  \label{fig:matrix_mul}
+\end{subfigure}
+% FIXME : \caption is broken
+\captionwide{Typing rules for matrices}
+\label{fig:matrix_rules}
+\end{figure*}
 
@@
 
-Two matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
+The conditions for adding two matrices are given in Figure~\ref{fig:matrix_add} --- the matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
 numeric (for example integers, or floating point numbers), and if $A$
-and $B$ have the same matrix type:
+and $B$ have the same matrix type.
 %
-\[
-  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
-  {
-    x : R, y : C \vdash \isadd{E(x, y)}
-    &\quad
-    A : \Matrix{R}{C}{E}{rs}{cs}
-    &\quad
-    B : \Matrix{R}{C}{E}{rs}{cs}
-  }
-\]
+%\[
+%  \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
+%  {
+%    x : R, y : C \vdash \isadd{E(x, y)}
+%    &\quad
+%    A : \Matrix{R}{C}{E}{rs}{cs}
+%    &\quad
+%    B : \Matrix{R}{C}{E}{rs}{cs}
+%  }
+%\]
 %
+
 Matrix multiplication is trickier, because it does not demand that $A$
 and $B$ have the same type --- rather, it demands that $A$ and $B$
-have \remph{compatible} types:
+have \remph{compatible} types, see Figure~\ref{fig:matrix_mul}.
 %
-\[
-  \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
-  {
-    \ismult{E}{E'}{E''}
-    &\quad
-    A : \Matrix{R}{C}{E}{rs}{ms}
-    &\quad
-    B : \Matrix{R}{C}{E'}{ms}{cs}
-  }
-\]
+%\[
+%  \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
+%  {
+%    \ismult{E}{E'}{E''}
+%    &\quad
+%    A : \Matrix{R}{C}{E}{rs}{ms}
+%    &\quad
+%    B : \Matrix{R}{C}{E'}{ms}{cs}
+%  }
+%\]
 The entry types $E$, $E'$, $E''$ are multipliable if for all $x$, $y$
@@
 \begin{verbatim}
-%> A :: [ i <- 12,  j <- 5 ] (if j == 4 then Q T else Q 1)
-%> B :: [ i <- 5 ] (if i == 4 then Q (ML^2T^(-4)I^(-1)) else Q (ML^2T^(-3)I^(-1)))
+% > A :: [ i <- 12,  j <- 5 ]
+      (if j == 4 then Q T else Q 1)
+% > B :: [ i <- 5 ]
+      (if i == 4
+         then Q (ML^2T^(-4)I^(-1))
+         else Q (ML^2T^(-3)I^(-1)))
 \end{verbatim}
@@
 \begin{verbatim}
-%> B :: [ Q TODO; Q TODO; ?aTy; Q TODO; Q TODO; Q TODO ]
+% > B :: [ Q TODO; Q TODO; ?aTy;
+           Q TODO; Q TODO ]
 \end{verbatim}

### 2024-04-09 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fixing Section 2

@@
 \affiliation{%
-\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , Glasgow, UK}  \\
-\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, Teddington, UK}}
+\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , UK}  \\
+\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}}
 
@@
 
-This calculation can be converted into a MATLAB problem with LabMate annotations as follows.
+This calculation can be converted into a MATLAB problem with LabMate annotations as follows, using MATLAB's builtin left division operator \texttt{A \textbackslash{} b} to compute a least squares fit.
+We then ask for the type of the resulting solution.
 Note that for reasons of space only twelve values are listed.
@@
 
-%> B :: [ 5 x 1 ] double
-B = [ a aCal c cCal m ]
+%> b :: [ 12 x 1 ] double
+b = [ v1 v2 v3 v4 v5 v6 ...
+        v7 v8 v9 v10 v11 v12 ]
 
-C = A * B
+x = A \ b
 
-%> typeof C
+%> typeof x
 \end{verbatim}
 %
-\todo{Is computing C from A and B what we want?
-\\ Keith 05/04/2024 ***No.
-We are computing B from A and C using a least squares line fit.*** \\ FNF 08/04/2024 Should the above be rewritten to do such a line fit, then?}
 The lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{input directives} by LabMate.
@@
 Did you notice the error in the above code?
-The matrix \texttt{B} should be of size $5 \times 1$, but is erroneously of size $1 \times 5$ instead.
-LabMate detects this error and produces the following as output, where we have suppressed the same definition of \texttt{A} as above:
+The vector \texttt{b} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead.
+LabMate detects this error and produces the following as output.
 \begin{verbatim}
 %> A :: [ 12 x 5 ] double
-A = ...
+A = [ 1  0  1  0  t1
+      1  0  1  0  t2
+      1  0  1  0  t3
+     -1  0  1  0  t4
+     -1  0  1  0  t5
+     -1  0  1  0  t6
+      0  1  0  1  t7
+      0  1  0  1  t8
+      0  1  0  1  t9
+      0 -1  0  1  t10
+      0 -1  0  1  t11
+      0 -1  0  1  t12 ]
 
-%> B :: [ 5 x 1 ] double
-B = [ a aCal c cCal m ]
+%> b :: [ 12 x 1 ] double
+%< unsolved constraints 12 =? 1, 1 =? 12
+b = [ v1 v2 v3 v4 v5 v6 ...
+        v7 v8 v9 v10 v11 v12 ]
 
-C = A * B
+x = A \ B
 
-%> typeof C
-%< the type of C is quite a puzzle;
-%<   unsolved constraints 5 =? 1, 1 =? 5
+%> typeof x
+%< the type of x is quite a puzzle
 \end{verbatim}
 %
-That is, LabMate says that for \texttt{C} to have a sensible type, we need the (impossible) constraints $5 = 1$ and $1 = 5$ to be true, arising from the fact that we have declared \texttt{B} to be a $5 \times 1$ matrix, but in fact it is a $1 \times 5$ matrix.
-Note that the original source code has been retained, but a response has been inserted to our \texttt{typeof} query.
-If we change \texttt{B} to
+That is, LabMate says that for \texttt{x} to have a sensible type, we need the (impossible) constraints $5 = 1$ and $1 = 5$ to be true, arising from the fact that we have declared \texttt{b} to be a $12 \times 1$ matrix, but in fact it is a $1 \times 12$ matrix.
+Note that the original source code has been retained exactly as is, including whitespace characters, but a response has been inserted to our \texttt{typeof} query.
+If we change \texttt{b} to
 \begin{verbatim}
-B = [ a; aCal; c; cCal; m ]
+b = [ v1 v2 v3 v4 v5 v6 ...
+        v7 v8 v9 v10 v11 v12 ]'
 \end{verbatim}
-instead, then LabMate responds
+using MATLAB's transpose operator instead, then LabMate responds
 \begin{verbatim}
-%< C :: [ 12 x 1 ] double
+%< x :: [ 5 x 1 ] double
 \end{verbatim}
-This way, LabMate can detect matrix size errors at development time, rather than at runtime.
+This way, LabMate can detect matrix size errors at development time, rather than at runtime --- running the original code would indeed crash with the error \textcolor{red}{\textbf{\texttt{Matrix dimensions must agree}}}.
 

### 2024-04-10 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fix Section 5

@@
 \begin{verbatim}
-% > A :: [ i <- 12,  j <- 5 ]
+%> A :: [ i <- 12,  j <- 5 ]
           (if j == 4 then Q T else Q 1)
-% > B :: [ i <- 5 ]
+
+%> b :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
+\end{verbatim}
+This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities.
+Meanwhile all the entries of \texttt{b} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$.
+LabMate reports that the type of \texttt{x} should be
+\begin{verbatim}
+%< x :: [ i <- 5 ]
         (if i == 4
@@
 \end{verbatim}
-This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities.
-Meanwhile the entries of \texttt{B} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first three rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) respectively.
-LabMate reports that the type of \texttt{C} should be
+that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first four rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) in the final row.
+%
+Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{x}, and to leave the type of some of the entries in \texttt{b} as an unknown
 \begin{verbatim}
-%< C :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
+%> b :: [ i <- 5 ]
+        (if i == 4
+           then ?bTy
+           else Q (ML^2T^(-3)I^(-1)))
 \end{verbatim}
-that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$.
-Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{C}, and to leave the type of one of the entries in \texttt{B} as an unknown:
-\begin{verbatim}
-% > B :: [ Q TODO; Q TODO; ?aTy;
-           Q TODO; Q TODO ]
-\end{verbatim}
-We can then ask LabMate to solve the \remph{meta variable} \texttt{?aTy}, to which LabMate responds that we must have \todo{\texttt{?aTy} = fill in}.
+We can then ask LabMate to solve the \remph{meta variable} \texttt{?bTy}, to which LabMate responds that we must also have \texttt{?aTy} = $\mathsf{Q}\,(\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1})$.
 This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.
@@
 Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from Section \ref{sec:example}.
-Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper.
+Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper, and to Professor Neil Ghani for his support.
 

### 2024-04-10 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] abstract and (commented out) keywords

@@
 \newcommand{\todo}[1]{\textcolor{red}{\textbf{TODO: #1}}}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{abstract}
-  TODO
+  Many computations in science and engineering are implemented in the
+  programming language MATLAB. However the high-level meaning of such
+  MATLAB programs stay informal, which can lead to implementation
+  errors and bugs, for example relating to incompatible units of
+  measure for quantities, or incompatible sizes of matrices at
+  runtime. We are in the process of developing LabMate, which is a
+  tool for reifying current informal programmer practices into a
+  language of formal comments. These comments are ignored by MATLAB,
+  but acted on and checked by LabMate. We outline the design
+  principles behind LabMate, our current progress, and our future
+  plans.
 \end{abstract}
-\begin{keywords}
-  TODO
-\end{keywords}
+%\begin{keywords}
+%  MATLAB, software correctness, dimensional consistency, type theory
+%\end{keywords}
 

### 2024-04-10 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] typo
[ writing ] section 2 wip

@@
 
-\title{LabMate: supporting types for MATLAB}
-\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
+\title{LabMate: a prospectus for types for MATLAB}
+\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Andr\'e Videla  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
 \affiliation{%
@@
 \end{abstract}
-%\begin{keywords}
-%  MATLAB, software correctness, dimensional consistency, type theory
-%\end{keywords}
+
+\begin{keywords}
+  MATLAB, software correctness, dimensional consistency, type theory
+\end{keywords}
 
@@
 %
-The matrix on the left is made up of the following values:
+The matrix left of $\times$ is made up of the following values:
 %
@@
 
-    \item The forth column is the opposite of the third.
+    \item The fourth column is the opposite of the third.
 
@@
 
-The first row vector contains values to be calculated using the line fit.
-The second column vector contains voltage readings.
-The measured value of the resistor is calculated using the contents of the first row vector.
-Details of that calculation are beyond the scope of this paper.
+The column vector right of $\times$ and left of $=$ contains values to
+be calculated using the line fit which then determine the measured
+value of the resistor by a calculation which is beyond the scope of
+this paper. The column vector right of $=$ contains voltage readings.
 
-This calculation can be converted into a MATLAB problem with LabMate annotations as follows, using MATLAB's builtin left division operator \texttt{A \textbackslash{} b} to compute a least squares fit.
-We then ask for the type of the resulting solution.
-Note that for reasons of space only twelve values are listed.
-A real measurement would contain considerably more values:
+This calculation can be converted into a MATLAB problem using the
+builtin left division operator \texttt{\textbackslash} to compute a
+least squares fit. Let us step through how we intend to use LabMate to
+help us deliver this solution.
+
+We then ask for the type of the resulting solution. Note
+that for reasons of space only twelve values are listed. A real
+measurement would contain considerably more values:
 \begin{verbatim}
-%> A :: [ 12 x 5 ] double
-A = [ 1  0  1  0  t1
-      1  0  1  0  t2
-      1  0  1  0  t3
-     -1  0  1  0  t4
-     -1  0  1  0  t5
-     -1  0  1  0  t6
-      0  1  0  1  t7
-      0  1  0  1  t8
-      0  1  0  1  t9
-      0 -1  0  1  t10
-      0 -1  0  1  t11
-      0 -1  0  1  t12 ]
+%> z3 :: [ 3 x 1 ] double
+z3 = [ 0; 0; 0 ]
+%> i3 :: [ 3 x 1 ] double
+i3 = [ 1; 1; 1 ]
+%> times :: [ 12 x 1 ] double
+% define times as a column vector
+  of time measurements
+
+ddnc = [ i3  z3  i3  z3
+        -i3  z3  i3  z3
+         z3  i3  z3  i3
+         z3 -i3  z3  i3 ]
+A = [ ddnc times ]
+%> typeof A
+%< A :: [ 12 x 5 ] double
 
@@
 LabMate detects this error and produces the following as output.
+
+%%> A :: [ 12 x 5 ] double
+%A = [ 1  0  1  0  t1
+%      1  0  1  0  t2
+%      1  0  1  0  t3
+%     -1  0  1  0  t4
+%     -1  0  1  0  t5
+%     -1  0  1  0  t6
+%      0  1  0  1  t7
+%      0  1  0  1  t8
+%      0  1  0  1  t9
+%      0 -1  0  1  t10
+%      0 -1  0  1  t11
+%      0 -1  0  1  t12 ]
+
 \begin{verbatim}
-%> A :: [ 12 x 5 ] double
-A = [ 1  0  1  0  t1
-      1  0  1  0  t2
-      1  0  1  0  t3
-     -1  0  1  0  t4
-     -1  0  1  0  t5
-     -1  0  1  0  t6
-      0  1  0  1  t7
-      0  1  0  1  t8
-      0  1  0  1  t9
-      0 -1  0  1  t10
-      0 -1  0  1  t11
-      0 -1  0  1  t12 ]
+
+%> z3 :: [ 3 x 1 ] double
+z3 = [0; 0; 0]
+%> i3 :: [ 3 x 1 ] double
+i3 = [1; 1; 1]
+%> times :: [ 12 x 1 ] double
+ddnc = [ i3  z3  i3  z3
+        -i3  z3  i3  z3
+         z3  i3  z3  i3
+         z3 -i3  z3  i3 ]
+A = [ ddnc times ]
+%> typeof A
+%< A :: [ 12 x 5 ] double
 
@@
 
-x = A \ B
+x = A \ b
 
@@
 %
-Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
+Matrix types formation can thus be summarised as the inference rule
+given in Figure~\ref{fig:matrix_intro}.
 %\[

### 2024-04-10 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] section 2 with more incident

@@
 \newcommand{\lsb}{\texttt{\symbol{91}}}
+\newcommand{\bsl}{\texttt{\symbol{92}}}
 \newcommand{\rsb}{\texttt{\symbol{93}}}
@@
 This calculation can be converted into a MATLAB problem using the
-builtin left division operator \texttt{\textbackslash} to compute a
+builtin left division operator {\bsl} to compute a
 least squares fit. Let us step through how we intend to use LabMate to
@@
 
-We then ask for the type of the resulting solution. Note
-that for reasons of space only twelve values are listed. A real
-measurement would contain considerably more values:
+We begin by declaring the types of the input data: the \texttt{times} and the \texttt{voltages}.
+Our example calculation uses twelve time and voltage inputs, arranged as two rows to be read from a file.
+A real measurement would involve considerably larger datasets.
+\begin{verbatim}
+%> times    :: [ 1 x 12 ] double
+%> voltages :: [ 1 x 12 ] double
+\end{verbatim}
+Lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{directives} by LabMate, here giving us enough information not only to insist that the values of \texttt{times} and \texttt{voltages} \remph{conform} to the given sizes, but to determine \remph{how} to read them from a file.
+We would issue the further directive
+\begin{verbatim}
+%> readfrom "inputs.txt" times voltages
+\end{verbatim}
+LabMate is a program \remph{transducer}: its output is a modified version of its input, allowing it to respond to directives with information in comments or by generating code.
+Here we would expect to see something like
+\begin{verbatim}
+%> readfrom "inputs.txt" times voltages
+%<{
+h=fopen("inputs.txt");
+c=textscan(h,'%f');
+fclose(h);
+src = c{1};
+readPtr = 1;
+for i = 1:12
+  times[i] = src[readPtr];
+  readPtr = readPtr + 1;
+end
+for i = 1:12
+  voltages[i] = src[readPtr];
+  readPtr = readPtr + 1;
+end
+%<}
+\end{verbatim}
+where the comment lines `\pct\res\lcb' and `\pct\res\rcb' visually delimit the extent of the generated boilerplate code, so that humans may safely ignore it.
+LabMate knows enough about scope in MATLAB to ensure that auxiliary variable naming never creates conflict.
+We have kept to one datum per line for simplicity, but we are free to extend our format specification language as we see fit~\cite{mgen}.
+
+Next, let us build up the matrix in the calculation. The method gives the first four columns a pattern of blocks, pasted in a grid.
 \begin{verbatim}
@@
 i3 = [ 1; 1; 1 ]
-%> times :: [ 12 x 1 ] double
-% define times as a column vector
-  of time measurements
 
@@
          z3 -i3  z3  i3 ]
-A = [ ddnc times ]
-%> typeof A
-%< A :: [ 12 x 5 ] double
+\end{verbatim}
+LabMate will check that this pasting strategy delivers a rectangular matrix, well in advance of MATLAB runtime.
+Pasting on the measured times (transposed to a column) completes the matrix.
+Let us ask for its type.
+LabMate responds to information-seeking directives by writing comments of its own, beginning with `\pct\res'.
+\begin{verbatim}
+M = [ ddnc times' ]
+%> typeof M
+%< M :: [ 12 x 5 ] double
+\end{verbatim}
 
-%> b :: [ 12 x 1 ] double
-b = [ v1 v2 v3 v4 v5 v6 ...
-        v7 v8 v9 v10 v11 v12 ]
-
-x = A \ b
-
+Now that we have assembled the data, we may solve the equations and ask for the type of the solution.
+\begin{verbatim}
+x = M \ voltages
 %> typeof x
 \end{verbatim}
-%
-The lines starting with `\texttt{\%>}' are interpreted as ordinary comments by MATLAB, but as \remph{input directives} by LabMate.
-LabMate acts by reading the source code and outputting it again, but possibly with some additional \remph{responses} to the directives added.
-
-Did you notice the error in the above code?
-The vector \texttt{b} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead.
-LabMate detects this error and produces the following as output.
-
-%%> A :: [ 12 x 5 ] double
-%A = [ 1  0  1  0  t1
-%      1  0  1  0  t2
-%      1  0  1  0  t3
-%     -1  0  1  0  t4
-%     -1  0  1  0  t5
-%     -1  0  1  0  t6
-%      0  1  0  1  t7
-%      0  1  0  1  t8
-%      0  1  0  1  t9
-%      0 -1  0  1  t10
-%      0 -1  0  1  t11
-%      0 -1  0  1  t12 ]
+Did you notice the error in the above code? The vector \texttt{voltages} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate detects this error and responds:
 
 \begin{verbatim}
-
-%> z3 :: [ 3 x 1 ] double
-z3 = [0; 0; 0]
-%> i3 :: [ 3 x 1 ] double
-i3 = [1; 1; 1]
-%> times :: [ 12 x 1 ] double
-ddnc = [ i3  z3  i3  z3
-        -i3  z3  i3  z3
-         z3  i3  z3  i3
-         z3 -i3  z3  i3 ]
-A = [ ddnc times ]
-%> typeof A
-%< A :: [ 12 x 5 ] double
-
-%> b :: [ 12 x 1 ] double
-%< unsolved constraints 12 =? 1, 1 =? 12
-b = [ v1 v2 v3 v4 v5 v6 ...
-        v7 v8 v9 v10 v11 v12 ]
-
-x = A \ b
-
+x = M \ voltages
+%< unsolved constraint 12 =? 1
 %> typeof x
@@
 %
-That is, LabMate says that for \texttt{x} to have a sensible type, we need the (impossible) constraints $5 = 1$ and $1 = 5$ to be true, arising from the fact that we have declared \texttt{b} to be a $12 \times 1$ matrix, but in fact it is a $1 \times 12$ matrix.
-Note that the original source code has been retained exactly as is, including whitespace characters, but a response has been inserted to our \texttt{typeof} query.
-If we change \texttt{b} to
+That is, LabMate says that for \texttt{x} to have a sensible type, we need \texttt{M} and \texttt{voltages} to have the same number of rows.
+We forgot to transpose \texttt{voltages}!
+If we make the necessary repair, we should now see something more satisfying.
 \begin{verbatim}
-b = [ v1 v2 v3 v4 v5 v6 ...
-        v7 v8 v9 v10 v11 v12 ]'
-\end{verbatim}
-using MATLAB's transpose operator instead, then LabMate responds
-\begin{verbatim}
+x = M \ voltages'
+%> typeof x
 %< x :: [ 5 x 1 ] double
 \end{verbatim}
-This way, LabMate can detect matrix size errors at development time, rather than at runtime --- running the original code would indeed crash with the error \textcolor{red}{\textbf{\texttt{Matrix dimensions must agree}}}.
+This way, LabMate can detect matrix size errors at development time, rather than at runtime --- running the original code would indeed crash with the error:
+\[\textcolor{red}{\textbf{\texttt{Matrix dimensions must agree}}}\]
 

### 2024-04-10 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] \mathit FTW

@@
 
+\newcommand{\rs}{\mathit{rs}}
+\newcommand{\ms}{\mathit{ms}}
+\newcommand{\cs}{\mathit{cs}}
+
 Under the hood, LabMate translates MATLAB expressions to typed
@@
 which might vary depending on the values of parameters $x : R$ and
-$y : C$, as well as two lists $rs : \List{R}$ and $\List{C}$ --- the
-idea being that the $i$th entry in the list $rs$ and the $j$th entry
-in the list $cs$ determines the type of the $(i, j)$ coordinate in the
+$y : C$, as well as two lists $\rs : \List{R}$ and $\List{C}$ --- the
+idea being that the $i$th entry in the list $\rs$ and the $j$th entry
+in the list $\cs$ determines the type of the $(i, j)$ coordinate in the
 matrix.
@@
 %\[
-%  \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
+%  \infer{\istype{\Matrix{R}{C}{E}{rs}{\cs}}}
 %  {
@@
     \[
-      \infer{\istype{\Matrix{R}{C}{E}{rs}{cs}}}
+      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
       {
@@
         &\quad
-        rs : \List{R}
+        \rs : \List{R}
         &\quad
-        cs : \List{C}
+        \cs : \List{C}
       }
@@
     \[
-      \infer{A + B : \Matrix{R}{C}{E}{rs}{cs}}
+      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
       {
@@
         &\quad
-        A : \Matrix{R}{C}{E}{rs}{cs}
+        A : \Matrix{R}{C}{E}{\rs}{\cs}
         &\quad
-        B : \Matrix{R}{C}{E}{rs}{cs}
+        B : \Matrix{R}{C}{E}{\rs}{\cs}
       }
@@
     \[
-      \infer{A * B : \Matrix{R}{C}{E''}{rs}{cs}}
+      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
       {
@@
         &\quad
-        A : \Matrix{R}{C}{E}{rs}{ms}
+        A : \Matrix{R}{C}{E}{\rs}{\ms}
         &\quad
-        B : \Matrix{R}{C}{E'}{ms}{cs}
+        B : \Matrix{R}{C}{E'}{\ms}{\cs}
       }
@@
 and $y$ must be the unique element of the unit type), and the only
-information available in the lists $rs : \List{\One}$ and
-$cs : \List{\One}$ is their length.
+information available in the lists $\rs : \List{\One}$ and
+$\cs : \List{\One}$ is their length.
 %
@@
 Another noteworthy special case which we want to support takes $R$ and
-$R$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
+$C$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
 type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
@@
 \[
-  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{rs}{(cs_A \append cs_B)}}
+  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{\rs}{(cs_A \append cs_B)}}
   {
-    A : \Matrix{R}{C}{E}{rs}{cs_A}
+    A : \Matrix{R}{C}{E}{\rs}{cs_A}
     &\quad
-    B : \Matrix{R}{C}{E}{rs}{cs_B}
+    B : \Matrix{R}{C}{E}{\rs}{cs_B}
   }
@@
 \[
-  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(rs_A \append rs_B)}{cs}}
+  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(\rs_A \append \rs_B)}{cs}}
   {
-    A : \Matrix{R}{C}{E}{rs_A}{cs}
+    A : \Matrix{R}{C}{E}{\rs_A}{cs}
     &\quad
-    B : \Matrix{R}{C}{E}{rs_B}{cs}
+    B : \Matrix{R}{C}{E}{\rs_B}{cs}
   }

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] picture of a matrix

@@
 \newcommand{\todo}[1]{\textcolor{red}{\textbf{TODO: #1}}}
+
+\newcommand{\param}{parametrize}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Under the hood, LabMate translates MATLAB expressions to typed
-expressions in its own core type theory.
+Under the hood, LabMate translates selected well behaved MATLAB expressions
+to typed expressions in its own core theory of types and values.
 %
-Since matrices feature heavily in MATLAB code, the type of matrices
-plays a central role in this type theory.
+As matrices feature heavily in MATLAB code, the types of matrices
+play a central role in this theory.
 %
+In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
+%
+A matrix type looks like
+\[\Matrix{R}{C}{E}{\rs}{\cs}
+\]
+%
+We thus \param{} matrix types by two arbitrary types $R$, for the individuals the rows concern, and $C$ for the columns.
+We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals.
+A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry with coordinates $(i,j)$.
+\[\begin{array}{r|ccccc|}
+\;     & c_1      & \cdots & c_j     & \cdots & \multicolumn{1}{c}{c_n}     \\
+\hline
+r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
+\vdots &          &        &         &        &         \\
+r_i    & e_{i,1}  &        & e_{i,j} &        & e_{i,n} \\
+\vdots &          &        &         &        &         \\
+r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
+      \cline{2-6}
+\end{array}\]
+
 We make this type rather precise, by allowing the type of each entry
@@
 %
-we parameterise the type of matrices by a type $R$ associated with
+we \param{} the type of matrices by a type $R$ associated with
 each row, a type $C$ associated with each column, and a type $E(x,y)$

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] more matrix picture

@@
 We thus \param{} matrix types by two arbitrary types $R$, for the individuals the rows concern, and $C$ for the columns.
-We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals.
-A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry with coordinates $(i,j)$.
+We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals.
+A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$.
 \[\begin{array}{r|ccccc|}
-\;     & c_1      & \cdots & c_j     & \cdots & \multicolumn{1}{c}{c_n}     \\
-\hline
+\multicolumn{1}{c}{\;} & c_1      & \cdots & c_j     & \cdots & \multicolumn{1}{c}{c_n}     \\
+      \cline{2-6}
 r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
@@
 \end{array}\]
-
-We make this type rather precise, by allowing the type of each entry
-in the matrix to vary according to the row and column of the entry:
-%
-we \param{} the type of matrices by a type $R$ associated with
-each row, a type $C$ associated with each column, and a type $E(x,y)$
-which might vary depending on the values of parameters $x : R$ and
-$y : C$, as well as two lists $\rs : \List{R}$ and $\List{C}$ --- the
-idea being that the $i$th entry in the list $\rs$ and the $j$th entry
-in the list $\cs$ determines the type of the $(i, j)$ coordinate in the
-matrix.
 %

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] typing rules made prose...

@@
 A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$.
+In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code, while the $r$s and $c$s outside the rectangle are `metadata' coming from the matrix type.
 \[\begin{array}{r|ccccc|}
@@
 \end{array}\]
+To check that $\Matrix{R}{C}{E}{\rs}{\cs}$ is a valid type, LabMate must:
+\begin{enumerate}
+\item check that $R$ and $C$ are both valid types
+\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$
+\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$
+\end{enumerate}
+
+A notable special case is when both $R$ and $C$ are the unit type $R = C = \One$, that is, the type with exactly one element.
 %
+$\List{\One}$ amounts to a mere `tally', counting indistinct things, which we allow to be written as a number in decimal notation.
+%
+In this case, there is only one type of entries $E(r,c)$ (because both $r$
+and $c$ must be the unique element of the unit type), and the only
+information available in the lists $\rs : \List{\One}$ and
+$\cs : \List{\One}$ are their lengths.
+%
+Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
+$m \times n$ matrices with entries of type $E$, where $m$ and $n$ are
+the unique lists over the unit type of length $m$ and $n$
+respectively.
+
 Matrix types formation can thus be summarised as the inference rule
@@
 \end{figure*}
-
-A notable special case is when both $R$ and $C$ are the unit type
-$R = C = \One$, that is, the type with exactly one element.
-%
-In this case, there is only one type of entries $E$ (because both $x$
-and $y$ must be the unique element of the unit type), and the only
-information available in the lists $\rs : \List{\One}$ and
-$\cs : \List{\One}$ is their length.
-%
-Thus, we have recovered $\Matrix{\One}{\One}{E}{m}{n}$ as the type of
-$m \times n$ matrices with entries of type $E$, where $n$ and $m$ are
-the unique lists over the unit type of length $n$ and $m$
-respectively.
 

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] hjux picture
[ writing ] more pics

@@
 %
-A matrix type looks like
+Internally, a matrix type looks like
 \[\Matrix{R}{C}{E}{\rs}{\cs}
@@
 %
-We thus \param{} matrix types by two arbitrary types $R$, for the individuals the rows concern, and $C$ for the columns.
+We thus \param{} matrix types by two arbitrary `header' types $R$, for the individuals the rows concern, and $C$ for the columns.
 We may then give two lists, $\rs = [r_1,\ldots,r_m]\in\List{R}$ (so that each $r_i\in R$) and $\cs = [c_1,\ldots,c_n]\in\List{C}$ of those individuals.
 A \param{}d type $E(r,c)$ gives the type of entry for any combination of row and column individual, with $E(r_i,c_j)$ being the type of the entry $e_{i,j}$.
-In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code, while the $r$s and $c$s outside the rectangle are `metadata' coming from the matrix type.
+In the diagram below, the $e$s inside the rectangle are the matrix entries computed by the MATLAB code.
+The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
 \[\begin{array}{r|ccccc|}
-\multicolumn{1}{c}{\;} & c_1      & \cdots & c_j     & \cdots & \multicolumn{1}{c}{c_n}     \\
-      \cline{2-6}
+\; & c_1      & \cdots & c_j     & \cdots & c_n     \\
+\hline
 r_1    & e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
@@
 r_m    & e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
-      \cline{2-6}
+\hline
 \end{array}\]
@@
 respectively.
+%
+That is what we render in LabMate directives and responses as
+\[
+\lsb \;m \;\texttt{x}\; n\; \rsb\; E
+\]
+as in our \texttt{M :: [ 1 x 5 ] double} in Section~\ref{sec:example}.
 
-Matrix types formation can thus be summarised as the inference rule
-given in Figure~\ref{fig:matrix_intro}.
+So much for what it is to be a valid matrix type: what does it mean to be a valid matrix \remph{in} such a type?
+%
+Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
+\[\begin{array}[b]{r|c|l}
+\; & c \\
+\cline{1-2}
+r  & e\in E(r,c) & \quad \in \Matrix{R}{C}{E}{[r]}{[c]} \\
+\cline{1-2}
+\end{array} 
+\]
+
+\[\begin{array}{l}
+\begin{array}{r|c|c|}
+  & \cs_A & \cs_b \\
+\hline
+&&\\
+&A \in & B \in \\
+\rs &\Matrix{R}{C}{E}{\rs}{\cs_A} & \Matrix{R}{C}{E}{\rs}{\cs_B} \\
+&&\\
+\hline
+\end{array}\\
+\;\;\; \in \Matrix{R}{C}{E}{\rs}{(\cs_A\append\cs_B)}
+\end{array}\]
+
+%Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
 %\[
@@
 \begin{figure*}[th]
-\begin{subfigure}{\textwidth}
-  \begin{center}
-    \[
-      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
-      {
-        \istype{R}
-        &\quad
-        \istype{C}
-        &\quad
-        x : R, y : C \vdash \istype{E(x,y)}
-        &\quad
-        \rs : \List{R}
-        &\quad
-        \cs : \List{C}
-      }
-    \]
-  \end{center}
-  \caption{Type formation rule for matrix types}
-  \label{fig:matrix_intro}
-\end{subfigure}
-\hfill
+%\begin{subfigure}{\textwidth}
+%  \begin{center}
+%    \[
+%      \infer{\istype{\Matrix{R}{C}{E}{\rs}{\cs}}}
+%      {
+%        \istype{R}
+%        &\quad
+%        \istype{C}
+%        &\quad
+%        x : R, y : C \vdash \istype{E(x,y)}
+%        &\quad
+%        \rs : \List{R}
+%        &\quad
+%        \cs : \List{C}
+%      }
+%    \]
+%  \end{center}
+%  \caption{Type formation rule for matrix types}
+%  \label{fig:matrix_intro}
+%\end{subfigure}
+%\hfill
 \begin{subfigure}{\textwidth}

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] vjux

@@
 So much for what it is to be a valid matrix type: what does it mean to be a valid matrix \remph{in} such a type?
-%
+
 Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
@@
 
+Secondly, to check a horizontal pasting of matrices, we must ensure that the header and entry types match, that the row headers are the same for left and right, and that the column headers can be split into a prefix for the left matrix and a suffix for the right.
+In other words, the column headers for the whole are the concatenation ($\append$) of the column headers for the parts.
 \[\begin{array}{l}
@@
 \hline
-\end{array}\\
+\end{array}\bigskip\\
 \;\;\; \in \Matrix{R}{C}{E}{\rs}{(\cs_A\append\cs_B)}
 \end{array}\]
+
+Thirdly, vertical pasting is similar, except that it is the row headers which split into prefix and suffix while the column headers are shared.
+\[
+\begin{array}[c]{r|c|}
+  & \cs \\
+  \hline
+  \;    & \\
+  \;    & A \in \\
+  \rs_A & \Matrix{R}{C}{E}{\rs_A}{\cs} \\
+  \;    & \\
+  \hline
+  \;    & \\
+  \;    & B \in \\
+  \rs_B & \Matrix{R}{C}{E}{\rs_B}{\cs} \\
+  \;    & \\
+  \hline
+\end{array}
+\in
+\begin{array}{l}
+\Matrix{R}{C}{E}{\\ \quad (\rs_A\append\rs_B)}{\cs}
+\end{array}
+\]
 

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] adding

@@
 
-\newcommand{\append}{+\!\!\!+}
+\newcommand{\append}{\mathop{+\!\!+}}
 \newcommand{\hjux}[2]{{#1}|{#2}}
@@
 \begin{enumerate}
-\item check that $R$ and $C$ are both valid types
-\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$
-\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$
+\item check that $R$ and $C$ are both valid types;
+\item check that $E(r,c)$ is a valid type whenever $r\in R$ and $c\in C$;
+\item check that $\rs\in\List{R}$ and $\cs\in\List{C}$.
 \end{enumerate}
@@
 Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
-\[\begin{array}[b]{r|c|l}
+\[\begin{array}[b]{r|@{}c@{}|l}
 \; & c \\
@@
 \]
+
+Now that we have matrices, we must explain how to typecheck matrix operations.
+
+To check if we may add matrices $A$ and $B$, LabMate must:
+\begin{enumerate}
+\item check that $A$ and $B$ both have the \remph{same} matrix type $\Matrix{R}{C}{E}{\rs}{\cs}$;
+\item check that \remph{every} entry type $E(r,c)$ admits addition.
+\end{enumerate}
+In the short term, LabMate can get by with a hard-coded table of which types admit addition, but more flexible and general approaches to operator overloading (e.g., Haskell's `type classes'~\cite{wadler.blott}) are readily adoptable in due course.
 
@@
 
-Terms of matrix type in the core type theory are built up from
-$1 \times 1$ matrices and pasting matrices horizontally and vertically.
-%
-Singleton matrices have the following typing rule
-\[
-  \infer{[e] : \Matrix{R}{C}{E}{[r]}{[c]}}
-    {r : R
-      &\quad
-      c : C
-      &\quad
-      e : E(r, c)
-    }
-\]
-%
-while horizontal and vertical pasting horizontally have the following
-typing rules:
-\[
-  \infer{\hjux{A}{B} : \Matrix{R}{C}{E}{\rs}{(cs_A \append cs_B)}}
-  {
-    A : \Matrix{R}{C}{E}{\rs}{cs_A}
-    &\quad
-    B : \Matrix{R}{C}{E}{\rs}{cs_B}
-  }
-\]
-\[
-  \infer{\vjux{A}{B} : \Matrix{R}{C}{E}{(\rs_A \append \rs_B)}{cs}}
-  {
-    A : \Matrix{R}{C}{E}{\rs_A}{cs}
-    &\quad
-    B : \Matrix{R}{C}{E}{\rs_B}{cs}
-  }
-\]
 When comparing matrices for equality, we reassociate horizontal and
@@
 
-The conditions for adding two matrices are given in Figure~\ref{fig:matrix_add} --- the matrices $A$ and $B$ can be added if all entry types $E(x, y)$ are
-numeric (for example integers, or floating point numbers), and if $A$
-and $B$ have the same matrix type.
 %

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] multiplication - it's the name of the game

@@
 The $r$s and $c$s outside the rectangle are \remph{meta}data coming from the matrix type, known only to LabMate, and not available as data to MATLAB.
-\[\begin{array}{r|ccccc|}
+\[\begin{array}{@{}r|ccccc|}
 \; & c_1      & \cdots & c_j     & \cdots & c_n     \\
@@
 Firstly, a singleton entry matrix must have singleton header lists $[r]$ and $[c]$.
-\[\begin{array}[b]{r|@{}c@{}|l}
+\[\begin{array}[b]{@{}r|@{}c@{}|l}
 \; & c \\
@@
 \[\begin{array}{l}
-\begin{array}{r|c|c|}
+\begin{array}{@{}r|c|c|}
   & \cs_A & \cs_b \\
@@
 \[
-\begin{array}[c]{r|c|}
+\begin{array}[c]{@{}r|c|}
   & \cs \\
@@
 \end{enumerate}
-In the short term, LabMate can get by with a hard-coded table of which types admit addition, but more flexible and general approaches to operator overloading (e.g., Haskell's `type classes'~\cite{wadler.blott}) are readily adoptable in due course.
+In the short term, LabMate can get by with a hard-coded table of which types (e.g.
+\texttt{double}) admit addition, but more flexible and general approaches to operator overloading (e.g., Haskell's `type classes'~\cite{wadler.blott}) are readily adoptable in due course.
+
+Multplication is rather more involved, of course. The usual requirement that the matrix sizes `meet in the middle' generalises straightforwardly to lists of headers,
+\[\begin{array}{@{}c@{}}
+\begin{array}[c]{@{}r|c|}
+  \;  & \ms \\
+  \hline
+  \;  &     \\
+  \;  &  A \in \\
+  \rs & \Matrix{R}{M}{E_1}{\rs}{\ms} \\
+  \;  & \\
+  \hline
+\end{array}
+  \times
+\begin{array}[c]{@{}r|c|}
+  \;  & \cs \\
+  \hline
+  \;  &     \\
+  \;  &     \\
+  \;  &  B \in \\
+  \ms & \Matrix{M}{C}{E_2}{\\ & \ms}{\cs} \\
+  \;  &     \\
+  \;  & \\
+  \hline
+\end{array}
+\bigskip\\
+\in \qquad
+\begin{array}[c]{@{}r|c|}
+  \;  & \cs \\
+  \hline
+  \;  &     \\
+  \rs & \Matrix{R}{C}{E_3}{\\ & \rs}{\cs} \\
+  \;  & \\
+  \hline
+\end{array}
+\end{array}
+\]
+but we must also check that the entry types are compatible. Specifically, LabMate must:
+\begin{enumerate}
+  \item check, for all $r$, $m$ and $c$, that if $a\in E_1(r,m)$ and $b\in E_2(m,c)$, then
+    $a\times b \in E_3(r,c)$;
+  \item check that \remph{every} entry type $E_3(r,c)$ admits addition.
+\end{enumerate}
+
+These conditions are straightforward in the special case where $R = C = \One$ and the $E$s are standard numerical types. There are, however, more motivating possibilities.
 
@@
 %\hfill
-\begin{subfigure}{\textwidth}
-  \begin{center}
-    \[
-      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
-      {
-        x : R, y : C \vdash \isadd{E(x, y)}
-        &\quad
-        A : \Matrix{R}{C}{E}{\rs}{\cs}
-        &\quad
-        B : \Matrix{R}{C}{E}{\rs}{\cs}
-      }
-    \]
-  \end{center}
-  \caption{Addition rule for matrices}
-  \label{fig:matrix_add}
-\end{subfigure}
-\hfill
+%\begin{subfigure}{\textwidth}
+%  \begin{center}
+%    \[
+%      \infer{A + B : \Matrix{R}{C}{E}{\rs}{\cs}}
+%      {
+%        x : R, y : C \vdash \isadd{E(x, y)}
+%        &\quad
+%        A : \Matrix{R}{C}{E}{\rs}{\cs}
+%        &\quad
+%        B : \Matrix{R}{C}{E}{\rs}{\cs}
+%      }
+%    \]
+%  \end{center}
+%  \caption{Addition rule for matrices}
+%  \label{fig:matrix_add}
+%\end{subfigure}
+%\hfill
 \begin{subfigure}{\textwidth}

### 2024-04-11 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] algebra fight

@@
 
+Of particular interest is the case when the header types, $R$ and
+$C$, are types of physical dimensions, with $E(r, c)$ being the
+type of physical quantities of dimension $r^{-1}\cdot c$.
+%
+Such quantities admit addition only if the dimensions match. Meanwhile, their multiplication correspondingly multiplies the dimensions.
+%
+This way, we can reduce dimensional consistency checking \`a la
+Kennedy~\cite{kennedyUOM} to typechecking, but also for programs involving matrices of non-uniform dimension, following the work of Hart~\cite{hart}.
+%
+We will see in Section~\ref{sec:example-revisited} how this can be helpful.
+
 %Matrix types formation can thus be summarised as the inference rule given in Figure~\ref{fig:matrix_intro}.
@@
 %
-\begin{figure*}[th]
+%\begin{figure*}[th]
 %\begin{subfigure}{\textwidth}
@@
 %\hfill
-\begin{subfigure}{\textwidth}
-  \begin{center}
-    \[
-      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
-      {
-        \ismult{E}{E'}{E''}
-        &\quad
-        A : \Matrix{R}{C}{E}{\rs}{\ms}
-        &\quad
-        B : \Matrix{R}{C}{E'}{\ms}{\cs}
-      }
+%\begin{subfigure}{\textwidth}
+%  \begin{center}
+%    \[
+%      \infer{A * B : \Matrix{R}{C}{E''}{\rs}{\cs}}
+%      {
+%        \ismult{E}{E'}{E''}
+%        &\quad
+%        A : \Matrix{R}{C}{E}{\rs}{\ms}
+%        &\quad
+%        B : \Matrix{R}{C}{E'}{\ms}{\cs}
+%      }
+%    \]
+%  \end{center}
+%  \caption{Multiplication rule for matrices}
+%  \label{fig:matrix_mul}
+%\end{subfigure}
+%% FIXME : \caption is broken
+%\captionwide{Typing rules for matrices}
+%\label{fig:matrix_rules}
+%\end{figure*}
+
+Checking the compatibility between operations and data boils down to checking equality of types.
+LabMate types are \remph{dependent} types, in the style of the state-of-the-art functional programming languages Agda and Idris~\cite{ydtm,agdathing,idristhing}.
+Dependent types express the reality that data validity is often a \remph{relative} notion, with the compatibility requirements for matrix multiplication being a standard example.
+Types which mention (hence depend on) some values (e.g., our header lists) are used to classify other values (e.g., our matrices) with respect to them.
+
+Equality of types therefore demands some notion of equality for the values they mention.
+We equip every type in our core theory with an algorithmic decision procedure testing when two expressions in the type are sufficiently obviously equal to guarantee compatibility of types.
+Languages like Agda and Idris generate their notion of expression equality by extending the evaluation rules used to compute with values at runtime to cope with expressions involving variables by getting stuck whenever a variable prevents a condition from being tested.
+In MATLAB, the matrix pastings \texttt{[[A B] C]} and \texttt{[A [B C]]} both mean the same thing as \texttt{[A B C]}, but for us to see that requires \remph{algebraic} properties of the $\append$ operator for our header lists.
+The state-of-the-art languages turn up to this algebra fight armed only with arithmetic!
+They force their programmers to give explicit proofs of type compatibility, doing the algebra by hand.
+We would not dream of inflicting this bureaucracy on LabMate users, especially when the small amount of algebra we need is perfectly straightforward~\cite{nueqns}.
+
+We incorporate the specific algebraic theories we need to manage matrices and dimensions into LabMate's expression equality.
+Indeed, we also consider when \remph{matrix} expressions are equal.
+We reassociate horizontal and vertical pastings, so that
+\[\begin{array}[c]{|c|@{}c@{}|c|}
+\cline{1-1} \cline{3-3}
+A &\;& B \\
+\cline{1-1} \cline{3-3}
+C &\;& D \\
+\cline{1-1} \cline{3-3}
+\end{array}
+\quad = \quad
+\begin{array}[c]{|c|c|}
+  \hline
+  A & B \\
+  \hline
+  \hline
+  C & D \\
+  \hline
+\end{array}
 \]
-  \end{center}
-  \caption{Multiplication rule for matrices}
-  \label{fig:matrix_mul}
-\end{subfigure}
-% FIXME : \caption is broken
-\captionwide{Typing rules for matrices}
-\label{fig:matrix_rules}
-\end{figure*}
-
-Another noteworthy special case which we want to support takes $R$ and
-$C$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
-type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
-%
-This way, we can reduce dimensional consistency checking \`a la
-Kennedy~\cite{kennedyUOM} also for programs involving matrices,
-following the work of Hart~\cite{hart}.
-%
-We will see in Section~\ref{sec:example-revisited} how this can be helpful.
-
-When comparing matrices for equality, we reassociate horizontal and
-vertical pastings, so that $\hjux{\vjux{A}{C}}{\vjux{B}{D}}$ and
-$\vjux{\hjux{A}{B}}{\hjux{C}{D}}$ indeed are considered equal.
 
@@
 
-Matrix multiplication is trickier, because it does not demand that $A$
-and $B$ have the same type --- rather, it demands that $A$ and $B$
-have \remph{compatible} types, see Figure~\ref{fig:matrix_mul}.
+%Matrix multiplication is trickier, because it does not demand that $A$
+%and $B$ have the same type --- rather, it demands that $A$ and $B$
+%have \remph{compatible} types, see Figure~\ref{fig:matrix_mul}.
 %
@@
 %\]
-The entry types $E$, $E'$, $E''$ are multipliable if for all $x$, $y$
-and $z$, we have that $E(x, y)$ and $E'(y, z)$ supports a
-multiplication operation with codomain $E''(x, z)$.
+%The entry types $E$, $E'$, $E''$ are multipliable if for all $x$, $y$
+%and $z$, we have that $E(x, y)$ and $E'(y, z)$ supports a
+%multiplication operation with codomain $E''(x, z)$.
 %
-This is certainly the case if all entry types involved are simply
-integers or floating point numbers, but this constraint can also
-ensure that the physical units involved line up.
+%This is certainly the case if all entry types involved are simply
+%integers or floating point numbers, but this constraint can also
+%ensure that the physical units involved line up.
 

### 2024-04-12 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] bash the bibstyle
[ writing ] honesty at first

@@
 engineering toolchains and practices.
+
+LabMate is under constant development; the latest version can be found at \url{https://github.com/msp-strath/LabMate/tree/devel}.
+%
+At time of writing, the necessary core type system is in place: we are busily extending the variety of MATLAB expressions it can make sense of, and the constraints it can solve.
+%
+We are close to covering the examples in this paper, but not quite there yet.
+%
+We aim to eliminate the speculative future tense from our examples before camera-ready copy is required.
+%
+This paper effectively documents how far we have come, and the roadmap to our next milestone.
 
@@
 
-LabMate is under constant development; the latest version can be found at \url{https://github.com/msp-strath/LabMate/tree/devel}.
-
 \section{Supporting dimensional consistency}
@@
 
-\bibliographystyle{ws-procs9x6}
+%\bibliographystyle{ws-procs9x6}
+\bibliographystyle{plain}
 \bibliography{labmate}

### 2024-04-12 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] section 4 expanded

@@
 
-\section{Implementation}
+\section{Implementation and Current Status}
 
@@
 
+The meaning of whitespace in MATLAB is highly context-sensitive --- e.g., $\mbox{\texttt{(2 - 2)}}$, $\mbox{\texttt{[2 - 2]}}$ and $\mbox{\texttt{(2 -2)}}$ all meaning zero, but $\mbox{\texttt{[2 -2]}}$ being a $2\times1$ matrix.
+%
+Another precursor to typechecking is the deduction of variable scope: as proof of concept, we implemented the directive
+\begin{verbatim}
+%> rename x y
+\end{verbatim}
+which requests LabMate to rename all and only the occurrences of the \texttt{x} in scope at that point to \texttt{y}, but only if that does not conflict with other things called \texttt{y}.
+%
+To our surprise, we discovered that a similar facility offered by MATLAB's own editor sometimes changes the semantics of programs, which is presumably not what the user intends.
+%
+Writing a typechecker, we cannot afford the luxury of waiting until runtime to figure out which $\texttt{x}$ is which.
+
 LabMate elaborates MATLAB expressions and commands into terms of its own internal core type theory.
 %
-Because we want to ergonomically handle physical dimensions, matrices
-and their types, this type theory has a richer equational theory than
-other state-of-the-art type theories, which we achieve by a refined
-normalisation algorithm for free Abelian groups, and free monoids.
+Although the simple examples in this paper use \emph{constant} sizes and dimensions, any practicable library requires size, header and dimension expressions to contain \emph{parameters}.
 %
-The actual typechecking is implemented as a stack based virtual machine which uses unification and a limited form of backtracking to solve typing constraints.
+For example, if \texttt{A} is $i\times j$ and \texttt{B} is $i\times k$, then
+\begin{verbatim}
+[ A B
+  B A ]
+\end{verbatim}
+should be a valid matrix pasting, because $j+k = k+j$. Our notion of equality for numbers thus induces commutative monoid structure.
 %
-After the virtual machine has solved as many problems as it can, we analyse its final state to reconstruct the output source code, perhaps with inserted warnings and error messages, or additional responses to queries.
+Physical dimensions are modelled as the free Abelian group generated by some base set of dimensions~\cite{dim.group}, so we decide that theory, too.
+%
+The core theory we need is in place.
+
+The actual typechecking is implemented as a stack based virtual machine which traverses the syntax tree, deducing its translation into the core theory by propagating type information, generating typing constraints, and solving them by unification~\cite{robinson65}.
+%
+In order to support a much more expressive language of types, we must move on from the classic requirement that all constraints can be solved or rejected on first inspection and hence that the program can be processed exactly once and in a specific order~\cite{damas.milner}.
+%
+Our virtual machine supports repeated refocusing to wherever in the program there is progress to be made.
+%
+Moreover, as MATLAB notation is somewhat overloaded (e.g., \texttt{*} can mean either scalar or matrix multiplication), so LabMate elaboration has a (limited) need for backtracking search, so that, too, is facilitated.
+
+Elaboration for matrix multiplication and addition currently generate the correct constraints, but to check our examples, we need to improve our constraint solver and also implement the $\bsl$ operator.
+%
+As division is specified in terms of multiplication, its typing discipline requires a similar coincidence of header lists, but for the rows of both inputs: the technology we need is more of the same.
+
+After the virtual machine has solved as many problems as it can, we analyse its final state to reconstruct the output source code, perhaps with inserted warnings and error messages, or additional responses to queries and generated code.
 

### 2024-04-12 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] dimensions

@@
 \newcommand{\rsb}{\texttt{\symbol{93}}}
+\newcommand{\crt}{\texttt{\symbol{94}}}
 
@@
 \end{verbatim}
-Did you notice the error in the above code? The vector \texttt{voltages} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate detects this error and responds:
+Did you notice the error in the above code? The vector \texttt{voltages} should be of size $12 \times 1$, but is erroneously of size $1 \times 12$ instead. LabMate will detect this error and respond:
 
@@
 %
-That is, LabMate says that for \texttt{x} to have a sensible type, we need \texttt{M} and \texttt{voltages} to have the same number of rows.
+That is, for \texttt{x} to have a sensible type, we need \texttt{M} and \texttt{voltages} to have the same number of rows.
 We forgot to transpose \texttt{voltages}!
@@
 \]
-as in our \texttt{M :: [ 1 x 5 ] double} in Section~\ref{sec:example}.
+as in our \texttt{M :: [ 12 x 5 ] double} in Section~\ref{sec:example}.
 
@@
 
-Coming back to the example from Section~\ref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
+Coming back to the example from Section~\ref{sec:example} again, we aim to give both \texttt{M} and \texttt{voltages} more precise types, also taking their physical dimensions into account,
 \begin{verbatim}
-%> A :: [ i <- 12,  j <- 5 ]
-          (if j == 4 then Q T else Q 1)
+%> M :: [ 12 x i::D<-[{},{},{},{},T] ] Q(i)
+%> voltages :: [ 1 x 12 ] Q(V)
+\end{verbatim}
+  but we had better unpack what we mean.
 
-%> b :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
+Firstly, \texttt{M}'s row header type remains $\One$, so our row header list is still abbreviated by a number.
+
+Secondly, we see \texttt{Q(}$d$\texttt{)} as the type of \remph{quantities} with physical dimension $d$, represented in MATLAB as a mere \texttt{double} with standardised units.
+We shall define a type $\texttt{D}$ of dimensions presently.
+The notation
+\begin{verbatim}
+i::D<-[{},{},{},{},T]
 \end{verbatim}
-This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities.
-Meanwhile all the entries of \texttt{b} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$.
-LabMate reports that the type of \texttt{x} should be
+will tell LabMate that the column header type is $\texttt{D}$, and that the matrix entry type \texttt{Q(i)} varies in accordance with the header, locally named \texttt{i}, which is drawn from the given list of dimensions.
+%
+But what type is \texttt{D}? We will have defined, in LabMate directives, rather than MATLAB,
 \begin{verbatim}
-%< x :: [ i <- 5 ]
-        (if i == 4
-           then Q (ML^2T^(-4)I^(-1))
-           else Q (ML^2T^(-3)I^(-1)))
+%{>
+Base =
+  ['Time, 'Length, 'Mass, 'Current,
+   'Temperature, 'Amount, 'Luminosity]
+D = Abel(Enum(Base))
+%}
 \end{verbatim}
-that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first four rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) in the final row.
-%
-Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{x}, and to leave the type of some of the entries in \texttt{b} as an unknown
+where \texttt{Base} is a list of symbolic constants standing for base dimensions, \texttt{Enum} converts any such list to a finite enumeration type, and \texttt{Abel} constructs the free Abelian group on any type of generators, with unit \lcb\rcb, singleton generators \lcb$g$\rcb, multiplication \texttt{*}, division \texttt{/}, and exponentiation by integer constants \crt.
+We further define
 \begin{verbatim}
-%> b :: [ i <- 5 ]
-        (if i == 4
-           then ?bTy
-           else Q (ML^2T^(-3)I^(-1)))
+%{>
+T = {'Time}
+V = {'Mass}*{'Length}^2/T^3/{'Current}
+%}
 \end{verbatim}
-We can then ask LabMate to solve the \remph{meta variable} \texttt{?bTy}, to which LabMate responds that we must also have \texttt{?aTy} = $\mathsf{Q}\,(\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1})$.
-This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.
+and \texttt{voltages} becomes a row vector uniformly containing quantities in \texttt{Q(V)}.
 
-%GravCalc example again: $t_i$ time ($\mathsf{T}$; in seconds); $v_i$ voltage ($\mathsf{ML}^2\mathsf{T}^{-3}\mathsf{I}^{-1}$; in volt)
+With this refinement in place
+\begin{verbatim}
+x = M \ voltages'
+%> typeof x
+\end{verbatim}
+will yield
+\begin{verbatim}
+%< x :: [ i::D<-[{},{},{},{},T] ] Q(V/i)
+\end{verbatim}
+
+In other words, \texttt{x} is a column vector comprising four quantities in \texttt{Q(V)} and a fifth in \texttt{Q(V/T)}.
+LabMate has thus propagated the dimensional variation from the type of \texttt{M} to the type of \texttt{x}, while retaining the uniformity of \texttt{V} in the type of \texttt{voltages}.
+Checking multiplicability for matrices of quantities \texttt{Q(}$d(i,j)$\texttt{)} and
+\texttt{Q(}$d(j,k)$\texttt{)} amounts to checking that $d(i,j)\texttt{*}d(j,k)$ is independent of $j$.
+In our example, there is no dimensional variation in the `middle' of the inputs, so multiplicability is evident.
+There is a degree of freedom to multiply headers by an arbitrary dimension and compensate in the entry type, but we avoid ambiguity by sticking to Hart's discipline~\cite{hart}, namely
+\begin{enumerate}
+\item ensure any list of dimensions starts with the unit, \lcb\rcb;
+\item divide by dimensions coming from a row header;
+\item multiply by dimensions coming from a column header.
+\end{enumerate}
+Valid operations which respect Hart's discipline will always have header lists which meet in the middle exactly and yield outputs which retain Hart's discipline~\cite{dimTypes}.
 

### 2024-04-15 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] redrafted conclusion; some bib fixes

@@
 
-We have demonstrated how LabMate can help write MATLAB programs that are correct both in the sense that they avoid runtime errors, as well as dimensional inconsistency for physical dimensions.
+We have illustrated our prospectus for LabMate, a third-party programming assistant for MATLAB.
 %
-Getting an error for types not matching can sometimes feel like a stick, but there are also plenty of opportunities for carrots, for example where the specification encoded in the type of the program admits a single unique solution, which could be generated automatically by LabMate.
+Its type system offers expressivity well beyond the dynamic type tagging that MATLAB does itself.
 %
-Similarly, LabMate already has support for generating MATLAB code for serialising and deserialising complex data formats to and from files, based on a more high-level description of the data format (for example also taking units of measure into account)~\cite{mgen}.
+Typechecking with LabMate is no ironclad guarantee of complete correctness, but it is relatively lightweight and has the potential to rule out large classes of silly mistake, including dimensional inconsistency.
 
-So far, our focus has been on matrices, as this is what makes MATLAB
+However, if all LabMate does is complain about mistakes, it will not appeal to programmers in a hurry.
+%
+We have already shown a simple example where LabMate does not merely check but rather generates boilerplate code using its types as specification.
+%
+There is considerable potential to push further in this direction, little by little.
+%
+For example, we could equip LabMate with its own expression language for calculating with quantities which have not only dimensions but \remph{units}, compiling to MATLAB with appropriate conversion factors correctly inserted.
+%
+Our mission is, ever so gently, to tilt the balance towards programming at a higher level of structure and meaning in LabMate's directive language.
+
+So far, our focus has been on calculation with matrices, as this is what makes MATLAB
 unique as a programming language.
@@
 what is known about program fragments at runtime.
+%
+We also have our work cut out to retrofit LabMate types to MATLAB \remph{libraries}.
+
+Inspired again by Kennedy's work --- his gradualist reformation of PHP~\cite{hack} ---
+we have set out on this journey to bring the benefits of expressive type systems to scientists and engineers \remph{where they are}.
+%
+It would be unrealistic of us to expect practitioners to abandon their existing toolchains and learn whole new programming styles and languages that, in any case, exist only as academic research prototypes.
+%
+Moreover, it has very much \remph{not} turned out that, in addressing MATLAB, we were merely \remph{applying} type system research already well established.
+%
+This project has pushed us in new directions and generated raised expectations of the \remph{algebra} to be done by the typechecker.
+%
+There is so much to learn outside our ivory tower!
 

### 2024-04-16 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] fixes and suggestions by NPL

@@
 \affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , UK}  \\
-\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}}
-
+\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, UK}} % \\
+%\affiliationmark{*}{Corresponding author. Email address: \url{fredrik.nordvall-forsberg@strath.ac.uk}}}
 \maketitle
@@
   programming language MATLAB. However the high-level meaning of such
-  MATLAB programs stay informal, which can lead to implementation
+  MATLAB programs stays informal, which can lead to implementation
   errors and bugs, for example relating to incompatible units of
@@
 
-To rectify this situation, we have developed LabMate, which is a tool
+To rectify this situation, we are developing LabMate, which is a tool
 to reify current virtuous engineering practices as a formal language
@@
 Behind the scenes, LabMate is retrofitting an expressive type system
-to MATLAB, with the more of the meaning of programs recorded in their
+to MATLAB, with more of the meaning of programs recorded in their
 types.
@@
 
-LabMate is under constant development; the latest version can be found at \url{https://github.com/msp-strath/LabMate/tree/devel}.
-%
+ LabMate is under constant development; we encourage interested readers to get in touch for access to the latest version.
  At time of writing, the necessary core type system is in place: we are busily extending the variety of MATLAB expressions it can make sense of, and the constraints it can solve.
 %
+% LabMate is under constant development; the necessary core type system is in place: we are extending the variety of MATLAB expressions it can make sense of, and the constraints it can solve.
 We are close to covering the examples in this paper, but not quite there yet.
@@
 
-An electrical current with a known value (the applied signal) is passed through the resistor being measured, and a series of voltage readings are taken at fixed time intervals (the recorded output).
+An electric current with a known value (the applied signal) is passed through the resistor being measured, and a series of voltage readings are taken at fixed time intervals (the recorded output).
 After a specified number of readings are taken, the direction of the input signal must be reversed in order to separate the offset and drift from the voltage readings.
@@
 
-The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~1%\ref{fig:rec_out}
+The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~1 %\ref{fig:rec_out}
 provides an example of recorded output, using simulated data.
@@
 for i = 1:12
-  times[i] = src[readPtr];
+  times(i) = src(readPtr);
   readPtr = readPtr + 1;
@@
 for i = 1:12
-  voltages[i] = src[readPtr];
+  voltages(i) = src(readPtr);
   readPtr = readPtr + 1;
@@
 \end{verbatim}
-where the comment lines `\pct\res\lcb' and `\pct\res\rcb' visually delimit the extent of the generated boilerplate code, so that humans may safely ignore it.
+where the comment lines `\pct\res\lcb' and `\pct\res\rcb' visually delimit the extent of the code generated by LabMate, so that humans may safely ignore it.
 LabMate knows enough about scope in MATLAB to ensure that auxiliary variable naming never creates conflict.
@@
 %
-In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows ans columns of a spreadsheet do.
+In particular, we acknowledge that the rows and columns of a matrix might pertain to distinct individual entities, rather as the rows and columns of a spreadsheet do.
 %
@@
 \end{verbatim}
-should be a valid matrix pasting, because $j+k = k+j$. Our notion of equality for numbers thus induces commutative monoid structure.
+should be a valid matrix pasting, because $j+k = k+j$. Our notion of equality for numbers thus includes tacit applications of commutative and associative laws. % induces commutative monoid structure.
 %
@@
 \begin{verbatim}
-%> M :: [ 12 x i::D<-[{},{},{},{},T] ] Q(i)
+%> M :: [ 12 x k::D<-[{},{},{},{},T] ] Q(k)
 %> voltages :: [ 1 x 12 ] Q(V)
@@
 \begin{verbatim}
-i::D<-[{},{},{},{},T]
+k::D<-[{},{},{},{},T]
 \end{verbatim}
-will tell LabMate that the column header type is $\texttt{D}$, and that the matrix entry type \texttt{Q(i)} varies in accordance with the header, locally named \texttt{i}, which is drawn from the given list of dimensions.
+will tell LabMate that the column header type is $\texttt{D}$, and that the matrix entry type \texttt{Q(k)} varies in accordance with the header, locally named \texttt{k}, which is drawn from the given list of dimensions.
 %
@@
 \begin{verbatim}
-%< x :: [ i::D<-[{},{},{},{},T] ] Q(V/i)
+%< x :: [ k::D<-[{},{},{},{},T] ] Q(V/k)
 \end{verbatim}
@@
 %
-We have already shown a simple example where LabMate does not merely check but rather generates boilerplate code using its types as specification.
+We have already shown a simple example where LabMate does not merely check but rather generates code using its types as specification.
 %
@@
 %
-There is so much to learn outside our ivory tower!
+There is so much to learn from applications of dependent types in the wild!
 
@@
 Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from Section \ref{sec:example}.
-Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper, and to Professor Neil Ghani for his support.
+Thanks also to NPL colleagues Peter Harris, Louise Wright and Ian Smith for reviewing this paper, and to Professor Neil Ghani for his support.
 
@@
 %%% End:
+
+% LocalWords:  Idris

### 2024-08-26 writing/IMEKOWC2024/presentation/presentation.tex
[ writing ] imeko2024 presentation

\documentclass[]{beamer}
\usepackage[utf8x]{inputenc}
\usepackage{pgfpages}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{lstautogobble}
\usepackage{xspace}
\usepackage{xparse}
\usepackage{matlab-prettifier}
%\setbeameroption{show notes on second screen}

\usepackage{microtype}

\usepackage{tikz}
\usetikzlibrary{shapes.callouts, shapes.geometric, arrows, arrows.meta, chains}

% speech bubbles
\tikzset{
  invisible/.style={opacity=0,text opacity=0},
  visible on/.style={alt=#1{}{invisible}},
  alt/.code args={<#1>#2#3}{%
    \alt<#1>{\pgfkeysalso{#2}}{\pgfkeysalso{#3}} % \pgfkeysalso doesn't change the path
  },
}
\newcommand{\tikzmark}[1]{\tikz[overlay,remember picture,baseline=-0.5ex] \node (#1) {};}

\NewDocumentCommand{\mycallout}{r<> O{opacity=0.8,text opacity=1} m m}{%
  \tikz[remember picture, overlay]\node[align=center, fill=blue!20, text width=12em,
  #2,visible on=<#1>, rounded corners,
  draw,rectangle callout,anchor=pointer,callout relative pointer={(180:1em)}]
  at (#3) {#4};
}
\NewDocumentCommand{\mycalloutup}{r<> O{opacity=0.8,text opacity=1} m m}{%
  \tikz[remember picture, overlay]\node[align=center, fill=blue!20, text width=12em,
  #2,visible on=<#1>, rounded corners,
  draw,rectangle callout,anchor=pointer,callout relative pointer={(240:1em)}]
  at (#3) {#4};
}

% flowcharts
\tikzset{
  reg/.style={
    rectangle,
    rounded corners,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=black
  },
  arrow/.style={thick,->},
  fade/.style={
    rectangle,
    rounded corners,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=gray!30,
    text=gray!30,
  },
  fadearrow/.style={thick,->, gray!30},
  highlight/.style={
    rectangle,
    rounded corners,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=black,
    fill=blue!10
  },
  end/.style={
    tape,
    tape bend top=none,tape bend height=2mm,
    tape bend bottom=out and in,
    minimum width=3cm,
    minimum height=1cm,
    align=center,
    draw=black
  },
  fadeend/.style={
  tape,
  tape bend top=none,tape bend height=2mm,
  tape bend bottom=out and in,
  minimum width=3cm,
  minimum height=1cm,
  align=center,
  draw=gray!30,
  text=gray!30
  }
}

\lstset{ %
  language=Matlab,
  style=Matlab-editor,
  autogobble=true,
  escapechar=!,
  belowskip=0em
}

\newcommand{\pct}{\texttt{\symbol{37}}}
\newcommand{\dir}{\texttt{\symbol{62}}}
\newcommand{\res}{\texttt{\symbol{60}}}
\newcommand{\lcb}{\texttt{\symbol{123}}}
\newcommand{\rcb}{\texttt{\symbol{125}}}
\newcommand{\lsb}{\texttt{\symbol{91}}}
\newcommand{\rsb}{\texttt{\symbol{93}}}
\newcommand{\keyword}[1]{\textcolor{blue}{#1}}
\newcommand{\lm}{\textsc{LabMate}\xspace}
\newcommand{\repo}{\url{https://github.com/msp-strath/LabMate}}
\newcommand{\ma}{\textsc{Matlab}\xspace}
\newcommand{\List}{\mathsf{List}}
\newcommand{\Unit}{\mathsf{Unit}}
\newcommand{\Abel}{\mathsf{Abel}}
\newcommand{\Int}{\mathsf{Int}}
\newcommand{\Matrix}{\mathsf{Matrix}}
\newcommand{\Enum}{\mathsf{Enum}}
\newcommand{\LL}{\mathsf{L}}
\newcommand{\TT}{\mathsf{T}}
\newcommand{\MM}{\mathsf{M}}
\newcommand{\Kg}{\mathsf{kg}}
\newcommand{\Sec}{\mathsf{s}}
\newcommand{\Km}{\mathsf{m}}
\newcommand{\NameTimes}{\!\!\times\!\!}

\setbeamertemplate{frametitle}[default][center]
\title{\huge \lm: \\ \centerline{Supporting Types for \ma} }
\author[McBride, Nakov, Nordvall Forsberg et al]{\small Conor McBride$^{1}$, \underline{Georgi Nakov}$^{1}$, Fredrik Nordvall Forsberg$^{1}$,\\ Andr\'e Videla$^{1}$, Alistair Forbes$^{2}$, Keith Lines$^{2}$}
\institute[]{$^{1}$University of Strathclyde, UK\\$^{2}$National Physical Laboratory, UK}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}{The Problem}
  \begin{itemize}
  \item Loads of computational software in science and engineering is
    written in \ma
    \begin{itemize}
    \item may contain errors and bugs, as with any software
    \end{itemize}
  \item Developers often leave comments what are the corresponding
    physical systems of their data and how it should be interpreted,
    e.g., units of measure for quantities.
  \item \ma is oblivious to these high-level, semantic comments, and
    instead performs low-level compatibility checks during execution.
  \end{itemize}
\end{frame}

\begin{frame}{Our Plan}
  Can we do better?
  \begin{itemize}
  \item Make these developers' comments formal
  \item \ldots and create a tool to make use of them --- \lm
    \begin{itemize}
    \item keep the existing \ma code and toolchains, no need to
      rewrite in a new language
    \end{itemize}
  \item Distill the essence of the developers' comments in
    \lm's expressive type system
    \begin{itemize}
    \item a set of logical rules that assign a domain of admissible
      values to the expressions in our program
    \end{itemize}
  \item Run \lm multiple times while writing the code to get instant
    feedback and guidance, do not delay until execution
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{How does \lm Work?}
  \begin{itemize}
  \item \lm is a program transducer: reads \ma code with formal comments, and outputs a modified version of the input
  \item These formal comments are directives --- they start with \texttt{\%<}
  \item Input the program:
   \begin{lstlisting}[xleftmargin=2em]
    %> rename n x !\tikzmark{ren}!
    n = 5;
    display(n);
   \end{lstlisting}
  \pause
  \mycallout<+>{ren}{this is an input directive to rename a variable}
  \pause
\item Run \lm to get:
  \begin{lstlisting}[xleftmargin=2em]
%< LabMate 0.2.0.0 !\tikzmark{ver}!
%< renamed n x !\tikzmark{renr}!
x = 5;
display(x);
   \end{lstlisting}
  \pause
  \mycallout<+>{renr}{LabMate response to the input directive}
  \mycallout<+>{ver}{mark the file as processed by LabMate}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Matrix Types}
  \begin{itemize}[<+->]
  \item Matrices feature heavily in \ma code
  \item \lm support type annotations for matrices
\begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
%> A :: [ 1 x 2 ] int  !\tikzmark{ann1}!
A = [ 2 3 ]
%> B :: [ 2 x 4 ] int  !\tikzmark{ann2}!
B = [ 1 1 1 1
      3 4 5 6 ]
C = A * B
\end{lstlisting}
  \pause
  \mycallout<+>{ann1}{type annotations at declaration of A}
  \mycallout<+>{ann2}{type annotations at declaration of B}
  \item We can ask for type information
\begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
%> typeof C !\tikzmark{query}!
%< C :: [Matrix 1 4 int]!\tikzmark{resp}!
\end{lstlisting}
 \pause
  \mycallout<+>{query}{query for the type of C}
  \mycalloutup<+>{resp}{LabMate can infer the dimensions}
  \item \ldots and can easily spot incompatible sizing
\begin{lstlisting}[xleftmargin=0em]
D = B * A
%> typeof D
%< The expression D is!\tikzmark{bad}! quite a puzzle
\end{lstlisting}
    \pause
      \mycalloutup<+>{bad}{LabMate can point out an error with D}
\end{itemize}
\end{frame}

\begin{frame}[fragile]{Matrix Types}
  \begin{itemize}[<+->]
  \item \lm processes arbitrary \ma code
    \begin{itemize}
    \item \lm does not rely on constant values, works on variables as well
    \end{itemize}
  \item
\begin{lstlisting}[xleftmargin=0em]
function B = f(A)
  %> B :: [ 1 x 3 ] int
  B = [ 1 A ]

  %> typeof A
  %< A :: [Matrix int 1 2]!\tikzmark{fromb}!
end

%> typeof A
%< The expression A is!\tikzmark{scope}! quite a puzzle
\end{lstlisting}
   \pause
   \mycalloutup<+>{fromb}{LabMate infers A from the type annotation on B}
   \mycalloutup<+>{scope}{tracks Matlab scope}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Dimensions and Quantities}
  \begin{itemize}[<+->]
  \item \lm has support for arbitrary quantities
    \begin{lstlisting}[xleftmargin=0em,belowskip=-1em]
%> dimensions V for Q over `Mass, `Time, `Length!\tikzmark{base}!
%> unit kg :: Q({!\tikzmark{unit1}! `Mass })!\tikzmark{unit}!
\end{lstlisting}
\pause
   \mycallout<+>{base}{define some base set of dimensions}
   \mycalloutup<+>{unit}{and a canonical unit of measure}
   \mycalloutup<+>{unit1}{can be arbitrary group expression over V}
\pause
\begin{lstlisting}[xleftmargin=0em]
%<{
kg = 1; !\tikzmark{kg}!
%<}
\end{lstlisting}
\pause
   \mycallout<+>{kg}{this is a "magic" response that LabMate emits}
\pause
\item we can then use the unit of measure to create quantities in our program:
\begin{lstlisting}[xleftmargin=0em]
y = 5*kg !\tikzmark{q}!
%> typeof y
%< y :: Quantity (Enum [`Mass, `Time, `Length]) {`Mass}
\end{lstlisting}
  \pause
   \mycallout<+>{q}{turn a value of a dimensionless type into a quantity}
  \end{itemize}
\end{frame}

\begin{frame}[fragile]{Dimensional Consistency for Matrices}
  \begin{itemize}[<+->]
  \item Most \ma program does not work on uniform matrices, the type of the entry $e_{i,j}$ might depend on the indices $i$ and $j$
  \item A common scenario when working with matrices of quantities
%    $\begin{array}{ccccc}
% e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
%         &        &         &        &         \\
% e_{i,1}  &        & e_{i,j}\tikzmark{mat} &        & e_{i,n} \\
%         &        &         &        &         \\
% e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
%\end{array}$
%    \mycallout<+>{mat}{the type might depend on i and j}
\begin{lstlisting}[xleftmargin=0em]
%> dimensions V for Q over `Length, `Mass, `Time
%> unit metre :: Q({ `Length })
%> unit kg :: Q({ `Mass })
%> unit sec :: Q({ `Time })
\end{lstlisting}
\item Work in progress: \lm support for such matrices
\begin{lstlisting}[xleftmargin=0em]
% > x :: [ i <- [{} {`Time}]
%        x j <- [{} {`Length}]
%        ] Q({`Mass * j / i})
x = [ 2*kg       5*kg*metre
      3*kg/sec   4*kg*metre/sec ]
\end{lstlisting}
  \end{itemize}
\end{frame}

\begin{frame}{Implementation Details}
  \begin{itemize}[<+->]
  \item \ma program are modelled as trees of commands, rather than sequence of commands
    \begin{itemize}
    \item the type information is propagated (consistently) throughout the tree; can put type annotations after variable declaration
    \item not every annotation is needed, document the interesting ones
    \end{itemize}
  \item \ma expressions are translated to \lm internal core type theory:
    \begin{itemize}
    \item Matrix types are parametrised over 5 parameters with dependencies between them
    \item Quantities are modelled as the free Abelian group over a base set of dimensions
    \item The typechecker understands some nontrivial algebraic properties
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}{Current Progress \& Future Plans}
  \begin{itemize}[<+->]
  \item \lm is under active development
    \begin{itemize}
    \item available on GitHub, please get it in touch if interested
    \end{itemize}
  \item Work in the pipeline:
    \begin{itemize}
    \item \keyword{Uniqueness of representation}: currently, a matrix with quantities can have more than one corresponding type; this might lead to odd behaviour during typechecking
      \item \keyword{Quality of life improvements}: better messages and more readable responses from \lm
      \end{itemize}
  \item We want to extend our coverage to loops and conditionals in the future.
  \end{itemize}
  \pause
  %\large \centerline{Thank you for your attention!}
\end{frame}

\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2024-08-26 src/CoreTT.hs
more aggressively infer type of string literals

@@ -569,10 +569,11 @@ termToNFList ty tm
       Just tm -> pure [Left tm]
       Nothing -> termToNFList ty tm
   | Intg i <- tm = withScope $ pure $ replicate (fromInteger i) (Right nil)
+  | Strg s <- tm = withScope $ pure $ [Right (SL s :$ U :^ no natty)]
   | Nil <- tm  = pure []
   | Just (Sone, [t]) <- tagEh tm = checkEval ty t >>= \r -> pure [Right r]
   | Just (Splus, [s, t]) <- tagEh tm = (++) <$> termToNFList ty s <*> termToNFList ty t
-  | otherwise = error "termToNFList: no"
+  | otherwise = withScope $ error $ "termToNFList: no\nty = " ++ show ty ++ "\n m = " ++ show tm
 
 termToNFAbel
   :: Typ ^ n     -- type of generators

### 2024-08-26 src/Machine.hs

@@ -1984,6 +1984,16 @@ runElabTask sol meta@Meta{..} etask = nattily (vlen mctxt) $ do
       pushProblems [xProb, yProb, zProb]
       newProb . Elab sol $ Await (conjunction [jc, hc, zstat]) zSol
       run
+    TypeExprTask whereAmI _ (TyStringLiteral s) -> do
+      (_, cs) <- constrain "IsListChar" $ Constraint
+          { constraintCtx = fmap Hom <$> mctxt
+          , constraintType = Hom (atom SType)
+          , lhs = mtype
+          , rhs = mk SList (atom SChar)
+          }
+      newProb . Elab sol $ Await cs (ixKI mtype (lit s))
+      run
+{-
     TypeExprTask whereAmI _ (TyStringLiteral s) -> case tagEh mtype of
       Just (SList, [genTy]) -> do
         (_, cs) <- constrain "IsChar" $ Constraint
@@ -1995,6 +2005,7 @@ runElabTask sol meta@Meta{..} etask = nattily (vlen mctxt) $ do
         newProb . Elab sol $ Await cs (ixKI mtype (lit s))
         run
       _ -> move worried
+-}
     TypeExprTask MatLab EnsureCompatibility (TyJux dir x (TyNil _ :<=: _)) -> do
       (rowTy, colTy, cellTy, rs, cs, tystat) <- ensureMatrixType mctxt mtype
       (xSol, xProb) <- elab "vjuxTop" mctxt (mk SMatrix rowTy colTy cellTy rs cs) (TypeExprTask MatLab EnsureCompatibility <$> x)
@@ -2384,6 +2395,7 @@ unelabType ty | Just ct <- tagEh ty = case ct of
         let sig = subSnoc (sub0 (R $^ nil <&> rowTy)) (R $^ nil <&> colTy)
         unelabType =<< normalise emptyContext (atom SType) (cellTy //^ sig)
   (SAbel, [ty]) | Just (SOne, []) <- tagEh ty -> pure [nom "int"]
+  (SList, [ty]) | Just (SChar, []) <- tagEh ty -> pure [nom "string"]
   (SEnum, [as]) -> do
     tas <- unelabTerm (mk SList $ atom SAtom) as
     pure $ [sym "Enum", spc 1] ++ tas

### 2024-08-26 src/Term.hs

@@ -211,6 +211,9 @@ instance Literal String  where ctor = SL
 pattern Intg :: () => (s ~ Chk) => Integer -> Term s ^ n
 pattern Intg i <- I i :$ U :^ _
 
+pattern Strg :: () => (s ~ Chk) => String -> Term s ^ n
+pattern Strg s <- SL s :$ U :^ _
+
 infixr 5 <&>
 (<&>) :: Term s ^ n -> Term t ^ n -> Term (Prd s t) ^ n
 (tl :^ th) <&> (tr :^ ph) | u :^ ps <- cop th ph = P tl u tr :^ ps

### 2024-08-26 writing/IMEKOWC2024/presentation/presentation.tex
[ IMEKO WC 2024 ] polish slides

@@
 \setbeamertemplate{frametitle}[default][center]
+\beamertemplatenavigationsymbolsempty
+
 \title{\huge \lm: \\ \centerline{Supporting Types for \ma} }
@@
   \begin{itemize}
-  \item Loads of computational software in science and engineering is
+  \item Much software in science and engineering.
     written in \ma
+    \smallskip
     \begin{itemize}
-    \item may contain errors and bugs, as with any software
+    \item May contain errors and bugs, as with any software.
     \end{itemize}
-  \item Developers often leave comments what are the corresponding
-    physical systems of their data and how it should be interpreted,
-    e.g., units of measure for quantities.
-  \item \ma is oblivious to these high-level, semantic comments, and
-    instead performs low-level compatibility checks during execution.
+    \bigskip
+  \item Developers often leave comments about how their data should be
+    interpreted, e.g., units of measure for quantities.
+%  \item Developers often leave comments what are the corresponding
+%    physical systems of their data and how it should be interpreted,
+%    e.g., units of measure for quantities.
+    \bigskip
+  \item However \ma is oblivious to these high-level comments, and
+    instead performs low-level %compatibility
+    checks during execution.
   \end{itemize}
@@
 \begin{frame}{Our Plan}
-  Can we do better?
+  \textit{Can we do better?}
+
   \begin{itemize}
-  \item Make these developers' comments formal
-  \item \ldots and create a tool to make use of them --- \lm
+    \medskip
+  \item Make these developer comments formal.
+    \medskip
+  \item \ldots and create a tool to make use of them --- \lm.
     \begin{itemize}
-    \item keep the existing \ma code and toolchains, no need to
-      rewrite in a new language
+      \smallskip
+    \item Keep existing \ma code and toolchains; no need to
+      switch to a new language.
     \end{itemize}
-  \item Distill the essence of the developers' comments in
-    \lm's expressive type system
+    \medskip
+  \item Distill the essence of the developer comments in
+    \lm's expressive type system.
     \begin{itemize}
-    \item a set of logical rules that assign a domain of admissible
-      values to the expressions in our program
+      \smallskip
+    \item A set of logical rules that assign domains of admissible
+      values to program expressions.
     \end{itemize}
-  \item Run \lm multiple times while writing the code to get instant
-    feedback and guidance, do not delay until execution
+    \medskip
+  \item \lm is meant to be used while writing the code to get instant
+    feedback and guidance --- do not delay until execution.
   \end{itemize}
@@
 
+\definecolor{commentgreen}{RGB}{034,139,034}
+
 \begin{frame}[fragile]{How does \lm Work?}
   \begin{itemize}
-  \item \lm is a program transducer: reads \ma code with formal comments, and outputs a modified version of the input
-  \item These formal comments are directives --- they start with \texttt{\%<}
+  \item \lm is a program transducer: reads \ma code with formal comments, and outputs a modified version of the input.
+    \medskip
+  \item These formal comments are directives --- they start with \textcolor{commentgreen}{\texttt{\%<}}.
+    \medskip
   \item Input the program:
@@
   \begin{lstlisting}[xleftmargin=2em]
-%< LabMate 0.2.0.0 !\tikzmark{ver}!
+%< LabMate 0.2.0.1 !\tikzmark{ver}!
 %< renamed n x !\tikzmark{renr}!
@@
   \begin{itemize}[<+->]
-  \item Matrices feature heavily in \ma code
-  \item \lm support type annotations for matrices
+  \item Matrices feature heavily in \ma code.
+    \medskip
+  \item \lm supports type annotations for matrices:
 \begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
 %> A :: [ 1 x 2 ] int  !\tikzmark{ann1}!
-A = [ 2 3 ]
+A = [ 3 4 ]
 %> B :: [ 2 x 4 ] int  !\tikzmark{ann2}!
 B = [ 1 1 1 1
-      3 4 5 6 ]
+      5 6 7 8 ]
 C = A * B
@@
   \mycallout<+>{ann2}{type annotations at declaration of B}
-  \item We can ask for type information
+  \item We can ask for type information:
 \begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
@@
   \mycalloutup<+>{resp}{LabMate can infer the dimensions}
-  \item \ldots and can easily spot incompatible sizing
+  \item \ldots and can easily spot incompatible sizing:
 \begin{lstlisting}[xleftmargin=0em]
@@
   \begin{itemize}[<+->]
-  \item \lm processes arbitrary \ma code
+  \item \lm processes arbitrary \ma code.
+    \smallskip
     \begin{itemize}
-    \item \lm does not rely on constant values, works on variables as well
+    \item \lm does not rely on constants, variables work as well.
     \end{itemize}
@@
 
+A = 'hello'
 %> typeof A
-%< The expression A is!\tikzmark{scope}! quite a puzzle
+%< A :: string!\tikzmark{scope}!
 \end{lstlisting}
    \pause
-   \mycalloutup<+>{fromb}{LabMate infers A from the type annotation on B}
+   \mycalloutup<+>{fromb}{LabMate infers type of A from the annotation on B}
    \mycalloutup<+>{scope}{tracks Matlab scope}
@@
   \begin{itemize}[<+->]
-  \item \lm has support for arbitrary quantities
+  \item \lm has support for arbitrary quantities.
     \begin{lstlisting}[xleftmargin=0em,belowskip=-1em]
-%> dimensions V for Q over `Mass, `Time, `Length!\tikzmark{base}!
+%> dimensions V for Q over `Mass!\tikzmark{base}!, `Time
 %> unit kg :: Q({!\tikzmark{unit1}! `Mass })!\tikzmark{unit}!
 \end{lstlisting}
-\pause
-   \mycallout<+>{base}{define some base set of dimensions}
+%\pause
+   \mycalloutup<+>{base}{define some base set of dimensions}
    \mycalloutup<+>{unit}{and a canonical unit of measure}
@@
 \pause
-   \mycallout<+>{kg}{this is a "magic" response that LabMate emits}
+   \mycallout<+>{kg}{this is a ``magic'' response that LabMate emits}
 \pause
-\item we can then use the unit of measure to create quantities in our program:
+\item We can then use the unit of measure for quantities:
 \begin{lstlisting}[xleftmargin=0em]
@@
 %> typeof y
-%< y :: Quantity (Enum [`Mass, `Time, `Length]) {`Mass}
+%< y :: Quantity (Enum [`Mass, `Time])
+%<               {`Mass}
 \end{lstlisting}
-  \pause
+%  \pause
    \mycallout<+>{q}{turn a value of a dimensionless type into a quantity}
@@
   \begin{itemize}[<+->]
-  \item Most \ma program does not work on uniform matrices, the type of the entry $e_{i,j}$ might depend on the indices $i$ and $j$
+  \item Most \ma programs do not work on uniform matrices: type of the entry $e_{i, j}$ might depend on the indices $i$ and $j$.
+    \medskip
   \item A common scenario when working with matrices of quantities
@@
 \begin{lstlisting}[xleftmargin=0em]
-%> dimensions V for Q over `Length, `Mass, `Time
-%> unit metre :: Q({ `Length })
-%> unit kg :: Q({ `Mass })
-%> unit sec :: Q({ `Time })
+%> dimensions V for Q over `L, `M, `T
+%> unit metre :: Q({ `L })
+%> unit kg :: Q({ `M })
+%> unit sec :: Q({ `T })
 \end{lstlisting}
+\medskip
 \item Work in progress: \lm support for such matrices
 \begin{lstlisting}[xleftmargin=0em]
-% > x :: [ i <- [{} {`Time}]
-%        x j <- [{} {`Length}]
-%        ] Q({`Mass * j / i})
-x = [ 2*kg       5*kg*metre
+% > A :: [ i <- [{} {`T}]
+%        x j <- [{} {`L}]
+%        ] Q({`M * j / i})
+A = [ 2*kg       5*kg*metre
       3*kg/sec   4*kg*metre/sec ]
@@
   \begin{itemize}[<+->]
-  \item \ma program are modelled as trees of commands, rather than sequence of commands
+  \item \ma programs are modelled as trees of commands, rather than sequence of commands.
+    \medskip
     \begin{itemize}
-    \item the type information is propagated (consistently) throughout the tree; can put type annotations after variable declaration
-    \item not every annotation is needed, document the interesting ones
+    \item Type information is propagated (consistently) throughout the tree; can put type annotations after variable declarations.
+      \medskip
+    \item Not every annotation is needed; document the interesting ones.
     \end{itemize}
-  \item \ma expressions are translated to \lm internal core type theory:
+    \medskip
+  \item \ma expressions are translated to \lm internal core type theory.
+    \medskip
     \begin{itemize}
-    \item Matrix types are parametrised over 5 parameters with dependencies between them
-    \item Quantities are modelled as the free Abelian group over a base set of dimensions
-    \item The typechecker understands some nontrivial algebraic properties
+    \item Matrix types are parametrised over 5 parameters with dependencies between them.
+      \medskip
+    \item Quantities are modelled as the free Abelian group over a base set of dimensions.
+      \medskip
+    \item The typechecker understands nontrivial algebraic properties.
     \end{itemize}
@@
   \begin{itemize}[<+->]
-  \item \lm is under active development
+  \item \lm is under active development.
+    \smallskip
     \begin{itemize}
-    \item available on GitHub, please get it in touch if interested
+    \item Available on GitHub, please get it in touch if interested.
     \end{itemize}
+    \bigskip
   \item Work in the pipeline:
+    \smallskip
     \begin{itemize}
-    \item \keyword{Uniqueness of representation}: currently, a matrix with quantities can have more than one corresponding type; this might lead to odd behaviour during typechecking
-      \item \keyword{Quality of life improvements}: better messages and more readable responses from \lm
+    \item \keyword{Uniqueness of representation}: currently, a matrix with quantities can have more than one corresponding type; this might lead to odd behaviour during typechecking.
+      \medskip
+    \item \keyword{Quality of life improvements}: better messages and more readable responses from \lm.
+      \medskip
       \end{itemize}
@@
   \end{itemize}
-  \pause
+  %\pause
   %\large \centerline{Thank you for your attention!}

### 2024-08-27 src/CoreTT.hs
[ progress ] more constraint solving

@@ -95,7 +95,6 @@ metaLookup s = TC $ \ st _ -> case s `Map.lookup` st of
 
 typeEh :: Term Chk ^ n -> TC n ()
 typeEh ty | Just cts <- tagEh ty = case cts of
-  (SOne, [])   -> pure ()
   (STwo, [])   -> pure ()
   (SAtom, [])  -> pure ()
   (SType, [])  -> pure ()
@@ -110,14 +109,14 @@ typeEh ty | Just cts <- tagEh ty = case cts of
   (SSig, [s, t]) | Just (x, t') <- lamNameEh t -> do
     typeEh s
     under (x, s) $ typeEh t'
-  (SMatrix, [rowTy, colTy, cellTy, rs, cs])
+  (SMatrix, [rowGenTy, colGenTy, cellTy, rs, cs])
     | Just (r, cellTy) <- lamNameEh cellTy
     , Just (c, cellTy) <- lamNameEh cellTy -> withScope $ do
-      typeEh rowTy
-      typeEh colTy
-      under (r, rowTy) $ under (c, wk colTy) $ typeEh cellTy
-      checkEh (mk SList rowTy) rs
-      checkEh (mk SList colTy) cs
+      typeEh rowGenTy
+      typeEh colGenTy
+      under (r, rowGenTy) $ under (c, wk colGenTy) $ typeEh cellTy
+      checkEh (mk SList (tag SAbel [rowGenTy])) rs
+      checkEh (mk SList (tag SAbel [colGenTy])) cs
   (SDest, [t]) -> typeEh t
   (SQuantity, [g, v]) -> withScope $ do
     typeEh g
@@ -169,7 +168,6 @@ checkCanEh ty tm | Just x <- tagEh ty = withScope $ case x of
   (SEnum, [as]) -> do
     nfs <- termToNFList (atom SAtom) as
     True <$ checkEnumEh nfs tm
-  (SOne, []) -> pure True
   (STwo, []) | Intg i <- tm, i == 0 || i == 1 -> pure True
   (SChar, []) | Intg i <- tm, i >= 0 && i <= 128 -> pure True
   (SAtom, []) | Atom _ <- tm -> pure True
@@ -179,12 +177,11 @@ checkCanEh ty tm | Just x <- tagEh ty = withScope $ case x of
     checkEh s a
     True <$ checkEh (t' //^ sub0 (R $^ a <&> s)) d
[…]

### 2024-08-27 src/Machine.hs

@@ -6,17 +6,23 @@ import Control.Monad
 import Control.Newtype
 import Control.Monad.State
 import Data.Char
+import Data.Either
 import Data.List
 import Data.Maybe (fromMaybe, isNothing)
 import Data.Monoid
 import Data.Map (Map)
 import qualified Data.Map as Map
 
+import Data.Traversable
+
 -- mgen
 import TranslateContext (TargetLang(..))
 import Translate
 
+import MissingLibrary
+
 import Bwd
+import Term.Indexed
 
 import Lex
 import Parse
@@ -37,6 +43,8 @@ import Control.Monad.Reader
 debug = const id -- trace
 debugCatch = const id -- trace
 debugMatrix = const id -- trace
+debug260824 = trace
+
 
 type Elab = State MachineState
 
@@ -385,6 +393,7 @@ prob = llup >>= \case
     push $ LocalNamespace oldNs
     prob
   Just (ConstraintFrame name c) -> do
+    c <- updateConstraint c
     debug ("`prob` attempting to solve constraint " ++ show c) $ solveConstraint name c
     prob
   Just f  -> do
@@ -459,6 +468,54 @@ lhsType (Het lty _ _) = lty
 rhsType (Hom ty) = ty
 rhsType (Het _ _ rty) = rty
 
+data SimplifiedContext m = forall n. SimplifiedContext
+  { simplifiedContext :: ConstraintContext n
+  , simplifySubst :: Subst m ^ n -- strengthening
+  }
+
+nonUnitTypes :: Vec j (String, ConstraintType m) -> Ex (Flip (<=) j)
+nonUnitTypes VN = Ex (Flip Ze)
+nonUnitTypes (ga :# (_, cty)) = case nonUnitTypes ga of
+  Ex (Flip th) -> case cty of
+    Hom t | isUnitType t -> Ex (Flip (No th))
+          | otherwise -> Ex (Flip (Su th))
+
+strengthenNil :: i <= j -> Subst j i
+strengthenNil Ze = Sub0
+strengthenNil (No th) = SubT (strengthenNil th) (leftAll (no (weeEnd th))) (R :$ P (A "" :$ U) ZZ oneTypeZ)
+strengthenNil (Su th) = wkSubst (strengthenNil th)
+
+simplifyContext :: ConstraintContext m -> SimplifiedContext m
+simplifyContext ga = case nonUnitTypes ga of
[…]

### 2024-08-27 src/MagicStrings.hs

@@ -2,7 +2,6 @@ module MagicStrings where
 
 -- core types
 pattern SType = "Type"
-pattern SOne  = "One"
 pattern STwo  = "Two"
 pattern SAbel = "Abel"
 pattern SList = "List"

### 2024-08-27 src/Term/Thinning.hs

@@ -116,6 +116,9 @@ data (^) :: (Nat -> Type) -> Nat -> Type where
 scopeOf :: p ^ n -> Natty n
 scopeOf (_ :^ th) = bigEnd th
 
+withScopeOf :: p ^ n -> (NATTY n => b) -> b
+withScopeOf x k = nattily (scopeOf x) k
+
 -- (^) is the free Thinny on (p :: Nat -> Type)
 instance Thinny ((^) (p :: Nat -> Type)) where
   (p :^ th) -< ph = p :^ (th -< ph)

### 2024-08-27 src/Machine.hs
[ more ] structural subtyping

@@ -980,6 +980,10 @@ solveConstraint name c@SubtypingConstraint{..} = let n = vlen constraintCtx in n
         (genTyWant, aStat) <- ensureAbel wantCtx wantType
         stat <- subtype "subTypeAbelGot" genTyGot genTyWant
         metaDefn name (conjunction [aStat, stat])
+      Just (tag, [elTyGot]) | tag `elem` [SList, SDest] -> do
+        (elTyWant, dStat) <- ensureUTag tag wantCtx wantType
+        stat <- subtype "subTypeUTagGot" elTyGot elTyWant
+        metaDefn name (conjunction [dStat, stat])
       Just (SEnum, [genTyGot]) -> do
         (genTyWant, eStat) <- ensureEnum wantCtx wantType
         stat <- prefixConstraint constraintCtx genTyGot genTyWant
@@ -1008,6 +1012,10 @@ solveConstraint name c@SubtypingConstraint{..} = let n = vlen constraintCtx in n
           (genTyGot, aStat) <- ensureAbel gotCtx gotType
           stat <- subtype "subTypeAbelWant" genTyGot genTyWant
           metaDefn name (conjunction [aStat, stat])
+        Just (tag, [elTyWant]) | tag `elem` [SList, SDest] -> do
+          (elTyGot, dStat) <- ensureUTag tag gotCtx gotType
+          stat <- subtype "subTypeUTagGot" elTyGot elTyWant
+          metaDefn name (conjunction [dStat, stat])
         Just (SEnum, [genTyWant]) -> do
           (genTyGot, eStat) <- ensureEnum gotCtx gotType
           stat <- prefixConstraint constraintCtx genTyGot genTyWant
@@ -1034,7 +1042,7 @@ solveConstraint name c@SubtypingConstraint{..} = let n = vlen constraintCtx in n
         _ -> push $ ConstraintFrame name SubtypingConstraint{..}
  where
   prefixConstraint :: NATTY n => ConstraintContext n -> Term Chk ^ n -> Term Chk ^ n -> Elab BOOL
-  prefixConstraint ctx Nil ws = pure $ conjunction []
[…]

### 2024-09-09 src/Machine.hs
[ pprint ] slightly better /

@@ -2993,14 +2993,20 @@ unelabTerm ctx ty tm | Just ty <- tagEh ty = case ty of
                 1 -> t
                 _ -> t ++ [sym "^", Tok (show n) Dig dump]
     go genTy tm | Just (Splus, [x, y]) <- tagEh tm = do
-       x <- go genTy x
        case tupEh y of
          Just [Intg n, y] | n < 0 -> do
+            x <- go genTy x
             y <- go genTy (tup [lit (-n), y])
             pure $ x ++ [spc 1, sym "/", spc 1] ++ y
-         _ -> do
-            y <- go genTy y
-            pure $ x ++ [spc 1, sym "*", spc 1] ++ y
+         _ -> case tupEh x of
+                Just [Intg n, x] | n < 0 -> do
+                  x <- go genTy (tup [lit (-n), x])
+                  y <- go genTy y
+                  pure $ y ++ [spc 1, sym "/", spc 1] ++ x
+                _ -> do
+                  x <- go genTy x
+                  y <- go genTy y
+                  pure $ y ++ [spc 1, sym "*", spc 1] ++ x
     go genTy tm = pure [sym $ show tm]
 
 diagnosticMove :: ForkCompleteStatus -> Elab ()

### 2024-09-09 src/Machine.hs
[ print ] pretty printing for Quantity

@@ -2936,6 +2936,7 @@ unelabType ctx ty | Just ct <- tagEh ty = case ct of
   (SEnum, [as]) -> do
     tas <- unelabTerm ctx (mk SList $ atom SAtom) as
     pure $ [sym "Enum", spc 1] ++ tas
+  _ | ty == doubleType -< no (scopeOf ty) -> pure [nom "double"]
   (SQuantity, [enumTy, exp]) -> do
         tenum <- unelabType ctx enumTy
         texp <- unelabTerm ctx (mk SAbel enumTy) exp
@@ -2957,7 +2958,7 @@ unelabTerm ctx ty tm | Just ty <- tagEh ty = case ty of
          -> Bool -- are we in the middle of printing rights
          -> Elab [Tok]
       go ctx ty [] True = pure [sym "]"]
-      go ctx ty [] False = pure []
+      go ctx ty [] False = pure [sym "[]"]
       go ctx ty [Left x] rightEh = do
         ttm <- unelabTerm ctx ty (E $^ x)
         pure $ (if rightEh then [sym "]"] else []) ++ ttm

### 2024-09-09 src/CoreTT.hs
[ new ] support for subtraction

@@ -13,7 +13,7 @@ import Term
 
 import Debug.Trace
 
-track = const id --trace
+track = const id -- track
 
 data Status = Crying | Waiting | Hoping | Abstract | ProgramVar
   deriving (Ord, Eq, Show)
@@ -534,6 +534,9 @@ evalSynth tm = withScope $ case tm of
         , Atom Stranspose <- dstr -> do
             let sg = subSnoc (subSnoc (idSubst natty :^ No (No (io natty))) (var 0)) (var 1)
             pure (E $^ D $^ ((R $^ (tgt <&> tgtTy)) <&> dstr), mk SMatrix colGenTy rowGenTy (lam c $ lam r $ cellTy //^ sg) cs rs)
+      Just (SMatrix, [rowGenTy, colGenTy, cellTy, rs, cs])
+        | Atom Suminus <- dstr ->
+            pure (E $^ D $^ ((R $^ (tgt <&> tgtTy)) <&> dstr), tgtTy)
       _ -> fail "evalSynth: eliminator for an unknown type"
   tm | Just tm <- MX $? tm, (lmx, rmx) <- split tm -> do
     (lmx, lmxTy) <- evalSynth lmx

### 2024-09-09 src/Machine.hs

@@ -865,7 +865,7 @@ data FlexEh (gamma :: Nat) where
     -> Subst n ^ delta -- how to map the permitted variables into the
                        -- flex variable's context
     -> FlexEh gamma
-    
+
 data FlexEhWork (gamma' :: Nat)(delta' :: Nat) where
   FlexEhWork
     :: n <= gamma'      -- the thinning for the candidate solution's
@@ -2338,8 +2338,7 @@ runElabTask sol meta@Meta{..} etask = nattily (vlen mctxt) $ do
       run
     TypeExprTask whereAmI synthMode (TyBinaryOp Plus x y) -> do
       -- TODO:
-      -- 1. make sure `mtype` admits plus
-      -- 2. find the correct way of doing plus in `mtype`
+      -- 1. find the correct way of doing plus in `mtype`
       (xTy, yTy, stats) <- case synthMode of
         EnsureCompatibility -> pure (mtype, mtype, [])
         FindSimplest -> do
@@ -2356,6 +2355,12 @@ runElabTask sol meta@Meta{..} etask = nattily (vlen mctxt) $ do
       (ySol, yProb) <- elab "plusYTy" mctxt yTy (TypeExprTask whereAmI FindSimplest <$> y)
       pushProblems [xProb, yProb]
       newProb . Elab sol $ Await (conjunction stats) (mk Splus xSol ySol) -- FIXME: do proper addition, eg if mtype is a matrix type
+      move winning
+    TypeExprTask whereAmI synthMode (TyUnaryOp UMinus x) -> do
+      (_, _, _, _, _, stat) <- ensureMatrix mctxt mtype
+      (xSol, xProb) <- elab "minusXTy" mctxt mtype (TypeExprTask whereAmI synthMode <$> x)
+      pushProblems [xProb]
+      newProb . Elab sol $ Await stat (E $^ matrixUminus (R $^ xSol <&> mtype))
       move winning
     TypeExprTask whereAmI synthMode (TyBinaryOp (Mul False{- x*y -} Times) x y) -> do
       -- Two main issues:
@@ -2444,48 +2449,101 @@ runElabTask sol meta@Meta{..} etask = nattily (vlen mctxt) $ do
           }
       newProb . Elab sol $ Await cs (ixKI mtype (lit s))
       run
-{-
-    TypeExprTask whereAmI _ (TyStringLiteral s) -> case tagEh mtype of
-      Just (SList, [genTy]) -> do
-        (_, cs) <- constrain "IsChar" $ Constraint
[…]

### 2024-09-09 src/MagicStrings.hs

@@ -19,6 +19,7 @@ pattern Sfst  = "fst"
 pattern Ssnd  = "snd"
 pattern Sinverse = "inverse"
 pattern Stranspose = "transpose"
+pattern Suminus = "uminus"
 
 -- constructors
 pattern Splus = "plus"

### 2024-09-09 src/Syntax.hs

@@ -136,6 +136,11 @@ toTypeExpr' (BinaryOp (Mul d LDiv) x y) = do
   x <- toTypeExpr x
   y <- toTypeExpr y
   pure (TyBinaryOp (Mul d Times) (TyUnaryOp UInvert x :<=: xsrc) y)
+toTypeExpr' (BinaryOp Minus x y) = do
+  let ysrc = source y
+  x <- toTypeExpr x
+  y <- toTypeExpr y
+  pure (TyBinaryOp Plus x (TyUnaryOp UMinus y :<=: ysrc))
 toTypeExpr' (BinaryOp op x y) = TyBinaryOp op <$> toTypeExpr x <*> toTypeExpr y
 toTypeExpr' (Mat exps) = do
    exps <- traverse (traverse toTypeExpr) exps

### 2024-09-09 src/Term.hs

@@ -63,6 +63,9 @@ matrixInverse t = withScopeOf t $ (D $^ t <&> atom Sinverse)
 matrixTranspose :: Term Syn ^ n -> Term Syn ^ n
 matrixTranspose t = withScopeOf t $ (D $^ t <&> atom Stranspose)
 
+matrixUminus :: Term Syn ^ n -> Term Syn ^ n
+matrixUminus t = withScopeOf t $ (D $^ t <&> atom Suminus)
+
 type Root = Bwd (String, Int)
 type NameSupply = (Root, Int)
 newtype Name = Name { longname :: [(String, Int)] }

### 2024-09-09 writing/IMEKOWC2024/presentation/presentation.tex
[ IMEKO WC 2024 ] slide numbers

@@
 
+\defbeamertemplate*{footline}{my theme}
+{
+  \leavevmode%
+  \hbox{%
+    \usebeamerfont{date in head/foot}
+    \hspace*{0.97\paperwidth}\raisebox{2pt}{\insertframenumber{}}}
+  \vskip0pt%
+}
+
 \title{\huge \lm: \\ \centerline{Supporting Types for \ma} }
@@
 
-\begin{frame}
+\begin{frame}[plain]
   \titlepage
@@
 
+\setcounter{framenumber}{0}
+
 \begin{frame}{The Problem}
-  \begin{itemize}
-  \item Much software in science and engineering.
-    written in \ma
+  \begin{itemize}[<+->]
+  \item Much science and engineering software is written in \keyword{\ma}.
     \smallskip
-    \begin{itemize}
-    \item May contain errors and bugs, as with any software.
+    \begin{itemize}[<.->]
+    \item May contain \keyword{errors} and \keyword{bugs}, as with any software.
     \end{itemize}
     \bigskip
-  \item Developers often leave comments about how their data should be
-    interpreted, e.g., units of measure for quantities.
+  \item Developers often leave comments about how their
+    \keyword{data} should be \keyword{interpreted}, e.g., units of measure for
+    quantities.
 %  \item Developers often leave comments what are the corresponding
@@
     \bigskip
-  \item However \ma is oblivious to these high-level comments, and
-    instead performs low-level %compatibility
-    checks during execution.
+  \item However \ma is \keyword{oblivious} to these high-level
+    comments, and instead performs \keyword{low-level checks} during
+    execution.
   \end{itemize}
@@
 
-  \begin{itemize}
+  \begin{itemize}[<+->]
     \medskip
@@
   \item \ldots and create a tool to make use of them --- \lm.
-    \begin{itemize}
+    \begin{itemize}[<.->]
       \smallskip
-    \item Keep existing \ma code and toolchains; no need to
+    \item \keyword{Keep existing} \ma \keyword{code} and toolchains; no need to
       switch to a new language.
@@
   \item Distill the essence of the developer comments in
-    \lm's expressive type system.
-    \begin{itemize}
+    \lm's \keyword{expressive type system}.
+    \begin{itemize}[<.->]
       \smallskip
@@
     \medskip
-  \item \lm is meant to be used while writing the code to get instant
-    feedback and guidance --- do not delay until execution.
+  \item \lm is meant to be used while writing the code to get
+    \keyword{instant feedback} and guidance --- do not delay until
+    execution.
   \end{itemize}
@@
 \begin{frame}[fragile]{How does \lm Work?}
-  \begin{itemize}
-  \item \lm is a program transducer: reads \ma code with formal comments, and outputs a modified version of the input.
+  \begin{itemize}[<+->]
+  \item \lm is a \keyword{program transducer}: reads \ma code with formal comments, and outputs a modified version of the input.
     \medskip
-  \item These formal comments are directives --- they start with \textcolor{commentgreen}{\texttt{\%<}}.
+  \item These formal comments are \keyword{directives} --- they start with \textcolor{commentgreen}{\texttt{\%<}}.
     \medskip
@@
   \mycallout<+>{ren}{this is an input directive to rename a variable}
-  \pause
 \item Run \lm to get:
@@
   \mycallout<+>{renr}{LabMate response to the input directive}
-  \mycallout<+>{ver}{mark the file as processed by LabMate}
+  \mycallout<+>{ver}{Mark the file as processed by LabMate}
 \end{itemize}
@@
   \begin{itemize}[<+->]
-  \item Matrices feature heavily in \ma code.
+  \item \keyword{Matrices} feature heavily in \ma code.
     \medskip
-  \item \lm supports type annotations for matrices:
+  \item \lm supports \keyword{type annotations} for matrices:
 \begin{lstlisting}[xleftmargin=0em, belowskip=-0.5em]
@@
     \smallskip
-    \begin{itemize}
-    \item \lm does not rely on constants, variables work as well.
+    \begin{itemize}[<.->]
+    \item \lm does not rely on constants, \keyword{variables} work as well.
     \end{itemize}
@@
   %> typeof A
-  %< A :: [Matrix int 1 2]!\tikzmark{fromb}!
+  %< A :: [Matrix 1 2 int]!\tikzmark{fromb}!
 end
@@
   \begin{itemize}[<+->]
-  \item \lm has support for arbitrary quantities.
+  \item \lm has support for arbitrary \keyword{quantities}.
     \begin{lstlisting}[xleftmargin=0em,belowskip=-1em]
@@
 \pause
-\item We can then use the unit of measure for quantities:
+\item We can then use the \keyword{unit of measure} for quantities:
 \begin{lstlisting}[xleftmargin=0em]
@@
   \begin{itemize}[<+->]
-  \item Most \ma programs do not work on uniform matrices: type of the entry $e_{i, j}$ might depend on the indices $i$ and $j$.
+  \item Most \ma programs \keyword{do not work on uniform matrices}: type of the entry $e_{i, j}$ might depend on the indices $i$ and $j$.
     \medskip
-  \item A common scenario when working with matrices of quantities
+  \item A common scenario when working with matrices of quantities:
 %    $\begin{array}{ccccc}
@@
   \begin{itemize}[<+->]
-  \item \ma programs are modelled as trees of commands, rather than sequence of commands.
+  \item \ma programs are modelled as \keyword{trees of commands}, rather than sequence of commands.
     \medskip
@@
     \medskip
-  \item \ma expressions are translated to \lm internal core type theory.
+  \item \ma expressions are translated to \lm internal \keyword{core type theory}.
     \medskip
@@
       \medskip
-    \item The typechecker understands nontrivial algebraic properties.
+    \item The typechecker understands \keyword{nontrivial algebraic properties}.
     \end{itemize}
@@
     \smallskip
-    \begin{itemize}
+    \begin{itemize}[<.->]
     \item Available on GitHub, please get it in touch if interested.
@@
     \begin{itemize}
-    \item \keyword{Uniqueness of representation}: currently, a matrix with quantities can have more than one corresponding type; this might lead to odd behaviour during typechecking.
+    \item<.-> \keyword{Uniqueness of representation}: currently, a matrix with quantities can have more than one corresponding type; this might lead to odd behaviour during typechecking.
       \medskip

### 2024-09-10 examples/imeko.m
[ new ] basic readFrom directive

@@ -1,22 +1,25 @@
-%%> times    :: [ 1 x 12 ] double
+%  %> dimensions V for Q over metre for `Length, kg for `Mass, sec for `Time
 
-%%> voltages :: [ 1 x 12 ] double
+%> times    :: [ 1 x 12 ] double
+%> voltages :: [ 1 x 12 ] double
 
+%> readfrom 'inputs.txt' times voltages
 
 %> z3 :: [ 3 x 1 ] double
 z3 = [ 0; 0; 0 ]
 %> i3 :: [ 3 x 1 ] double
 i3 = [ 1; 1; 1 ]
 
-% %> ddnc :: [ 12 x 4 ] double
 ddnc = [ i3  z3  i3  z3
         -i3  z3  i3  z3
          z3  i3  z3  i3
          z3 -i3  z3  i3 ]
 %> typeof ddnc
 
-%M = [ ddnc times' ]
-%%> typeof M
+M = [ ddnc times' ]
+%> typeof M
 
-%x = M \ voltages
-%%> typeof x
+x = M \ voltages'
+%> typeof x
+
+

### 2024-09-10 src/Machine.hs

@@ -128,6 +128,8 @@ instance PrettyPrint DiagnosticData where
     pure ["unit $" ++ show un ++ " " ++ stat ++ " $" ++ show tn]
   pprint (UnitDs units) =
     pure ["units" ++ foldMap (\u -> " $" ++ show (nonce u)) units]
+  pprint (ReadFromD filename vars) =
+    pure ["readfrom" ++ filename ++ " " ++ show vars]
 
 instance PrettyPrint Frame where
   pprint (Declaration n UserDecl{..}) = do
@@ -165,6 +167,7 @@ data DiagnosticData
       Expr {- e, t = elab e -}
   | UnitD Nonce BOOL Nonce
   | UnitDs [WithSource String]
+  | ReadFromD String [Either String (String, TYPE)]
   deriving Show
 
 data Frame where
@@ -1966,6 +1969,15 @@ runDirective rl (dir :<=: src, body) = do
     InputFormat name | Just (InputFormatBody body) <- body -> do
       newProb $ InputFormatAction name body rl
       run
+    ReadFrom filename variables -> do
+      vs <- for variables $ \ v -> findDeclaration (UserDecl{varTy = Nothing, currentName = v, seen = False, newNames = [], capturable = False, whereAmI = MatLab}) >>= \case
+        Just (name, ty) -> do
+          pushDefinition name (E $^ M (name, Zy) $^ S0 $^ U :^ Ze)
+          pure $ Right (v, ty)
+        Nothing -> pure $ Left $ "Unknown variable " ++ v
+      push $ Diagnostic rl (ReadFromD filename vs)
+      newProb $ Done nil
+      run
     Declare xs ty -> do
       (sol, prob) <- elab "declType" emptyContext (atom SType) (TensorTask <$> ty)
       pushProblems (Sourced . fmap (DeclareAction sol) <$> xs)
@@ -2871,6 +2883,10 @@ normaliseFrame = \case
       stat <- normalise VN (atom STwo) stat
       pure $ UnitD un stat tn
     normaliseDiagnostic d@(UnitDs _) = pure d
+    normaliseDiagnostic d@(ReadFromD filename vs) = do
+      vs <- traverse (either (pure . Left)
+                     (\ (s, ty) -> (Right . (s,)) <$> normalise emptyContext (atom SType) ty)) vs
+      pure $ ReadFromD filename vs
 
 cleanup :: Mood -> Elab ()
 cleanup mood = do
@@ -2952,7 +2968,53 @@ diagnosticRun = llup >>= \case
           units ++
[…]

### 2024-09-10 src/Parse/Matlab.hs

@@ -120,6 +120,7 @@ pdirhead :: Parser DirHeader
 pdirhead = Declare <$> psep1 (pspc <|> punc ",") (pws pnomNotLabMateKey) <* pospc <* psym "::" <* pospc <*> ptensortype
     <|> Rename <$ pkin Nom "rename" <* pospc <*> pnomNotLabMateKey <* pspc <*> pnomNotLabMateKey
     <|> InputFormat <$ pkin Nom "input" <* pospc <*> pnom
+    <|> ReadFrom <$ pkin Nom "readfrom" <* pospc <*> pstringlit <* pspc <*> psep1 pspc pnom
     <|> Typecheck <$ pkin Nom "typecheck" <* pospc <*> ptensortype <* pspc <*> pexpr topCI
     <|> SynthType <$ pkin Nom "typeof" <* pospc <*> pexpr topCI
     <|> Dimensions <$ pkin Nom "dimensions" <* pspc

### 2024-09-10 src/Syntax.hs

@@ -32,6 +32,9 @@ data DirHeader
   = Declare [WithSource String] ConcreteType
   | Rename String String
   | InputFormat String {- name of function -}
+  | ReadFrom
+      String -- filename
+      [String] -- variables to generate code for
   | Typecheck ConcreteType Expr
   | SynthType Expr
   | Dimensions

### 2024-09-10 src/Machine.hs
[ fix ] fix for multiplying matrices - await all statuses

@@ -265,7 +265,7 @@ data ElabTask where
     -> Term Chk ^ n  -- the scoped solution we will commit if the
                      -- Boolean becomes true
     -> ElabTask
-  ConstrainTask :: NATTY n => WhereAmI -> (String, Constraint) -> Term Chk ^ n -> ElabTask
+  ConstrainTask :: NATTY n => WhereAmI -> (String, Constraint) -> [BOOL] -> Term Chk ^ n -> ElabTask
   ElimTask :: NATTY n => Natty n -> (Term Chk ^ n, Typ ^ n) -> [TypeExpr] -> ElabTask
   AbelTask :: TypeExpr' -> ElabTask -- the problem type must be Abel genTy for some genTy
   ConstantCellTask :: SynthesisMode -> TypeExpr' -> ElabTask
@@ -1501,7 +1501,7 @@ solveConstraint name c@HeadersCompatibility{..} = nattily (vlen constraintCtx) $
       let consCase (leftElement , leftList) (rightElement, rightList) stat = do
             (_, hstat) <- constrain "headCompat" $ HeaderCompatibility{..}
             (_, tstat) <- constrain "tailCompat" $ HeadersCompatibility{..}
-            metaDefn name (conjunction [tstat, hstat])
+            metaDefn name (conjunction [tstat, hstat, stat])
       case listView leftList of
         [] -> ensureNull rightCtx (mk SAbel rightGenType) rightList >>= metaDefn name
         Right x:xs -> do
@@ -2320,13 +2320,13 @@ runElabTask sol meta@Meta{..} etask = nattily (vlen mctxt) $ do
           move worried
         FindSimplest -> do
           invCellTy <- invent "invCellTy" (mctxt \\\ ("j", mk SAbel colGenTy) \\\ ("i", mk SAbel (wk rowGenTy))) (atom SType)
-          (_, lstat) <- constrain "sameLength" $ SameLength
+          {- (_, lstat) <- constrain "sameLength" $ SameLength
             { constraintCtx = fmap Hom <$> mctxt
             , leftEltType = mk SAbel rowGenTy
             , leftList = rs
             , rightEltType = mk SAbel colGenTy
             , rightList = cs
-            }
+            } -}
           (_, istat) <- constrain "inverseCell" $ InverseQuantities
[…]

### 2024-09-10 examples/imeko-bad.m
[ print ] more pretty printing of matrices

@@ -1,0 +1,23 @@
+%  %> dimensions V for Q over metre for `Length, kg for `Mass, sec for `Time
+
+%> times    :: [ 1 x 12 ] double
+%> voltages :: [ 1 x 12 ] double
+
+%> readfrom 'inputs.txt' times voltages
+
+%> z3 :: [ 3 x 1 ] double
+z3 = [ 0; 0; 0 ]
+%> i3 :: [ 3 x 1 ] double
+i3 = [ 1; 1; 1 ]
+
+ddnc = [ i3  z3  i3  z3
+        -i3  z3  i3  z3
+         z3  i3  z3  i3
+         z3 -i3  z3  i3 ]
+%> typeof ddnc
+
+M = [ ddnc times' ]
+%> typeof M
+
+x = M \ voltages
+%> typeof x

### 2024-09-10 src/Machine.hs

@@ -3051,6 +3051,28 @@ unelabType ctx ty | Just ct <- tagEh ty = case ct of
               , [spc 1, sym "]", spc 1]
               , cellTy
               ]
+          (_, _, cellTy' :^ Su (No th), Intg r, _) -> do
+            cs <- unelabTerm ctx (mk SList (tag SAbel [colGenTy])) cs
+            cellTy <- unelabType (ctx \\\ (j, mk SAbel colGenTy)) (cellTy' :^ Su th)
+            pure $ concat
+              [ intersperse (spc 1) [sym "[", sym (show r), sym "x", sym j, sym "<-"]
+              , [spc 1]
+              , cs
+              , [spc 1, sym "]", spc 1]
+              , cellTy
+              ]
+          (_, _, cellTy' :^ No (Su th), _, Intg c) -> do
+            rs <- unelabTerm ctx (mk SList (tag SAbel [rowGenTy])) rs
+            cellTy <- unelabType (ctx \\\ (i, mk SAbel rowGenTy)) (cellTy' :^ Su th)
+            pure $ concat
+              [ intersperse (spc 1) [sym "[", sym i, sym "<-"]
+              , [spc 1]
+              , rs
+              , [spc 1]
+              , intersperse (spc 1) [sym "x", sym (show c), sym "]"]
+              , [spc 1]
+              , cellTy
+              ]
           _ -> pure [sym $ show ty]
   (SAbel, [ty]) | isUnitType ty -> pure [nom "int"]
   (SList, [ty]) | Just (SChar, []) <- tagEh ty -> pure [nom "string"]

### 2024-09-10 src/Machine.hs
[ print ] better printing of qualities

@@ -45,7 +45,6 @@ debugCatch = const id -- trace
 debugMatrix = const id -- trace
 debug260824 = trace
 
-
 type Elab = State MachineState
 
 data MachineState = MS
@@ -53,6 +52,7 @@ data MachineState = MS
   , problem :: Problem
   , nameSupply :: (Root, Int)
   , nonceTable :: Map Nonce String
+  , refoldingMap :: Map TYPE String -- shitty version
   , metaStore :: Store
   , constraintStore :: Map Name Constraint
   , clock :: Int
@@ -65,6 +65,7 @@ initMachine f t = MS
   , problem = File f
   , nameSupply = (B0 :< ("labmate", 0), 0)
   , nonceTable = t
+  , refoldingMap = Map.empty
   , metaStore = Map.empty
   , constraintStore = Map.empty
   , clock = 0
@@ -1995,8 +1996,9 @@ runDirective rl (dir :<=: src, body) = do
       newProb eProb
       run
     Dimensions (g :<=: gsrc) (q :<=: qsrc) unitAtoms -> do
-      let generators = mk SEnum $ foldr (mk Splus . (mk Sone :: TERM -> TERM) . atom . what) nil (snd <$> unitAtoms) :: TERM
-      let gdef = mk SAbel generators :: TERM
+      let generators = mk SEnum $ foldr (mk Splus . (mk Sone :: TERM -> TERM) . atom . what) nil (snd <$> unitAtoms) :: TYPE
+      generators <- normalise emptyContext (atom SType) generators
+      let gdef = mk SAbel generators :: TYPE
       let gdecl = UserDecl
             { varTy = mk SType
             , currentName = g
@@ -2019,8 +2021,8 @@ runDirective rl (dir :<=: src, body) = do
                 }
           freshDeclaration qdecl >>= \case
             Nothing -> move worried
-            Just (q, qty) -> do
-              pushDefinition q (lam "d" $ mk SQuantity (wk generators) (evar 0))
+            Just (q', qty) -> do
+              pushDefinition q' (lam "d" $ mk SQuantity (wk generators) (evar 0))
               ps <- fmap concat . for (zip [0..] unitAtoms) $ \case
                 (_, (Nothing, _)) -> pure []
[…]

### 2024-09-10 src/Term/Thinning.hs

@@ -116,6 +116,9 @@ data (^) :: (Nat -> Type) -> Nat -> Type where
 scopeOf :: p ^ n -> Natty n
 scopeOf (_ :^ th) = bigEnd th
 
+supportOf :: p ^ n -> Ex Natty
+supportOf (_ :^ th) = Ex $ weeEnd th
+
 withScopeOf :: p ^ n -> (NATTY n => b) -> b
 withScopeOf x k = nattily (scopeOf x) k
 

### 2024-09-10 src/Machine.hs
[ wip ] give up on solving constraints with differing integral constants

@@ -975,38 +975,40 @@ solveConstraint name c@Constraint{..} = case (traverse (traverse isHom) constrai
     lhsFlexEh <- flexEh ctx ty lhs
     rhsFlexEh <- flexEh ctx ty rhs
     case (lhs, rhs) of
-      _ | lhs == rhs -> metaDefn name (I 1 :$ U :^ no Zy)
+      _ | lhs == rhs -> metaDefn name tRUE
       (_, t :^ ph)
         | Just (FlexEh x k th ta) <- lhsFlexEh
         , Just ps <- thicken ph th
         , x `Set.notMember` dependencies t -> do
               metaDefn x ((t :^ ps) //^ ta)
-              metaDefn name (I 1 :$ U :^ no Zy)
+              metaDefn name tRUE
       (t :^ ph, _)
         | Just (FlexEh x k th ta) <- rhsFlexEh
         , Just ps <- thicken ph th
         , x `Set.notMember` dependencies t -> do
               metaDefn x ((t :^ ps) //^ ta)
-              metaDefn name (I 1 :$ U :^ no Zy)
+              metaDefn name tRUE
       -- different atoms are never unifiable, raise the constraint as impossible
       _ | Just (s1, []) <- tagEh lhs
         , Just (s2, []) <- tagEh rhs
         , s1 /= s2 -> do
             debug ("Push Constraint n-5 " ++ show c) $ push $ ConstraintFrame name c
-            debug (concat ["Unifying different atoms: ", s1, " and ", s2, "."]) $ metaDefn name (I 0 :$ U :^ no Zy)
+            debug (concat ["Unifying different atoms: ", s1, " and ", s2, "."]) $ metaDefn name fALSE
       _ | Just (SList, [elty])  <- tagEh ty
         , isUnitType elty -> elabTC ctx (natTermCancel (lhs,rhs)) >>= \case
             Left err -> error err
-            Right (lhs', rhs') -> if lhs == lhs'
-              then push $ ConstraintFrame name c
-              else do
-                (_, cstat) <- constrain "cancellation" $ Constraint
-                  { constraintCtx = constraintCtx
-                  , constraintType = Hom ty
-                  , lhs = lhs'
-                  , rhs = rhs'
-                  }
-                metaDefn name cstat
+            Right (lhs', rhs') -> if Set.null (dependencies (tup [lhs', rhs']))
+              then metaDefn name fALSE
+              else if lhs == lhs'
+                then push $ ConstraintFrame name c
+                else do
[…]

### 2024-09-10 examples/imeko-bad.m
[ new ] more error reporting. 12 does not align with 1.

@@ -1,5 +1,3 @@
-%  %> dimensions V for Q over metre for `Length, kg for `Mass, sec for `Time
-
 %> times    :: [ 1 x 12 ] double
 %> voltages :: [ 1 x 12 ] double
 

### 2024-09-10 src/Machine.hs

@@ -43,7 +43,6 @@ import Control.Monad.Reader
 debug = const id -- trace
 debugCatch = const id -- trace
 debugMatrix = const id -- trace
-debug260824 = trace
 
 type Elab = State MachineState
 
@@ -53,6 +52,8 @@ data MachineState = MS
   , nameSupply :: (Root, Int)
   , nonceTable :: Map Nonce String
   , refoldingMap :: Map TYPE String -- shitty version
+  , elabStore :: Map Name [Tok]
+  , sourceGripeTable :: Map Nonce [(BOOL, [Either Unelabtask [Tok]])]
   , metaStore :: Store
   , constraintStore :: Map Name Constraint
   , clock :: Int
@@ -66,11 +67,28 @@ initMachine f t = MS
   , nameSupply = (B0 :< ("labmate", 0), 0)
   , nonceTable = t
   , refoldingMap = Map.empty
+  , elabStore = Map.empty
+  , sourceGripeTable = Map.empty
   , metaStore = Map.empty
   , constraintStore = Map.empty
   , clock = 0
   , epoch = 0
   }
+
+data Unelabtask = forall n . Unelabtask
+  { unElabContext :: Context n
+  , unElabType :: Term Chk ^ n
+  , unElabTerm :: Term Chk ^ n
+  }
+
+instance Show Unelabtask where
+  show _ = "<>" --TODO
+
+elabUnelabtask :: Unelabtask -> Elab [Tok]
+elabUnelabtask (Unelabtask ctx ty tm) = withScopeOf ty $ do
+  ty <- normalise ctx (atom SType) ty
+  tm <- normalise ctx ty tm
+  unelabTerm ctx ty tm
 
 -- whatever PPMonad is, it should have Monoid instance which lifts the
 -- monoid structure of values
@@ -937,7 +955,7 @@ ensureNull ctx ty tm = case partitionEithers (listView tm) of
   ([], []) -> pure tRUE
   (_, _:_) -> pure fALSE
   (es, []) -> conjunction <$> (for es $ \ e -> do
-      (_, stat) <- constrain "ensureNulls" $ Constraint
+      (_, stat) <- substrain "ensureNulls" $ Constraint
         { constraintCtx = fmap Hom <$> ctx
         , constraintType = Hom (mk SList ty)
         , lhs = E $^ e
@@ -951,7 +969,7 @@ ensureCons ctx ty tm = case listView tm of
   _ -> do
[…]

### 2024-09-11 src/Lex.hs
[ more ] error messages

@@ -435,7 +435,9 @@ groupRaw g ts = groupString g (ts >>= raw)
 nonceExpand :: Map Nonce String -> Tok -> String
 nonceExpand table t | Non n <- kin t = case Map.lookup n table of
   Just ts -> ts
-  Nothing -> error $ "should be impossible: " ++ show n
+  Nothing -> case n of
+    -1 -> "..."
+    _ -> error $ "should be impossible: " ++ show n
 nonceExpand table t | Grp g (Hide ts) <- kin t = groupString g (ts >>= nonceExpand table)
 nonceExpand _ t = raw t
 

### 2024-09-11 src/Machine.hs

@@ -53,7 +53,8 @@ data MachineState = MS
   , nonceTable :: Map Nonce String
   , refoldingMap :: Map TYPE String -- shitty version
   , elabStore :: Map Name [Tok]
-  , sourceGripeTable :: Map Nonce [(BOOL, [Either Unelabtask [Tok]])]
+  , sourceGripeTable :: Map Nonce [SourceIssue]
+  , sourceFudTable :: Map Nonce [(BOOL, SourceMessage)]
   , metaStore :: Store
   , constraintStore :: Map Name Constraint
   , clock :: Int
@@ -69,11 +70,20 @@ initMachine f t = MS
   , refoldingMap = Map.empty
   , elabStore = Map.empty
   , sourceGripeTable = Map.empty
+  , sourceFudTable = Map.empty
   , metaStore = Map.empty
   , constraintStore = Map.empty
   , clock = 0
   , epoch = 0
   }
+
+data SourceIssue = SourceIssue
+  { issueStatus :: BOOL
+  , messageIfFalse :: SourceMessage
+  , messageIfDoubt :: SourceMessage
+  } deriving Show
+
+type SourceMessage = [Either Unelabtask [Tok]]
 
 data Unelabtask = forall n . Unelabtask
   { unElabContext :: Context n
@@ -278,7 +288,7 @@ data ElabTask where
   TensorTask :: TensorType' -> ElabTask
   TypeExprTask :: WhereAmI -> SynthesisMode -> TypeExpr' -> ElabTask
   LHSTask :: LHS' -> ElabTask
-  ExprTask :: WhereAmI -> SynthesisMode -> Expr' -> ElabTask
+  ExprTask :: WhereAmI -> SynthesisMode -> Expr -> ElabTask
   Await :: NATTY n
     => BOOL          -- the closed Boolean value we hope will become true
     -> Term Chk ^ n  -- the scoped solution we will commit if the
@@ -977,27 +987,30 @@ ensureCons ctx ty tm = case listView tm of
       }
     pure (x, xs, stat)
 
-constrain :: String -> Constraint -> Elab (Name, BOOL)
-constrain s c = do
+constrainWithIssue :: String -> (SourceMessage, SourceMessage) -> Constraint -> Elab (Name, BOOL)
+constrainWithIssue s (fud, gripe) c =  do
   c <- updateConstraint c
   name <- metaDecl Hoping s emptyContext (atom STwo)
   solveConstraint name c
[…]

### 2024-09-11 src/Parse/Matlab.hs

@@ -168,7 +168,9 @@ ptypeexpr ci = go >>= more ci where
    <|> pws (TyAtom <$> patom)
    <|> pws (pgrp (== Bracket Curly) (TyBraces . join <$> optional (plink (id <$ pospc <*> optional (ptypeexpr topCI <* pospc)))))
   lhsstuff ci = (pws (TyVar <$> pnom) >>= lmore)
-            <|> pws (tyMat <$> pgrp (== Bracket Square) (many (prow ptypeexpr)))
+            <|> (do
+                        (e :<=: src) <- pws (pgrp (== Bracket Square) (many (prow ptypeexpr)))
+                        pure $ tyMat src e :<=: src)
   lmore l = pcond
               (pws' (nonce l) (TyApp l <$ pcxspc ci <*> pargs Round (ptypeexpr topCI)))
               lmore

### 2024-09-11 writing/NPLSep2024/examples-start/cubit.m
[ gig ] examples from the train

@@ -1,0 +1,12 @@
+%> dimensions D for Q over metre for `Length
+
+inch = 0.0254*metre;
+foot = 12*inch;
+cubit = 0.462788*metre;
+
+x = 5*metre + 12*foot;
+%> typeof x
+
+y = x/cubit
+%> typeof y
+

### 2024-09-11 writing/NPLSep2024/examples-start/dim-add.m

@@ -1,0 +1,45 @@
+%< LabMate 0.2.0.6
+%> dimensions D for Q over metre for `Length, kg for `Mass, sec for `Time
+%<{
+% LabMate tells MATLAB that units are 1s
+metre = 1;
+kg = 1;
+sec = 1;
+%<}
+
+%> x :: [ i <- [{} {`Time}] x j <- [{} {`Length}] ] Q({j/i})
+x = [ 3      2*metre
+      5/sec  4*metre/sec ]
+%> typeof 2*metre
+%< 2*metre :: Q({`Length})
+%> typeof x
+%< x :: [ i <- [{}, {`Time}] x j <- [{}, {`Length}] ] Q({j / i})
+                                                                            
+%> y :: [ i <- [{} {`Time}] x j <- [{} {`Length*`Mass}] ] Q({j/i/`Mass})
+y = [ 7/kg      8*metre
+      5/sec/kg  6*metre/sec ]
+%>typeof y
+%< y :: [ i <- [{}, {`Time}] x j <- [{}, {`Mass * `Length}] ] Q({j / i / `Mass})
+
+z = x + y
+%> typeof z
+%< There is something wrong with z. Relevant: x + y
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+

### 2024-09-11 writing/NPLSep2024/examples-start/vjux-bad.m

@@ -1,0 +1,19 @@
+%< LabMate 0.2.0.6
+%> x :: [ 3 x 2 ] double
+x = [ 1 2
+      3 4
+      5 6 ]
+%> typeof x
+
+%> y :: [ 2 x 3 ] double
+y = [ 1 2 3
+      4 5 6 ]
+%> typeof y
+
+z = [ x
+      y ]
+%> typeof z
+
+
+
+

### 2024-09-11 src/Lex.hs
[ fix ] multiline synth responses get a %< per line

@@ -49,6 +49,7 @@ data Grouping = Literal
               | Block
               | Bracket Bracket
               | Line LineTerminator
+              | Indentation String
               | Error
   deriving (Show, Eq)
 
@@ -418,7 +419,7 @@ seeToks ts = go 0 ts where
         _ -> return ()
 
 groupString :: Grouping -> String -> String
-groupString g s = prefix g ++ s ++ suffix g
+groupString g s = (prefix g ++ s ++ suffix g) >>= dent g
   where
     prefix g = case g of
       Bracket b -> fst (brackets b)
@@ -428,6 +429,8 @@ groupString g s = prefix g ++ s ++ suffix g
       Bracket b -> snd (brackets b)
       Generated -> "%<}"
       _ -> ""
+    dent (Indentation dent) '\n' = '\n' : dent
+    dent g t = [t]
 
 groupRaw :: Grouping -> [Tok] -> String
 groupRaw g ts = groupString g (ts >>= raw)

### 2024-09-11 src/Machine.hs

@@ -2054,12 +2054,12 @@ runDirective rl (dir :<=: src, body) = do
       run
     Typecheck ty e -> do
       (tySol, tyProb) <- elab "typeToCheck" emptyContext (atom SType) (TensorTask <$> ty)
-      (eSol, eProb) <- elab "exprToCheck" emptyContext tySol (ExprTask LabMate EnsureCompatibility e :<=: source e)
+      (eSol, eProb) <- elab "exprToCheck" emptyContext tySol (ExprTask MatLab EnsureCompatibility e :<=: source e)
       newProb tyProb
       run
     SynthType e -> do
       ty <- invent "typeToSynth" emptyContext (atom SType)
-      (eSol, eProb) <- elab "exprToSynth" emptyContext ty (ExprTask LabMate FindSimplest e :<=: source e)
+      (eSol, eProb) <- elab "exprToSynth" emptyContext ty (ExprTask MatLab FindSimplest e :<=: source e)
       push $ Diagnostic rl (SynthD ty eSol e)
       newProb eProb
       run
@@ -3031,10 +3031,12 @@ diagnosticRun = llup >>= \case
                          _ -> []
                     pure $ [sym " Relevant:", spc 1] ++ intercalate [sym ",", spc 1] ps'
                   [] -> pure []
+    quack dent toks =
+      [Tok "" (Grp (Indentation (replicate dent ' ' ++ "%< ")) (Hide toks)) dump]
     msg :: Int -> DiagnosticData -> Elab [Tok]
     msg dent = \case
-      SynthD ty tm (e :<=: (en, _)) -> do
-        let resp = [Tok "\n" Ret dump, spc dent, sym "%<", spc 1]
+      SynthD ty tm (e :<=: (en, _)) -> quack dent <$> do
+        let resp = [Tok "\n" Ret dump]
         estore <- gets elabStore
         statTy <- traverse (\ x -> (,Map.lookup x estore) <$> metaStatus x) . Set.toList $ dependencies ty
         statTm <- traverse (\ x -> (,Map.lookup x estore) <$> metaStatus x) . Set.toList $ dependencies tm

### 2024-09-11 src/Machine.hs
[ fix ] avoid missing source frame error by using -1 as in "nosource"

@@ -1770,7 +1770,9 @@ ensureDeclaration s = findDeclaration s >>= \case
 findNearestSource :: Elab Nonce
 findNearestSource = excursion go where
   go = pull >>= \case
-    Nothing -> error "findNearestSource: impossible no enclosing Source frame"
+    Nothing -> pure (-1)
+      -- TODO: what, really?
+      -- error "findNearestSource: impossible no enclosing Source frame"
     Just fr@(Source s@(n, Hide ts)) -> n <$ push fr
     Just f -> shup f >> go
 

### 2024-09-11 writing/NPLSep2024/examples-start/4-0-renaming-simple.m
[ gig ] renaming examples

@@ -1,0 +1,9 @@
+x = 5;
+y = x + x;
+
+%> rename x a
+%> rename y b
+
+function y = foo(x)
+  y = x + 7;
+end

### 2024-09-11 writing/NPLSep2024/examples-start/4-1-nested-captured.m

@@ -1,0 +1,19 @@
+
+f1()
+
+function x = f1()
+  % This is an example of a renaming the Matlab editor gets wrong.
+  % It should be disallowed, because it captures w in f2.
+
+  %> rename z w
+
+  z = 5;
+  k = f2();
+  x = z;
+
+  function y = f2()
+    w = 7;
+    y = w;
+  end
+end
+

### 2024-09-11 writing/NPLSep2024/examples-start/demo_5_1_cubit.m
[ gig ] demos in start config

@@ -1,4 +1,4 @@
-%> dimensions D for Q over metre for `Length
+%> dimensions D for Q over metre for `Length, kg for `Mass, sec for `Time
 
 inch = 0.0254*metre;
 foot = 12*inch;

### 2024-09-11 writing/NPLSep2024/examples-start/demo_5_2_dim_add.m

@@ -1,0 +1,33 @@
+%> dimensions D for Q over metre for `Length, kg for `Mass, sec for `Time
+
+%> x :: [ i <- [{} {`Time}] x j <- [{} {`Length}] ] Q({j/i})
+x = [ 3      2*metre
+      5/sec  4*metre/sec ]
+%> typeof 2*metre
+                                                                            
+%> y :: [ i <- [{} {`Time}] x j <- [{} {`Length}] ] Q({j/i})
+y = [ 7      8*metre
+      5/sec  6*metre/sec ]
+
+z = x + y
+%> typeof z
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+

### 2024-09-11 writing/NPLSep2024/examples-start/demo_5_5_hello_dimensions.m

@@ -1,0 +1,29 @@
+%> dimensions D for Q over metre for `Length, kg for `Mass, sec for `Time, amp for `Current
+
+newton = kg * metre / sec / sec
+joule = newton * metre
+watt = joule / sec
+volt = watt / amp
+%> typeof volt
+
+%> times    :: [ 1 x 12 ] double
+%> voltages :: [ 1 x 12 ] double
+
+%> readfrom 'inputs.txt' times voltages
+
+%> z3 :: [ 3 x 1 ] double
+z3 = [ 0; 0; 0 ]
+%> i3 :: [ 3 x 1 ] double
+i3 = [ 1; 1; 1 ]
+
+ddnc = [ i3  z3  i3  z3
+        -i3  z3  i3  z3
+         z3  i3  z3  i3
+         z3 -i3  z3  i3 ]
+%> typeof ddnc
+
+M = [ ddnc times' ]
+%> typeof M
+
+x = M \ voltages'
+%> typeof x

### 2024-09-11 writing/NPLSep2024/examples-start/dim-add.m

@@ -1,45 +1,0 @@
-%< LabMate 0.2.0.6
-%> dimensions D for Q over metre for `Length, kg for `Mass, sec for `Time
-%<{
-% LabMate tells MATLAB that units are 1s
-metre = 1;
-kg = 1;
-sec = 1;
-%<}
-
-%> x :: [ i <- [{} {`Time}] x j <- [{} {`Length}] ] Q({j/i})
-x = [ 3      2*metre
-      5/sec  4*metre/sec ]
-%> typeof 2*metre
-%< 2*metre :: Q({`Length})
-%> typeof x
-%< x :: [ i <- [{}, {`Time}] x j <- [{}, {`Length}] ] Q({j / i})
-                                                                            
-%> y :: [ i <- [{} {`Time}] x j <- [{} {`Length*`Mass}] ] Q({j/i/`Mass})
-y = [ 7/kg      8*metre
-      5/sec/kg  6*metre/sec ]
-%>typeof y
-%< y :: [ i <- [{}, {`Time}] x j <- [{}, {`Mass * `Length}] ] Q({j / i / `Mass})
-
-z = x + y
-%> typeof z
-%< There is something wrong with z. Relevant: x + y
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-
-

### 2024-09-12 writing/NPLSep2024/examples-live/demo_5_1_cubit.m
[ demo ] add live copy, fix minor things

@@ -1,0 +1,12 @@
+%> dimensions D for Q over metre for `Length, kg for `Mass, sec for `Time
+
+inch = 0.0254*metre;
+foot = 12*inch;
+cubit = 0.462788*metre;
+
+x = 5*metre + 12*foot;
+%> typeof x
+
+y = x/cubit
+%> typeof y
+

### 2024-09-12 writing/NPLSep2024/examples-live/demo_5_2_dim_add.m

@@ -27,7 +27,3 @@ z = x + y
 
 
 
-
-
-
-

### 2024-09-12 writing/IMEKOWC2024/presentation/presentation.tex

@@
 \institute[]{$^{1}$University of Strathclyde, UK\\$^{2}$National Physical Laboratory, UK}
-
+\date{IMEKO 2024 XXIV World Congress\\August 27, 2024}
 \begin{document}
@@
    \end{lstlisting}
-  \pause
   \mycallout<+>{ren}{this is an input directive to rename a variable}
@@
    \end{lstlisting}
-  \pause
+  %\pause
    \mycallout<+>{renr}{LabMate response to the input directive}
@@
     \end{lstlisting}
-  \pause
     \mycallout<+>{ann1}{type annotations at declaration of A}
@@
     \end{lstlisting}
- \pause
     \mycallout<+>{query}{query for the type of C}
@@
     \end{lstlisting}
-    \pause
     \mycalloutup<+>{bad}{LabMate can point out an error with D}
@@
      \end{lstlisting}
-   \pause
      \mycalloutup<+>{fromb}{LabMate infers type of A from the annotation on B}
-   \mycalloutup<+>{scope}{tracks Matlab scope}
+     \mycalloutup<+>{scope}{distinguishes and tracks Matlab scope}
   \end{itemize}
@@
     \end{lstlisting}
-%\pause
     \mycalloutup<+>{base}{define some base set of dimensions}
@@
    \end{lstlisting}
-%  \pause
    \mycallout<+>{q}{turn a value of a dimensionless type into a quantity}
@@
   \item A common scenario when working with matrices of quantities:
-%    $\begin{array}{ccccc}
-% e_{1,1}  &        & e_{1,j} &        & e_{1,n} \\
-%         &        &         &        &         \\
-% e_{i,1}  &        & e_{i,j}\tikzmark{mat} &        & e_{i,n} \\
-%         &        &         &        &         \\
-% e_{m,1}  &        & e_{m,j} &        & e_{m,n}  \\
-%\end{array}$
-%    \mycallout<+>{mat}{the type might depend on i and j}
     \begin{lstlisting}[xleftmargin=0em]
@@
     \begin{lstlisting}[xleftmargin=0em]
-% > A :: [ i <- [{} {`T}]
-%        x j <- [{} {`L}]
-%        ] Q({`M * j / i})
-A = [ 2*kg       5*kg*metre
-      3*kg/sec   4*kg*metre/sec ]
+% > A :: [ i <-!\tikzmark{indi}! [{} {`T}]
+%        x j <-!\tikzmark{indj}! [{} {`L}]
+%        ] Q({`M * j / i})!\tikzmark{cell}!
+A = [ 2*!\tikzmark{e00}!kg       5*!\tikzmark{e01}!kg*metre
+      3*kg/sec   4*!\tikzmark{e11}!kg*metre/sec]
     \end{lstlisting}
+    \mycalloutup<+>{cell}{The cell type Q depends on i and j}
+    \mycalloutup<+>{indi}{Index i tells us something about the rows}
+    \mycalloutup<+>{indj}{Index j disambiguates the columns}
+    \mycalloutup<+>{e00}{:: Q(\{ `M * \{\} / \{\} \})}
+    \mycalloutup<+>{e01}{:: Q(\{ `M * `L / \{\} \})}
+    \mycalloutup<+>{e11}{:: Q(\{ `M * `L / `T \})}
   \end{itemize}
@@
 
+\begin{frame}
+  \begin{center}
+    \LARGE Thanks for the attention!\\
+    \pause
+    \vspace{1em}
+    \Huge Questions?
+  \end{center}
+\end{frame}
+
 \end{document}
